---
title: "Time grouping in calculate_* functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time grouping in calculate_* functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "amadeus contributors"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(amadeus)
library(terra)
```

## How `.by` and `.by_time` work

`calculate_*()` functions now use `.by` / `.by_time` for optional summarization:

- If `.by = NULL` and `.by_time = NULL`, results keep the native processed time rows.
- If `.by` is set to a time unit (`"day"`, `"week"`, `"month"`, `"year"`), rows are
  grouped by site and that bucket.
- If `.by` is set to a column name (for example `"region"`), grouping uses that
  column. Then `.by_time` can add time grouping (either a time unit or an existing
  time column name).

## Minimal reproducible example using `calculate_geos()`

```{r}
# Build a toy 4-layer raster with GEOS-style layer names
r <- terra::rast(
  ncols = 2, nrows = 1, xmin = -81, xmax = -79, ymin = 35, ymax = 36,
  crs = "EPSG:4326", nlyrs = 4
)
terra::values(r[[1]]) <- c(1, 2)
terra::values(r[[2]]) <- c(3, 4)
terra::values(r[[3]]) <- c(5, 6)
terra::values(r[[4]]) <- c(7, 8)
names(r) <- c(
  "no2_lev=850_20200101_010000",
  "no2_lev=850_20200101_130000",
  "no2_lev=850_20200201_010000",
  "no2_lev=850_20200201_130000"
)

locs <- data.frame(
  site_id = c("A", "B"),
  lon = c(-80.5, -79.5),
  lat = c(35.5, 35.5)
)
```

### 1) Default behavior: native temporal resolution

```{r}
native_rows <- calculate_geos(
  from = r,
  locs = locs,
  locs_id = "site_id",
  radius = 0,
  geom = FALSE
)
head(native_rows)
```

### 2) Time bucketing with `.by`

```{r}
monthly <- calculate_geos(
  from = r,
  locs = locs,
  locs_id = "site_id",
  radius = 0,
  .by = "month",
  geom = FALSE
)
monthly
```

### 3) Group by a column and add time with `.by_time`

```{r}
site_month <- calculate_geos(
  from = r,
  locs = locs,
  locs_id = "site_id",
  radius = 0,
  .by = "site_id",
  .by_time = "month",
  geom = FALSE
)
site_month
```

## Practical guidance

1. Use defaults (`.by = NULL`, `.by_time = NULL`) when you want native temporal
   output.
2. Use `.by` for coarser temporal summaries or custom grouping columns.
3. Use `.by_time` only when `.by` is a non-time column and you also want temporal
   grouping.
