---
title: "USDA CropScape"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{USDA CropScape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_transform(sf::st_as_sf(readRDS(aqs_sample_path)), 4326)
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)
durham_hex_sf <- sf::st_transform(sf::st_as_sf(durham_hex), 4326)

workflow_bbox <- sf::st_bbox(c(
  sf::st_geometry(example_points_sf),
  sf::st_geometry(durham_hex_sf)
))
workflow_extent_wgs84 <- sf::st_as_sfc(sf::st_bbox(c(
  xmin = workflow_bbox[["xmin"]] - 0.15,
  xmax = workflow_bbox[["xmax"]] + 0.15,
  ymin = workflow_bbox[["ymin"]] - 0.15,
  ymax = workflow_bbox[["ymax"]] + 0.15
), crs = sf::st_crs(4326)))


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  preferred <- candidates[vapply(x[candidates], function(col) {
    is.numeric(col) || is.character(col) || is.factor(col)
  }, logical(1))]
  if (length(preferred) > 0) preferred[1] else candidates[1]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::theme_minimal()
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::theme_minimal()
}
```

This article demonstrates a compact workflow for USDA CropScape (Cropland Data Layer) rasters.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "cropscape_workflow")
download_data(
  dataset_name = "cropscape",
  year = 2019,
  source = "GMU",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE
)
download_data(
  dataset_name = "cropscape",
  year = 2020,
  source = "USDA",
  directory_to_save = file.path(directory_to_save, "usda_2020"),
  acknowledgement = TRUE,
  unzip = TRUE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
cropscape_path <- list.files(
  directory_to_save,
  pattern = "cdl_30m_.*2019.*\\.tif$",
  recursive = TRUE,
  full.names = TRUE
)[1]
cropscape_reference <- terra::rast(cropscape_path)
processing_extent <- terra::ext(
  terra::project(terra::vect(workflow_extent_wgs84), terra::crs(cropscape_reference))
)
processed_data <- process_covariates(
  covariate = "cropscape",
  path = cropscape_path,
  year = 2019,
  extent = processing_extent
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "cropscape",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 300,
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "cropscape",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 300,
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("USDA CropScape", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("USDA CropScape", ": Durham H3 polygons"))
```

## Notes

- The download section demonstrates both GMU and USDA CropScape sources so local runs cover more than one source configuration.
- CropScape uses categorical raster classes, so polygon summaries are best interpreted as class composition rather than a continuous surface.
- This workflow crops the source raster with a Durham-area window that is
  transformed from WGS84 into the CropScape raster CRS before processing.
