---
title: "Drought workflows (SPEI, EDDI, USDM)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drought workflows (SPEI, EDDI, USDM)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 8,
  fig.height = 5.5,
  out.width = "100%",
  fig.align = "center"
)
library(amadeus)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

drought_root <- c(
  file.path("tests", "testdata", "drought"),
  file.path("..", "tests", "testdata", "drought")
)
drought_root <- drought_root[file.exists(drought_root)][1]
if (is.na(drought_root) || drought_root == "") {
  stop("Could not locate tests/testdata/drought")
}

spei_path <- file.path(drought_root, "spei")
eddi_path <- file.path(drought_root, "eddi")
usdm_path <- file.path(drought_root, "usdm")

drought_reference <- data.frame(
  source = c("SPEI", "EDDI", "USDM"),
  description = c(
    "Standardized Precipitation-Evapotranspiration Index",
    "Evaporative Demand Drought Index",
    "U.S. Drought Monitor drought-class polygons"
  ),
  cadence = c("Monthly", "Weekly (Tuesday releases)", "Weekly (Tuesday releases)"),
  spatial_form = c("Raster (netCDF)", "Raster (netCDF)", "Polygon vector (shapefile)"),
  key_download_args = c(
    'source = "spei", timescale = <months>, date = c(start, end)',
    'source = "eddi", timescale = <months>, date = c(start, end)',
    'source = "usdm", date = c(start, end)'
  ),
  key_process_args = c(
    'source = "spei", timescale = <months>',
    'source = "eddi", timescale = <months>',
    'source = "usdm"'
  ),
  workflow = c(
    '`download_drought(...)`<br>`process_drought(...)`<br>`calculate_drought(...)`',
    '`download_drought(...)`<br>`process_drought(...)`<br>`calculate_drought(...)`',
    '`download_drought(...)`<br>`process_drought(...)`<br>`calculate_drought(...)`'
  ),
  stringsAsFactors = FALSE
)
```

This article demonstrates drought workflows in `amadeus` for SPEI, EDDI, and USDM using `download_drought()`, `process_drought()`, and `calculate_drought()`.

The live-download chunk below is provided as a reference and is not evaluated during vignette builds.

## Available inputs and data availability

```{r drought-reference, echo = FALSE, results = "asis"}
knitr::kable(
  drought_reference,
  col.names = c(
    "Source",
    "Description",
    "Temporal cadence",
    "Spatial form",
    "Key download args",
    "Key process args",
    "Workflow in amadeus"
  ),
  escape = FALSE
)
```

## Optional live download examples

```{r drought-download-live, eval = live_run}
drought_download_dir <- file.path(tempdir(), "drought_workflow_download")
date_range <- c("2020-01-01", "2020-03-31")
download_drought(
  source = "spei",
  date = date_range,
  timescale = 1L,
  directory_to_save = drought_download_dir,
  acknowledgement = TRUE
)

download_drought(
  source = "eddi",
  date = date_range,
  timescale = 1L,
  directory_to_save = drought_download_dir,
  acknowledgement = TRUE
)

download_drought(
  source = "usdm",
  date = date_range,
  directory_to_save = drought_download_dir,
  acknowledgement = TRUE
)
```

## Process representative drought data

```{r process-drought, eval = live_run}
spei_processed <- process_drought(
  source = "spei",
  path = drought_download_dir,
  date = date_range,
  timescale = 1L
)

eddi_processed <- process_drought(
  source = "eddi",
  path = drought_download_dir,
  date = date_range,
  timescale = 1L
)

usdm_processed <- process_drought(
  source = "usdm",
  path = drought_download_dir,
  date = date_range
)
```

## Plot processed SPEI and EDDI rasters

```{r plot-processed-rasters, fig.height = 6, fig.alt = "Processed SPEI and EDDI raster maps for the first date in each stack.", eval = live_run}
spei_map <- terra::as.data.frame(spei_processed[[1]], xy = TRUE, na.rm = TRUE)
names(spei_map)[3] <- "value"
spei_map$source <- "SPEI"

eddi_map <- terra::as.data.frame(eddi_processed[[1]], xy = TRUE, na.rm = TRUE)
names(eddi_map)[3] <- "value"
eddi_map$source <- "EDDI"

map_df <- rbind(spei_map, eddi_map)

ggplot2::ggplot(map_df, ggplot2::aes(x = x, y = y, fill = value)) +
  ggplot2::geom_raster() +
  ggplot2::facet_wrap(~ source, ncol = 2) +
  ggplot2::scale_fill_viridis_c(option = "C") +
  ggplot2::coord_equal() +
  ggplot2::labs(
    title = "Processed drought rasters",
    subtitle = "First available date from each processed stack",
    x = "Longitude",
    y = "Latitude",
    fill = "Index"
  ) +
  ggplot2::theme_minimal()
```

## Plot processed USDM polygons

```{r plot-processed-usdm, fig.alt = "Processed USDM drought monitor polygons for one weekly date.", eval = live_run}
usdm_dates <- sort(unique(terra::values(usdm_processed)$date))
usdm_one_day <- usdm_processed[terra::values(usdm_processed)$date == usdm_dates[1], ]
usdm_sf <- sf::st_as_sf(usdm_one_day)

ggplot2::ggplot(usdm_sf) +
  ggplot2::geom_sf(ggplot2::aes(fill = factor(DM)), color = NA) +
  ggplot2::scale_fill_brewer(palette = "YlOrRd", na.value = "grey85") +
  ggplot2::coord_sf(datum = NA) +
  ggplot2::labs(
    title = "Processed USDM drought classes",
    subtitle = paste("Date:", usdm_dates[1]),
    x = NULL,
    y = NULL,
    fill = "DM class"
  ) +
  ggplot2::theme_minimal()
```

## Calculate drought values at sample locations

```{r calculate-drought, eval = live_run}
sample_locs <- data.frame(
  site_id = c("site_1", "site_2", "site_3"),
  lon = c(-99.0, -97.5, -96.0),
  lat = c(37.0, 36.0, 35.0)
)

calc_spei <- calculate_drought(
  from = spei_processed,
  locs = sample_locs,
  locs_id = "site_id",
  radius = 0L,
  fun = "mean"
)

calc_eddi <- calculate_drought(
  from = eddi_processed,
  locs = sample_locs,
  locs_id = "site_id",
  radius = 0L,
  fun = "mean"
)

calc_usdm <- calculate_drought(
  from = usdm_processed,
  locs = sample_locs,
  locs_id = "site_id"
)
```

## Plot calculated drought time series with ggplot

```{r plot-calculated, fig.height = 6.5, fig.alt = "Calculated SPEI and EDDI drought values at sample point locations.", eval = live_run}
spei_col <- grep("^spei_", names(calc_spei), value = TRUE)
eddi_col <- grep("^eddi_", names(calc_eddi), value = TRUE)

calc_spei_plot <- data.frame(
  site_id = calc_spei$site_id,
  time = as.Date(calc_spei$time),
  value = calc_spei[[spei_col]],
  source = "SPEI"
)

calc_eddi_plot <- data.frame(
  site_id = calc_eddi$site_id,
  time = as.Date(calc_eddi$time),
  value = calc_eddi[[eddi_col]],
  source = "EDDI"
)

calc_all_plot <- data.table::rbindlist(
  list(calc_spei_plot, calc_eddi_plot),
  fill = TRUE
)

ggplot2::ggplot(
  calc_all_plot,
  ggplot2::aes(x = time, y = value, color = site_id, group = site_id)
) +
  ggplot2::geom_line(linewidth = 0.5) +
  ggplot2::geom_point(size = 1.3) +
  ggplot2::facet_wrap(~ source, scales = "free_y", ncol = 1) +
  ggplot2::scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  ggplot2::labs(
    title = "Calculated drought covariates at sample locations",
    subtitle = "SPEI and EDDI index values",
    x = "Date",
    y = "Calculated value",
    color = "Location"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))
```

## Notes

- SPEI and EDDI are raster products; USDM is a weekly polygon product, so processing and extraction differ by source.
- For SPEI/EDDI, `timescale` controls the accumulation period and must match the downloaded files.
- USDM extraction returns drought-monitor class (`DM`) values for each location/date.
