---
title: "US EPA Ecoregions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{US EPA Ecoregions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))


```

This article demonstrates a compact workflow for the US EPA ecoregion layers distributed through `amadeus`. The download is a single national layer that contains the multiple ecoregion levels used during point and polygon extraction.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "ecoregion_workflow")
download_data(
  dataset_name = "ecoregion",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
ecoregion_path <- list.files(
  directory_to_save,
  pattern = "\\.(gpkg|shp)$",
  recursive = TRUE,
  full.names = TRUE
)[1]
processed_data <- process_covariates(
  covariate = "ecoregion",
  path = ecoregion_path
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_ecoregion(
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  geom = "sf"
)
```



## Visualize the outputs
Print the output from calculate_ecoregion. The result is a binary(dummy) variable where each location is 1 if it falls in the specified ecoregion, and 0 otherwise. 

```{r calculated data, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
print(point_values)

```


## Notes

- EPA ecoregions are distributed as a single vector layer, so one download covers both the level-2 and level-3 classifications used downstream.
- The example points come from a small saved subset of AQS monitor locations. Pass an `extent` argument to `process_covariates()` if you want to restrict processing to a smaller region around those sites.
