---
title: "US EPA Ecoregions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{US EPA Ecoregions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)
library(terra)
live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))


```

This article demonstrates a compact workflow for the US EPA ecoregion layers distributed through `amadeus`. The download is a single national layer that contains the multiple ecoregion levels used during point and polygon extraction.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

- `download_data(dataset_name = "ecoregion", ...)` does not take product, year, scenario, or region arguments; it downloads the EPA U.S. ecoregion distribution used by the package.
- This is a static national vector layer with no temporal dimension.
- Downloads arrive as a zipped vector-data bundle that can be processed into workflow-ready ecoregion boundaries and identifiers.
- Ecoregions are a categorical covariate, so we have two options for calculation output: 
  - Binary (dummy) indicator columns for the intersecting level-2 and level-3 ecoregions at each location, with key-based names (for example `DUM_E2083_00000`, `DUM_E3064_00000`).
  - Fraction or proportion of each location that falls in each intersecting level-2 and level-3 ecoregion, with key-based names (for example `FRC_E2083_00000`, `FRC_E3064_00000`).

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "ecoregion_workflow")
download_data(
  dataset_name = "ecoregion",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
ecoregion_path <- list.files(
  directory_to_save,
  pattern = "\\.(gpkg|shp)$",
  recursive = TRUE,
  full.names = TRUE
)[1]
processed_data <- process_covariates(
  covariate = "ecoregion",
  path = ecoregion_path
)
plot(processed_data, main = "Level 3 Ecoregions of the Conterminous United States")
```

## Calculate dummy (bivariate) covariates at points

`calculate_ecoregion()` by default returns binary (dummy) indicator columns for the
intersecting level-2 and level-3 ecoregions at each location. By default
(`colnames = "coded"`), those indicators use key-based names (for example
`DUM_E2083_00000`, `DUM_E3064_00000`). If you set
`colnames = "full_ecoregion"`, indicators use sanitized full names (for
example `DUM_E2_SOUTHEASTERN_USA_PLAINS_00000`), with duplicate names
automatically disambiguated by suffixes like `_1`.

```{r calculate-points, eval = live_run}

sf_process <- sf::st_as_sf(processed_data)

example_points_terra <- terra::spatSample(
  processed_data,
  size = 25,
  method = "random"
)
 example_points_sf <- sf::st_as_sf(example_points_terra)
 example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))



point_values <- calculate_ecoregion(
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 10000,
  geom = "sf"
)

print(point_values)

```

## Calculate fractional covariates at points

`calculate_ecoregion()` can also return fractional (proportion) columns for the
intersecting level-2 and level-3 ecoregions at each location. By default
(`colnames = "coded"`), those indicators use key-based names (for example
`FRC_E2083_00000`, `FRC_E3064_00000`). If you set
`colnames = "full_ecoregion"`, indicators use sanitized full names (for
example `FRC_E2_SOUTHEASTERN_USA_PLAINS_00000`), with duplicate names
automatically disambiguated by suffixes like `_1`.

```{r calculate-points-fractional, eval = live_run}
point_values <- calculate_ecoregion(
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  geom = "sf",
  radius = 1000,
  frac = TRUE,
  drop = TRUE
)

print(point_values)

```
