---
title: "EDGAR Emissions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EDGAR Emissions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  preferred <- candidates[vapply(x[candidates], function(col) {
    is.numeric(col) || is.character(col) || is.factor(col)
  }, logical(1))]
  if (length(preferred) > 0) preferred[1] else candidates[1]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::scale_color_viridis_c(
      trans = log10_zero,
      option = "plasma",
      na.value = "grey70"
    ) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::guides(
      color = ggplot2::guide_colorbar(
        barwidth  = ggplot2::unit(0.4, "npc"),
        barheight = ggplot2::unit(0.4, "lines"),
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      legend.position   = "bottom",
      legend.title      = ggplot2::element_text(size = 8),
      legend.text       = ggplot2::element_text(size = 7),
      plot.title        = ggplot2::element_text(size = 10, hjust = 0.5),
      plot.margin       = ggplot2::margin(4, 4, 4, 4)
    )
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::scale_fill_viridis_c(
      trans = log10_zero,
      option = "plasma",
      na.value = "grey70"
    ) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::guides(
      fill = ggplot2::guide_colorbar(
        barwidth  = ggplot2::unit(0.4, "npc"),
        barheight = ggplot2::unit(0.4, "lines"),
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      legend.position   = "bottom",
      legend.title      = ggplot2::element_text(size = 8),
      legend.text       = ggplot2::element_text(size = 7),
      plot.title        = ggplot2::element_text(size = 10, hjust = 0.5),
      plot.margin       = ggplot2::margin(4, 4, 4, 4)
    )
}

# log10 transform that floors zeros (and negatives) to 0 instead of -Inf
log10_zero <- scales::trans_new(
  name      = "log10_zero",
  transform = function(x) ifelse(is.na(x) | x <= 0, 0, log10(x)),
  inverse   = function(x) 10^x,
  breaks    = scales::log_breaks(base = 10),
  domain    = c(0, Inf)
)
```

This article demonstrates a compact workflow for EDGAR emissions surfaces supported by `amadeus`.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available EDGAR variables and information

The `species` argument in `download_data(dataset_name = "edgar", ...)`
currently supports the following EDGAR air-pollutant emissions variables.
For the emissions downloads used in this vignette, values are reported in
tonnes.

```{r edgar-variable-reference, echo = FALSE}
edgar_variables <- data.frame(
  variable = c("BC", "CO", "NH3", "NMVOC", "NOx", "OC", "PM10", "PM2.5", "SO2"),
  description = c(
    "Black carbon (carbonaceous particulate fraction)",
    "Carbon monoxide",
    "Ammonia",
    "Non-methane volatile organic compounds",
    "Nitrogen oxides",
    "Organic carbon (carbonaceous particulate fraction)",
    "Particulate matter <= 10 um aerodynamic diameter",
    "Particulate matter <= 2.5 um aerodynamic diameter",
    "Sulfur dioxide"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  edgar_variables,
  col.names = c("amadeus variable name", "short description"),
  align = c("l", "l")
)
```

`download_edgar()` also exposes several other useful switches that determine
which EDGAR products are available:

- `version = "8.1"` covers the standard EDGAR air-pollutant products shown in
  this vignette.
- `version = "8.1_voc"` switches to VOC speciation products instead of the
  standard pollutant grids.
- `temp_res` supports `"yearly"`, `"monthly"`, and `"timeseries"` for
  `version = "8.1"`. VOC speciation downloads do not use `temp_res`.
- `output` supports `"emi"` for emissions and `"flx"` for flux products.
- `format` is typically `"nc"` or `"txt"`, but monthly outputs and flux
  outputs are only available in NetCDF form.
- `year_range` can be a single year or a range. Yearly EDGAR products span
  `1970-2022`, while monthly and VOC-speciation products span `2000-2022`.

The sector arguments depend on which EDGAR product family you request:

```{r edgar-option-reference, echo = FALSE}
edgar_options <- data.frame(
  input = c(
    "sector_yearly",
    "sector_monthly",
    "sector_voc",
    "voc"
  ),
  supported_values = c(
    paste(
      c(
        "AGS", "AWB", "CHE", "ENE", "IND", "MNM", "NMM", "PRU_SOL",
        "RCO", "REF_TRF", "SWD_INC", "SWD_LDF", "TNR_Aviation_CDS",
        "TNR_Aviation_CRS", "TNR_Aviation_LTO", "TNR_Aviation_SPS",
        "TNR_Other", "TNR_Ship", "TRO", "WWT"
      ),
      collapse = ", "
    ),
    paste(
      c(
        "AGRICULTURE", "BUILDINGS", "FUEL_EXPLOITATION",
        "IND_COMBUSTION", "IND_PROCESSES", "POWER_INDUSTRY",
        "TRANSPORT", "WASTE"
      ),
      collapse = ", "
    ),
    paste(
      c(
        "AGRICULTURE", "BUILDINGS", "FUEL_EXPLOITATION",
        "IND_COMBUSTION", "IND_PROCESSES", "POWER_INDUSTRY",
        "TRANSPORT", "WASTE"
      ),
      collapse = ", "
    ),
    "Integers 1-25 for VOC speciation groups"
  ),
  when_used = c(
    "Yearly standard EDGAR products (`version = \"8.1\"`)",
    "Monthly standard EDGAR products (`version = \"8.1\"`)",
    "VOC speciation products (`version = \"8.1_voc\"`)",
    "VOC speciation products (`version = \"8.1_voc\"`)"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  edgar_options,
  col.names = c("input", "supported values", "when used"),
  align = c("l", "l", "l")
)
```

EDGAR groups these variables broadly as ozone precursors (`CO`, `NOx`, `NMVOC`), acidifying gases (`NH3`, `NOx`, `SO2`), and primary particulates (`PM10`, `PM2.5`, `BC`, `OC`).

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "edgar_workflow")
download_data(
  dataset_name = "edgar",
  species = c("CO", "NOx"),
  temp_res = "yearly",
  sector_yearly = "ENE",
  year_range = 2021,
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
```

## Process one workflow-ready data product

Amadeus process functions returns a `terra::SpatRaster`  so we can apply the terra extent option on the bounding box of the points to reduce processing time.

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "edgar",
  path = list.files(
    directory_to_save,
    pattern = "\\.(tif|tiff|nc4?|grd|img)$",
    recursive = TRUE,
    full.names = TRUE
  ),
  extent = sf::st_bbox(example_points_sf)
)
```

## Calculate covariates at points

We start with a radius of 0, which extracts the raster gric cell value at each point with no smoothing or averaging over an area.

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "edgar",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  geom = "sf"
)

print(point_values)
```

At point-level extractions, we often see more NA because data has to overlap exactly. Here, we smooth out the covariate calculation by using a buffer radius.

```{r calculate-points-buffer, eval = live_run}
point_values_1km <- calculate_covariates(
  covariate = "edgar",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 1000, # distance units are in meters, see crs(processed_data)
  geom = "sf"
)

print(point_values_1km)

point_values_10km <- calculate_covariates(
  covariate = "edgar",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 10000,
  geom = "sf"
)

print(point_values_10km)

```
## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "edgar",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("EDGAR Emissions", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("EDGAR Emissions", ": Durham H3 polygons"))
```


## Notes

- This example requests two species (`CO` and `NOx`) to exercise multi-file EDGAR downloads.
- Geometry-returning EDGAR extractions work cleanly for both points and polygons.
