---
title: "NASA GEOS-CF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA GEOS-CF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)

geos_sample_path <- c(
  file.path("tests", "testdata", "geos", "c"),
  file.path("..", "tests", "testdata", "geos", "c")
)
geos_sample_path <- geos_sample_path[dir.exists(geos_sample_path)][1]
if (is.na(geos_sample_path) || geos_sample_path == "") {
  stop("Could not locate tests/testdata/geos/c")
}


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  numeric_candidates <- candidates[vapply(x[candidates], is.numeric, logical(1))]
  ranked <- if (length(numeric_candidates) > 0) numeric_candidates else candidates
  non_missing_share <- vapply(ranked, function(col) {
    mean(!is.na(x[[col]]))
  }, numeric(1))
  ranked[which.max(non_missing_share)]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::guides(
      color = ggplot2::guide_colorbar(title.position = "top")
    ) +
    ggplot2::theme_minimal(base_size = 13) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.key.width = grid::unit(2, "cm"),
      legend.title = ggplot2::element_text(face = "bold")
    )
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::guides(
      fill = ggplot2::guide_colorbar(title.position = "top")
    ) +
    ggplot2::theme_minimal(base_size = 13) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.key.width = grid::unit(2, "cm"),
      legend.title = ggplot2::element_text(face = "bold")
    )
}

summarize_na_counts <- function(x, value_pattern) {
  data <- sf::st_drop_geometry(sf::st_as_sf(x))
  value_cols <- grep(value_pattern, names(data), value = TRUE)
  if (length(value_cols) == 0) {
    fallback_excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", "time")
    fallback_cols <- setdiff(names(data), fallback_excluded)
    value_cols <- fallback_cols[vapply(data[fallback_cols], is.numeric, logical(1))]
  }
  if (length(value_cols) == 0) {
    stop("No covariate columns found for missing-value summary.")
  }
  data.frame(
    variable = value_cols,
    missing_values = vapply(data[value_cols], function(col) sum(is.na(col)), integer(1)),
    stringsAsFactors = FALSE
  )
}
```

This article demonstrates a compact workflow for NASA GEOS-CF data. GEOS-CF downloads require a NASA EarthData token; see `protected_datasets` for setup details.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "geos", ...)` wraps `download_geos()`.

- `collection` accepts six hourly GEOS-CF global NetCDF collections: `aqc_tavg_1hr_g1440x721_v1`, `chm_tavg_1hr_g1440x721_v1`, `met_tavg_1hr_g1440x721_x1`, `xgc_tavg_1hr_g1440x721_x1`, `chm_inst_1hr_g1440x721_p23`, and `met_inst_1hr_g1440x721_p23`.
- `date` can be a single day or a start/end range; each requested day expands to hourly `.nc4` files for every selected collection.
- Standard downloads retrieve the full 1440×721 global grid; study-area clipping usually happens later in `process_covariates()`.
- NASA EarthData authentication is required; `amadeus` reads `NASA_EARTHDATA_TOKEN` or accepts a token explicitly.
- The wrapper validates the first requested URL before downloading and stops early when a requested date is unavailable.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "geos_workflow")
geos_live_extent <- c(-79.2, 35.8, -78.6, 36.3)

geos_probe_url <- paste0(
  "https://portal.nccs.nasa.gov/datashare/gmao/geos-cf/v1/ana/",
  "Y2019/M09/D09/GEOS-CF.v01.rpl.chm_inst_1hr_g1440x721_p23.20190909_0000z.nc4"
)
geos_live_probe <- tryCatch({
  resp <- httr2::request(geos_probe_url) |>
    httr2::req_method("HEAD") |>
    httr2::req_timeout(20) |>
    httr2::req_perform()
  httr2::resp_status(resp)
}, error = function(e) {
  NA_integer_
})
geos_live_probe

geos_live_files <- character(0)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
geos_process_path <- if (exists("geos_live_files") && length(geos_live_files) > 0) {
  dirname(geos_live_files[1])
} else {
  geos_sample_path
}
geos_process_date <- if (exists("geos_live_files") && length(geos_live_files) > 0) {
  "2019-09-09"
} else {
  "2018-01-01"
}

processed_data <- process_covariates(
  covariate = "geos",
  variable = "O3",
  date = geos_process_date,
  path = geos_process_path,
  extent = terra::ext(
    geos_live_extent[1], geos_live_extent[3],
    geos_live_extent[2], geos_live_extent[4]
  )
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "geos",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "geos",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

point_na_summary <- summarize_na_counts(point_values, "^O3")
polygon_na_summary <- summarize_na_counts(polygon_values, "^O3")
knitr::kable(point_na_summary, col.names = c("Variable", "Point missing values"))
knitr::kable(polygon_na_summary, col.names = c("Variable", "Polygon missing values"))
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("NASA GEOS-CF", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("NASA GEOS-CF", ": Durham H3 polygons"))
```

## Notes

- The live download chunk validates authenticated GEOS-CF requests with `download = FALSE`; downstream processing uses downloaded files when present and otherwise falls back to packaged GEOS test files for reproducible rendering.
