---
title: "NASA GEOS-CF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA GEOS-CF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  preferred <- candidates[vapply(x[candidates], function(col) {
    is.numeric(col) || is.character(col) || is.factor(col)
  }, logical(1))]
  if (length(preferred) > 0) preferred[1] else candidates[1]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::theme_minimal()
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::theme_minimal()
}
```

This article demonstrates a compact workflow for NASA GEOS-CF data. GEOS-CF downloads require a NASA EarthData token; see `protected_datasets` for setup details.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "geos", ...)` wraps `download_geos()`.

- `collection` accepts six hourly GEOS-CF global NetCDF collections: `aqc_tavg_1hr_g1440x721_v1`, `chm_tavg_1hr_g1440x721_v1`, `met_tavg_1hr_g1440x721_x1`, `xgc_tavg_1hr_g1440x721_x1`, `chm_inst_1hr_g1440x721_p23`, and `met_inst_1hr_g1440x721_p23`.
- `date` can be a single day or a start/end range; each requested day expands to hourly `.nc4` files for every selected collection.
- Standard downloads retrieve the full 1440×721 global grid; study-area clipping usually happens later in `process_covariates()`. Set `use_opendap = TRUE` with an `extent` to request a server-side spatial subset and reduce transfer size.
- NASA EarthData authentication is required; `amadeus` reads `NASA_EARTHDATA_TOKEN` or accepts a token explicitly.
- The wrapper validates the first requested URL before downloading and stops early when a requested date is unavailable.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "geos_workflow")
download_data(
  dataset_name = "geos",
  collection = c("chm_inst_1hr_g1440x721_p23", "met_inst_1hr_g1440x721_p23"),
  date = "2019-09-09",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```

## Download with OPeNDAP — server-side spatial and variable subsetting

OPeNDAP enables server-side subsetting for GEOS-CF, so only the grid cells and
variables you request are transferred instead of the full 1440×721 global grid.
Set `use_opendap = TRUE`, provide an `extent` (`c(xmin, ymin, xmax, ymax)` in
WGS-84 degrees), and optionally a `variables` vector. Authentication is handled
by the same `NASA_EARTHDATA_TOKEN` environment variable used for standard
downloads.

The example below retrieves ozone (`O3`) and nitrogen dioxide (`NO2`) from the
`chm_inst_1hr_g1440x721_p23` collection over the US Mid-Atlantic for a single
day — a small fraction of the full-globe file.

```{r geos-opendap, eval = live_run}
directory_to_save_opendap <- file.path(tempdir(), "geos_opendap")
download_data(
  dataset_name = "geos",
  collection = "chm_inst_1hr_g1440x721_p23",
  date = "2019-09-09",
  directory_to_save = directory_to_save_opendap,
  acknowledgement = TRUE,
  use_opendap = TRUE,
  extent = c(-85, 33, -75, 42),
  variables = c("O3", "NO2")
)
list.files(directory_to_save_opendap, pattern = "\\.nc4?$", recursive = TRUE)
```

The downloaded file is a NetCDF4 spatial subset that can be processed with
`process_covariates()` just like a full-globe GEOS-CF file. Downstream
processing clips to `extent` as usual.

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "geos",
  variable = "O3",
  date = "2019-09-09",
  path = dirname(list.files(
    directory_to_save,
    pattern = "\\.nc4?$",
    recursive = TRUE,
    full.names = TRUE
  )[1]),
  extent = terra::ext(-79.2, -78.6, 35.8, 36.3)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "geos",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "geos",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("NASA GEOS-CF", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("NASA GEOS-CF", ": Durham H3 polygons"))
```

## Notes

- The download chunk requests both a chemistry and a meteorology collection so local live runs exercise more than one GEOS-CF collection.
- GEOS-CF examples can create many hourly files, so Durham-sized extents keep local inspection manageable.
