---
title: "NASA GEOS-CF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA GEOS-CF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 9,
  fig.height = 7,
  out.width = "100%",
  fig.align = "center"
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

geos_sample_path <- c(
  file.path("tests", "testdata", "geos", "c"),
  file.path("..", "tests", "testdata", "geos", "c")
)
geos_sample_path <- geos_sample_path[dir.exists(geos_sample_path)][1]
if (is.na(geos_sample_path) || geos_sample_path == "") {
  stop("Could not locate tests/testdata/geos/c")
}
```

This article demonstrates a compact workflow for NASA GEOS-CF data. GEOS-CF downloads require a NASA EarthData token; see `protected_datasets` for setup details.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two compact example surfaces created from the processed GEOS extent: two points for point extraction and a small hexagon grid for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "geos", ...)` wraps `download_geos()`.

- `collection` accepts six hourly GEOS-CF global NetCDF collections: `aqc_tavg_1hr_g1440x721_v1`, `chm_tavg_1hr_g1440x721_v1`, `met_tavg_1hr_g1440x721_x1`, `xgc_tavg_1hr_g1440x721_x1`, `chm_inst_1hr_g1440x721_p23`, and `met_inst_1hr_g1440x721_p23`.
- `date` can be a single day or a start/end range; each requested day expands to hourly `.nc4` files for every selected collection.
- Standard downloads retrieve the full 1440×721 global grid; study-area clipping usually happens later in `process_covariates()`.
- NASA EarthData authentication is required; `amadeus` reads `NASA_EARTHDATA_TOKEN` or accepts a token explicitly.
- The wrapper validates the first requested URL before downloading and stops early when a requested date is unavailable.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "geos_workflow")
geos_live_extent <- c(-79.2, 35.8, -78.6, 36.3)
geos_demo_collection <- "aqc_tavg_1hr_g1440x721_v1"
geos_demo_date <- "2019-09-09"
geos_download_error <- NULL


download_data(
  dataset_name = "geos",
  collection = geos_demo_collection,
  date = geos_demo_date,
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)

geos_live_files <- list.files(
  file.path(directory_to_save, geos_demo_collection),
  pattern = "\\.nc4$",
  recursive = TRUE,
  full.names = TRUE
)
geos_live_ready <- length(geos_live_files) > 0
if (!geos_live_ready) {
  message(
    "Using packaged GEOS test files for downstream chunks because live files ",
    "are unavailable."
  )
}
```

## Process one workflow-ready data product

```{r process, eval = live_run}
geos_process_path <- if (isTRUE(geos_live_ready)) {
  file.path(directory_to_save, geos_demo_collection)
} else {
  geos_sample_path
}
geos_process_date <- if (isTRUE(geos_live_ready)) {
  geos_demo_date
} else {
  "2018-01-01"
}

processed_data <- process_covariates(
  covariate = "geos",
  variable = "O3",
  date = geos_process_date,
  path = geos_process_path,
  extent = terra::ext(
    geos_live_extent[1], geos_live_extent[3],
    geos_live_extent[2], geos_live_extent[4]
  )
)

processed_extent <- terra::ext(processed_data)
processed_bbox <- sf::st_bbox(
  c(
    xmin = processed_extent[1],
    ymin = processed_extent[3],
    xmax = processed_extent[2],
    ymax = processed_extent[4]
  ),
  crs = sf::st_crs(4326)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
domain_x <- c(processed_bbox[["xmin"]], processed_bbox[["xmax"]])
domain_y <- c(processed_bbox[["ymin"]], processed_bbox[["ymax"]])
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
candidate_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
candidate_values <- terra::extract(
  processed_data[[1]],
  terra::vect(candidate_points_sf)
)[, 2]
valid_points <- candidate_points_sf[!is.na(candidate_values), ]
if (nrow(valid_points) < 2) {
  stop("Could not find at least two in-domain points with GEOS values.")
}
example_points_sf <- valid_points[1:2, ]
example_points_sf$site_id <- sprintf("geos_point_%d", seq_len(nrow(example_points_sf)))

point_values <- calculate_covariates(
  covariate = "geos",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Calculate covariates across in-domain hexagons

```{r calculate-polygons, eval = live_run}
processed_domain <- sf::st_as_sfc(processed_bbox)
hex_size <- min(domain_dx, domain_dy) / 5
hex_grid <- sf::st_make_grid(processed_domain, cellsize = hex_size, square = FALSE)
hex_sf <- sf::st_sf(geometry = hex_grid)
hex_centroids <- sf::st_centroid(hex_sf)
hex_values <- terra::extract(processed_data[[1]], terra::vect(hex_centroids))[, 2]
example_hex_sf <- hex_sf[!is.na(hex_values), ]
if (nrow(example_hex_sf) == 0) {
  stop("Could not build in-domain hexagons with GEOS values.")
}
example_hex_sf$h3_id <- sprintf("geos_hex_%03d", seq_len(nrow(example_hex_sf)))

polygon_values <- calculate_covariates(
  covariate = "geos",
  from = processed_data,
  locs = example_hex_sf,
  locs_id = "h3_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

point_data <- sf::st_drop_geometry(point_values)
point_value_cols <- grep("^O3", names(point_data), value = TRUE)
if (length(point_value_cols) == 0) {
  point_value_cols <- names(point_data)[vapply(point_data, is.numeric, logical(1))]
}
point_na_summary <- data.frame(
  variable = point_value_cols,
  missing_values = vapply(point_data[point_value_cols], function(col) sum(is.na(col)), integer(1)),
  stringsAsFactors = FALSE
)

polygon_data <- sf::st_drop_geometry(polygon_values)
polygon_value_cols <- grep("^O3", names(polygon_data), value = TRUE)
if (length(polygon_value_cols) == 0) {
  polygon_value_cols <- names(polygon_data)[vapply(polygon_data, is.numeric, logical(1))]
}
polygon_na_summary <- data.frame(
  variable = polygon_value_cols,
  missing_values = vapply(
    polygon_data[polygon_value_cols],
    function(col) sum(is.na(col)),
    integer(1)
  ),
  stringsAsFactors = FALSE
)

knitr::kable(point_na_summary, col.names = c("Variable", "Point missing values"))
knitr::kable(polygon_na_summary, col.names = c("Variable", "Polygon missing values"))
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
point_plot_col <- grep("^O3", names(point_values), value = TRUE)[1]
if (is.na(point_plot_col) || point_plot_col == "") {
  excluded <- c("site_id", "h3_id", attr(point_values, "sf_column"))
  fallback_cols <- setdiff(names(point_values), excluded)
  fallback_numeric <- fallback_cols[vapply(point_values[fallback_cols], is.numeric, logical(1))]
  point_plot_col <- fallback_numeric[1]
}

point_bbox <- sf::st_bbox(point_values)
point_pad_x <- max((point_bbox[["xmax"]] - point_bbox[["xmin"]]) * 0.08, 0.03)
point_pad_y <- max((point_bbox[["ymax"]] - point_bbox[["ymin"]]) * 0.08, 0.03)

ggplot2::ggplot(point_values) +
  ggplot2::geom_sf(
    ggplot2::aes(color = .data[[point_plot_col]]),
    size = 2.2
  ) +
  ggplot2::coord_sf(
    xlim = c(point_bbox[["xmin"]] - point_pad_x, point_bbox[["xmax"]] + point_pad_x),
    ylim = c(point_bbox[["ymin"]] - point_pad_y, point_bbox[["ymax"]] + point_pad_y),
    expand = FALSE,
    datum = NA
  ) +
  ggplot2::scale_color_viridis_c(option = "C") +
  ggplot2::labs(
    title = "NASA GEOS-CF: point extraction",
    color = point_plot_col
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(2, "cm"),
    legend.title = ggplot2::element_text(face = "bold")
  )
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across in-domain hexagons for this workflow."}
polygon_plot_col <- grep("^O3", names(polygon_values), value = TRUE)[1]
if (is.na(polygon_plot_col) || polygon_plot_col == "") {
  excluded <- c("site_id", "h3_id", attr(polygon_values, "sf_column"))
  fallback_cols <- setdiff(names(polygon_values), excluded)
  fallback_numeric <- fallback_cols[vapply(polygon_values[fallback_cols], is.numeric, logical(1))]
  polygon_plot_col <- fallback_numeric[1]
}

polygon_bbox <- sf::st_bbox(polygon_values)
polygon_pad_x <- max((polygon_bbox[["xmax"]] - polygon_bbox[["xmin"]]) * 0.08, 0.03)
polygon_pad_y <- max((polygon_bbox[["ymax"]] - polygon_bbox[["ymin"]]) * 0.08, 0.03)

ggplot2::ggplot(polygon_values) +
  ggplot2::geom_sf(
    ggplot2::aes(fill = .data[[polygon_plot_col]]),
    color = NA
  ) +
  ggplot2::coord_sf(
    xlim = c(polygon_bbox[["xmin"]] - polygon_pad_x, polygon_bbox[["xmax"]] + polygon_pad_x),
    ylim = c(polygon_bbox[["ymin"]] - polygon_pad_y, polygon_bbox[["ymax"]] + polygon_pad_y),
    expand = FALSE,
    datum = NA
  ) +
  ggplot2::scale_fill_viridis_c(option = "C") +
  ggplot2::labs(
    title = "NASA GEOS-CF: in-domain hexagon extraction",
    fill = polygon_plot_col
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(2, "cm"),
    legend.title = ggplot2::element_text(face = "bold")
  )
```

## Notes

- The live download chunk uses `download_data(dataset_name = "geos", collection = "aqc_tavg_1hr_g1440x721_v1", ...)`, which dispatches to `download_geos()`. Downstream processing uses downloaded files when present and otherwise falls back to packaged GEOS test files for reproducible rendering.
