---
title: "USGS GMTED2010"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{USGS GMTED2010}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


```

This article demonstrates a compact workflow for GMTED2010 terrain products.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

- `download_data(dataset_name = "gmted", ...)` accepts `statistic` values of `"Breakline Emphasis"`, `"Systematic Subsample"`, `"Median Statistic"`, `"Minimum Statistic"`, `"Mean Statistic"`, `"Maximum Statistic"`, and `"Standard Deviation Statistic"`.
- Available `resolution` values are `"7.5 arc-seconds"`, `"15 arc-seconds"`, and `"30 arc-seconds"`.
- GMTED2010 is a static terrain reference, so the wrapper has no year or scenario arguments; downloads arrive as zipped elevation grid bundles.
- Pick the statistic intentionally: each layer represents a different summary of elevation, so values from different statistics should not be treated as interchangeable.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "gmted_workflow")
download_data(
  dataset_name = "gmted",
  statistic = "Breakline Emphasis",
  resolution = "7.5 arc-seconds",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
download_data(
  dataset_name = "gmted",
  statistic = "Mean Statistic",
  resolution = "15 arc-seconds",
  directory_to_save = file.path(directory_to_save, "mean_15s"),
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
gmted_dirs <- list.dirs(directory_to_save, recursive = TRUE, full.names = TRUE)
gmted_path <- gmted_dirs[grepl("be75_grd$", gmted_dirs)][1]
processed_data <- process_covariates(
  covariate = "gmted",
  variable = c("Breakline Emphasis", "7.5 arc-seconds"),
  path = gmted_path,
  extent = terra::ext(-79.2, -78.6, 35.8, 36.3)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "gmted",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "gmted",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("USGS GMTED2010", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("USGS GMTED2010", ": Durham H3 polygons"))
```

## Notes

- GMTED download requests are shown for two statistic/resolution combinations so the vignette covers more than one terrain surface.
- Processing still focuses on the 7.5 arc-second breakline product because it is the most detailed of the two example requests.
