---
title: "Climatology Lab gridMET"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Climatology Lab gridMET}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)
library(ggplot2)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

```

This article demonstrates a compact workflow for Climatology Lab `gridMET` data using multiple variables, multiple summaries, and both point and polygon extraction.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

`download_data(dataset_name = "gridmet", ...)` accepts either the full variable names below or their gridMET codes.

| Code | Variable |
| --- | --- |
| `sph` | Near-Surface Specific Humidity |
| `vpd` | Mean Vapor Pressure Deficit |
| `pr` | Precipitation |
| `rmin` | Minimum Near-Surface Relative Humidity |
| `rmax` | Maximum Near-Surface Relative Humidity |
| `srad` | Surface Downwelling Solar Radiation |
| `tmmn` | Minimum Near-Surface Air Temperature |
| `tmmx` | Maximum Near-Surface Air Temperature |
| `vs` | Wind speed at 10 m |
| `th` | Wind direction at 10 m |
| `pdsi` | Palmer Drought Severity Index |
| `pet` | Reference grass evapotranspiration |
| `etr` | Reference alfalfa evapotranspiration |
| `ERC` | Energy Release Component |
| `BI` | Burning Index |
| `FM100` | 100-hour dead fuel moisture |
| `FM1000` | 1000-hour dead fuel moisture |

- Temporal resolution: daily; each download is an annual NetCDF file for one variable.
- Year input: use a single year or a start/end pair such as `c(2018, 2022)`.
- Availability check: the wrapper validates the first requested variable-year URL and stops if that request returns HTTP 404.
- Major constraint: gridMET downloads do not require authentication.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "gridmet_workflow")
download_data(
  dataset_name = "gridmet",
  variables = c("tmmx"),
  year = 2020,
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```

## Process workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "gridmet",
  variable = "tmmx",
  date = c("2020-01-01", "2020-01-31"),
  path = file.path(directory_to_save, "tmmx"),
  extent = terra::ext(-79.2, -78.6, 35.8, 36.3)
)

```

## Calculate covariates at points using native daily temporal resolution and use .by_time to calculate monthly means

```{r calculate-points-daily, eval = live_run}


domain_x <- c(terra::xmin(processed_data), terra::xmax(processed_data))
domain_y <- c(terra::ymin(processed_data), terra::ymax(processed_data))
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
example_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))


point_values <- calculate_covariates(
  covariate = "gridmet",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

point_values_month <- calculate_covariates(
  covariate = "gridmet",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  .by_time = "month",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

print(point_values)
print(point_values_month)

```

## Visualize the point outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}

ggplot(data = point_values) + geom_sf(aes(color = `tmmx_0`)) + ggtitle("Daily values")
```
