---
title: "Climatology Lab gridMET"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Climatology Lab gridMET}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  preferred <- candidates[vapply(x[candidates], function(col) {
    is.numeric(col) || is.character(col) || is.factor(col)
  }, logical(1))]
  if (length(preferred) > 0) preferred[1] else candidates[1]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::theme_minimal()
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::theme_minimal()
}
```

This article demonstrates a compact workflow for Climatology Lab `gridMET` data using multiple variables, multiple summaries, and both point and polygon extraction.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "gridmet", ...)` accepts either the full variable names below or their gridMET codes.

| Code | Variable |
| --- | --- |
| `sph` | Near-Surface Specific Humidity |
| `vpd` | Mean Vapor Pressure Deficit |
| `pr` | Precipitation |
| `rmin` | Minimum Near-Surface Relative Humidity |
| `rmax` | Maximum Near-Surface Relative Humidity |
| `srad` | Surface Downwelling Solar Radiation |
| `tmmn` | Minimum Near-Surface Air Temperature |
| `tmmx` | Maximum Near-Surface Air Temperature |
| `vs` | Wind speed at 10 m |
| `th` | Wind direction at 10 m |
| `pdsi` | Palmer Drought Severity Index |
| `pet` | Reference grass evapotranspiration |
| `etr` | Reference alfalfa evapotranspiration |
| `ERC` | Energy Release Component |
| `BI` | Burning Index |
| `FM100` | 100-hour dead fuel moisture |
| `FM1000` | 1000-hour dead fuel moisture |

- Temporal resolution: daily; each download is an annual NetCDF file for one variable.
- Year input: use a single year or a start/end pair such as `c(2018, 2022)`.
- Availability check: the wrapper validates the first requested variable-year URL and stops if that request returns HTTP 404.
- Major constraint: gridMET downloads do not require authentication.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "gridmet_workflow")
download_data(
  dataset_name = "gridmet",
  variables = c("tmmx", "pr"),
  year = 2020,
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "gridmet",
  variable = "tmmx",
  date = c("2020-01-01", "2020-01-31"),
  path = file.path(directory_to_save, "tmmx"),
  extent = terra::ext(-79.2, -78.6, 35.8, 36.3)
)

processed_data <- process_covariates(
  covariate = "gridmet",
  variable = "tmmx",
  date = c("2020-01-01", "2020-01-31"),
  path = file.path(directory_to_save, "tmmx"),
  extent = modis_extent
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "gridmet",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

point_values_month <- calculate_covariates(
  covariate = "gridmet",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  .by_time = "month",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

point_values_file_month <- calculate_covariates(
  covariate = "gridmet",
  from = file.path(directory_to_save, "tmmx"),
  locs = example_points_sf,
  locs_id = "site_id",
  .by_time = "month",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

point_values_sum <- calculate_covariates(
  covariate = "gridmet",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "sum",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "gridmet",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  fun = "sum",
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("Climatology Lab gridMET", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("Climatology Lab gridMET", ": Durham H3 polygons"))
```

## Notes

- The download chunk requests two gridMET variables so local runs exercise a true multi-variable request.
- Point extraction demonstrates both `mean` and `sum`, while the polygon example uses `sum` to summarize accumulated precipitation over each H3 cell.
