---
title: "NOAA HMS Smoke"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NOAA HMS Smoke}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})


```

This article demonstrates a compact workflow for NOAA HMS smoke plume polygons.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

`download_hms()` exposes the main choices that determine what NOAA smoke files are available:

- `date` accepts a single day or a start/end range in `YYYY-MM-DD` format, with availability beginning on `2005-08-05`.
- HMS smoke products are daily files, so multi-day requests download one plume file per day in the requested window.
- `data_format` supports `"Shapefile"` and `"KML"`.
- Shapefile requests download zipped archives that can be unzipped automatically; KML requests download raw `.kml` files directly.
- HMS does not require authentication, and availability is determined by the NOAA daily smoke-product archive for the dates requested.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "hms_workflow")
download_data(
  dataset_name = "hms",
  date = c("2022-06-10", "2022-06-13"),
  data_format = "Shapefile",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)

```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "hms",
  date = c("2022-06-10", "2022-06-13"),
  path = dirname(list.files(
    directory_to_save,
    pattern = "\\.shp$",
    recursive = TRUE,
    full.names = TRUE
  )[1]),
  extent = terra::ext(-124.8, -116.4, 41.9, 46.3)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}

domain_x <- c(terra::xmin(processed_data), terra::xmax(processed_data))
domain_y <- c(terra::ymin(processed_data), terra::ymax(processed_data))
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
example_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))

point_values <- calculate_covariates(
  covariate = "hms",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 5000,
  geom = "sf"
)

```


## Visualize the point  outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}

sf_process <- sf::st_as_sf(processed_data)
 # create 1-km buffers from points (in a projected CRS, then back to WGS84)
 buffers_5km <- example_points_sf |>
   st_transform(3857) |>
   st_buffer(dist = 5000) |>
   st_transform(4326)

 # join extracted values so fill matches each site
 buffers_5km <- merge(
   buffers_5km,
   st_drop_geometry(point_values)[, c("site_id", "light_05000", "medium_05000")],
   by = "site_id",
   all.x = TRUE
 )
 buffers_5km$smoke_class <- ifelse(
   is.na(buffers_5km$medium_05000), "No smoke",
   ifelse(
     buffers_5km$medium_05000 > 0, "Medium smoke",
     ifelse(buffers_5km$light_05000 > 0, "Light smoke", "No smoke")
   )
 )
 buffers_5km$smoke_class <- factor(
   buffers_5km$smoke_class,
   levels = c("No smoke", "Light smoke", "Medium smoke")
 )

  ggplot() +
    geom_sf(data = sf_process, fill = NA, color = "grey40") +
    geom_sf(data = buffers_5km, aes(fill = smoke_class), alpha = 0.8) +
   scale_fill_manual(
     values = c(
       "No smoke" = "#f7f7f7",
       "Light smoke" = "#fddbc7",
       "Medium smoke" = "#ef8a62"
     ),
     name = "Smoke category"
    ) +
    ggtitle("Light/medium smoke within 5km of points") +
    theme(legend.position = "bottom")

```
