---
title: "USGS Hydrologic Unit Codes (HUC)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{USGS Hydrologic Unit Codes (HUC)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)
library(terra)
library(ggplot2)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})


```

This article demonstrates a compact workflow for Hydrologic Unit Code boundaries.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

- `download_data(dataset_name = "huc", ...)` accepts `region = "Lower48"` or `"Islands"`; the `"Islands"` option covers Hawaii, Puerto Rico, and the Virgin Islands.
- The `type` argument is `"Seamless"` or `"OceanCatchment"` for the two national NHDPlus v2.1 delivery types exposed by the wrapper.
- HUC data are static reference boundaries rather than time-varying products, and downloads arrive as large `.7z` geodatabase archives; automatic unzip is not supported by this wrapper.
- For HUC boundaries, the wrapper docs note that `type = "Seamless"` is the right choice because it contains the HUC12 layer, which can then be aggregated to HUC6, HUC8, HUC10, and similar levels.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "huc_workflow")
download_data(
  dataset_name = "huc",
  region = "Islands",
  type = "Seamless",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE
)

```

## Process one workflow-ready data product

```{r process, eval = live_run}

huc_path <- 
  paste0(directory_to_save,"/NHDPlusNationalData/",
  "NHDPlusV21_National_Seamless_Flattened_HI_PR_VI_PI.gdb")

huc_field <- "HUC12"
layers <- sf::st_layers(huc_path)
processed_data <- process_covariates(
  covariate = "huc",
  path = huc_path,
  layer_name = "HUC12",
  extent =  terra::ext(-162.5, -153.5, 18, 23)
)

plot(processed_data, main = "HUC12 boundaries for Hawaii")
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}

df <- data.frame(
  site_id = c("site_1", "site_2", "site_3","site_4"),
  lon = c(-157.8583, -155.5319, -155.5828,-159.5),
  lat = c(21.3069, 19.5, 19.8968, 22.0)
)
example_points_sf <- sf::st_as_sf(
  df,
  coords = c("lon", "lat"),
  crs = 4326
)


point_values <- calculate_covariates(
  covariate = "huc",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  geom = "sf"
)

print(point_values)
```
