---
title: "Koppen-Geiger Climate Classes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Koppen-Geiger Climate Classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

```

This article demonstrates a compact workflow for Koppen-Geiger climate classification rasters.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

- `download_data(dataset_name = "koppen", ...)` accepts `data_resolution = "0.0083"`, `"0.083"`, or `"0.5"`.
- `time_period` is either `"Present"` for the 1980-2016 classification or `"Future"` for the 2071-2100 classification.
- Downloads arrive as zipped climate-classification raster files; the wrapper exposes period-level products rather than individual years or scenarios.
- Köppen-Geiger outputs are categorical climate classes, so treat the raster values as class codes instead of a continuous climate surface.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "koppen_workflow")
download_data(
  dataset_name = "koppen",
  data_resolution = "0.5",
  time_period = "Present",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)

```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "koppen",
  path = list.files(
    directory_to_save,
    pattern = "\\.tif$",
    recursive = TRUE,
    full.names = TRUE
  )[1],
  extent = terra::ext(-114.9, -102.0, 31.3, 41.1)
)

plot(processed_data, main = "Koppen-Geiger Climate Classes (0.5° resolution, 1980-2016)")
```

## Calculate covariates at points. Demonstrate the dummy variable and fraction options for categorical covariates.

```{r calculate-points, eval = live_run}

domain_x <- c(terra::xmin(processed_data), terra::xmax(processed_data))
domain_y <- c(terra::ymin(processed_data), terra::ymax(processed_data))
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
example_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))


point_values <- calculate_covariates(
  covariate = "koppen",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  geom = "sf"
)

print(point_values, n = 25)

frac_values <- calculate_covariates(
  covariate = "koppen",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  frac = TRUE,
  radius = 100000,
  geom = "sf"
)

print(frac_values, n = 25)

```



