---
title: "NASA MERRA-2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA MERRA-2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  preferred <- candidates[vapply(x[candidates], function(col) {
    is.numeric(col) || is.character(col) || is.factor(col)
  }, logical(1))]
  if (length(preferred) > 0) preferred[1] else candidates[1]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::theme_minimal()
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::theme_minimal()
}
```

This article demonstrates a compact workflow for NASA MERRA-2 data. MERRA-2 downloads require a NASA EarthData token.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "merra2", ...)` wraps `download_merra2()`.

- `collection` accepts a large set of MERRA-2 collection IDs across instantaneous (`inst*`), time-averaged (`tavg*`), and daily-statistic (`statD*`) families for 2-D and 3-D meteorology and chemistry fields; examples include `inst1_2d_asm_Nx`, `tavg1_2d_slv_Nx`, and `tavg3_3d_mst_Np`. It also accepts `"fwi"` for the daily corrected Global Fire Weather Index product (`MERRA2.CORRECTED`).
- `date` can be a single day or a start/end range, and requests return daily `.nc4` files for each selected collection.
- For `collection = "fwi"`, downloads return daily `.nc` files from the public GlobalFWI portal rather than GES DISC.
- Downloads are global only; clip to a study area during processing rather than at download time.
- NASA EarthData authentication is required for standard GES DISC collections, but not for `collection = "fwi"`.
- `amadeus` validates collection IDs against its built-in MERRA-2 table, routes requests to the correct GES DISC server automatically, and also saves companion `.xml` metadata files under each collection's `metadata/` folder.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "merra2_workflow")
download_data(
  dataset_name = "merra2",
  collection = c("inst1_2d_asm_Nx", "tavg1_2d_slv_Nx"),
  date = c("2022-02-14", "2022-02-15"),
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "merra2",
  variable = "CPT",
  date = c("2022-02-14", "2022-02-15"),
  path = dirname(list.files(
    directory_to_save,
    pattern = "\\.nc4?$",
    recursive = TRUE,
    full.names = TRUE
  )[1]),
  extent = terra::ext(-79.2, -78.6, 35.8, 36.3)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "merra2",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "merra2",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("NASA MERRA-2", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("NASA MERRA-2", ": Durham H3 polygons"))
```

## Notes

- The download request now spans two MERRA-2 collections so live runs pull more than one product family.
- Short date windows still exercise the processing and extraction pipeline without creating an unmanageable local cache.
