---
title: "NASA MERRA-2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA MERRA-2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  out.width = "100%"
)
library(amadeus)
library(sf)
library(ggplot2)
live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

find_merra2_collection_dir <- function(root, collection) {
  pattern <- if (collection == "fwi") {
    "^FWI\\..*\\.nc$"
  } else {
    paste0("^MERRA2_[0-9]{3}\\.", collection, "\\..*\\.nc4$")
  }
  files <- list.files(
    root,
    pattern = pattern,
    recursive = TRUE,
    full.names = TRUE
  )
  if (length(files) == 0) {
    stop("Could not locate files for collection ", collection, ".")
  }
  dirname(files[1])
}

has_merra2_collection_files <- function(root, collection) {
  pattern <- if (collection == "fwi") {
    "^FWI\\..*\\.nc$"
  } else {
    paste0("^MERRA2_[0-9]{3}\\.", collection, "\\..*\\.nc4$")
  }
  length(list.files(
    root,
    pattern = pattern,
    recursive = TRUE,
    full.names = TRUE
  )) > 0
}

summarize_na_counts <- function(x, value_pattern) {
  data <- if ("sf" %in% class(x)) {
    sf::st_drop_geometry(x)
  } else {
    x
  }
  value_cols <- grep(value_pattern, names(data), value = TRUE)
  if (length(value_cols) == 0) {
    return(data.frame(
      variable = character(0),
      missing_values = integer(0),
      stringsAsFactors = FALSE
    ))
  }
  data.frame(
    variable = value_cols,
    missing_values = vapply(data[value_cols], function(col) {
      sum(is.na(col))
    }, integer(1)),
    stringsAsFactors = FALSE
  )
}

merra2_demo_variables <- data.frame(
  collection = c(
    "`inst1_2d_int_Nx`",
    "`inst3_2d_gas_Nx`",
    "`statD_2d_slv_Nx`",
    "`tavg1_2d_adg_Nx`",
    "`fwi`",
    "`fwi`"
  ),
  variable = c(
    "`CPT`",
    "`AODANA`",
    "`HOURNORAIN`",
    "`BCEMAN`",
    "`FFMC`",
    "`FWI`"
  ),
  workflow_role = c(
    "Instantaneous meteorology / diagnostics example",
    "3-hourly gas or aerosol example",
    "Daily surface statistic example",
    "Hourly aerosol diagnostics example aggregated to a daily summary",
    "Daily fire weather layer used in extraction demos",
    "Daily fire weather layer used in extraction demos"
  ),
  stringsAsFactors = FALSE
)

merra2_collection_reference <- data.frame(
  collection_family = c(
    "`inst1_2d_asm_Nx`, `inst1_2d_int_Nx`, `inst1_2d_lfo_Nx`",
    "`inst3_2d_gas_Nx`",
    paste(
      "`inst3_3d_asm_Np`, `inst3_3d_aer_Nv`, `inst3_3d_asm_Nv`,",
      "`inst3_3d_chm_Nv`, `inst3_3d_gas_Nv`, `inst6_3d_ana_Np`,",
      "`inst6_3d_ana_Nv`"
    ),
    paste(
      "`tavg1_2d_adg_Nx`, `tavg1_2d_aer_Nx`, `tavg1_2d_chm_Nx`,",
      "`tavg1_2d_csp_Nx`, `tavg1_2d_flx_Nx`, `tavg1_2d_int_Nx`,",
      "`tavg1_2d_lfo_Nx`, `tavg1_2d_lnd_Nx`, `tavg1_2d_ocn_Nx`,",
      "`tavg1_2d_rad_Nx`, `tavg1_2d_slv_Nx`"
    ),
    paste(
      "`tavg3_3d_mst_Ne`, `tavg3_3d_trb_Ne`, `tavg3_3d_nav_Ne`,",
      "`tavg3_3d_cld_Np`, `tavg3_3d_mst_Np`, `tavg3_3d_rad_Np`,",
      "`tavg3_3d_tdt_Np`, `tavg3_3d_trb_Np`, `tavg3_3d_udt_Np`,",
      "`tavg3_3d_odt_Np`, `tavg3_3d_qdt_Np`, `tavg3_3d_asm_Nv`,",
      "`tavg3_3d_cld_Nv`, `tavg3_3d_mst_Nv`, `tavg3_3d_rad_Nv`,",
      "`tavg3_2d_glc_Nx`"
    ),
    "`statD_2d_slv_Nx`",
    "`fwi`"
  ),
  cadence = c(
    "Hourly instantaneous",
    "3-hourly instantaneous",
    "3-hourly or 6-hourly instantaneous",
    "Hourly time averaged",
    "3-hourly time averaged",
    "Daily statistics",
    "Daily corrected fire weather"
  ),
  files = c(
    "Daily `.nc4`",
    "Daily `.nc4`",
    "Daily `.nc4`",
    "Daily `.nc4`",
    "Daily `.nc4`",
    "Daily `.nc4`",
    "Daily `.nc`"
  ),
  auth = c(
    "NASA EarthData token required",
    "NASA EarthData token required",
    "NASA EarthData token required",
    "NASA EarthData token required",
    "NASA EarthData token required",
    "NASA EarthData token required",
    "Public GlobalFWI portal; no token required"
  ),
  example_variables = c(
    "`CPT`",
    "`AODANA`",
    "`AIRDENS`, `SLP`",
    "`BCEMAN`, `COCL`",
    "`DUDTANA`",
    "`HOURNORAIN`",
    "`DC`, `DMC`, `FFMC`, `ISI`, `BUI`, `FWI`"
  ),
  stringsAsFactors = FALSE
)

merra2_variable_reference <- data.frame(
  collection = c(
    "`inst1_2d_int_Nx`",
    "`inst3_2d_gas_Nx`",
    "`inst3_3d_chm_Nv`",
    "`inst6_3d_ana_Np`",
    "`statD_2d_slv_Nx`",
    "`tavg1_2d_chm_Nx`",
    "`tavg3_3d_udt_Np`",
    "`fwi`"
  ),
  example_variable = c(
    "`CPT`",
    "`AODANA`",
    "`AIRDENS`",
    "`SLP`",
    "`HOURNORAIN`",
    "`COCL`",
    "`DUDTANA`",
    "`FWI` or raw layer `MERRA2.CORRECTED_FWI`"
  ),
  output_time_structure = c(
    "Hourly layers with date and hour in layer names",
    "3-hourly layers with date and hour in layer names",
    "3-hourly layers; 3-D collection keeps pressure-level metadata",
    "6-hourly layers; 3-D collection keeps pressure-level metadata",
    "Daily layers with collection-specific timestamps",
    "Hourly layers with date and hour in layer names",
    "3-hourly layers; 3-D collection keeps pressure-level metadata",
    "Daily layers named as `MERRA2.CORRECTED.<var>_<YYYYMMDD>`"
  ),
  stringsAsFactors = FALSE
)
```

This article demonstrates a compact, multi-variable workflow for NASA MERRA-2 data. Standard MERRA-2 GES DISC downloads require a NASA EarthData token, while the public FWI collection does not.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

`download_data(dataset_name = "merra2", ...)` wraps `download_merra2()`.

`process_covariates(covariate = "merra2", variable = ...)` accepts native layer
names from the selected MERRA-2 collection. For the new public FWI product,
processable variables are `DC`, `DMC`, `FFMC`, `ISI`, `BUI`, and `FWI` (or the
raw layer name, such as `MERRA2.CORRECTED_FWI`).

```{r merra2-collection-reference, echo = FALSE, results = "asis"}
knitr::kable(
  merra2_collection_reference,
  col.names = c(
    "Supported collection family",
    "Temporal cadence",
    "File type returned by download",
    "Authentication / source",
    "Representative variables you can process"
  ),
  escape = FALSE
)
```

```{r merra2-variable-reference, echo = FALSE, results = "asis"}
knitr::kable(
  merra2_variable_reference,
  col.names = c(
    "Collection",
    "Example `variable =` value",
    "What the processed output looks like"
  ),
  escape = FALSE
)
```

- `date` can be a single day or a start/end range.
- Downloads are global only; clip to a study area during processing rather than at download time.
- Standard GES DISC collections save companion `.xml` metadata files under each collection's `metadata/` folder.
- If you need to discover additional variables inside a downloaded file, inspect the native layer names with `names(terra::rast(path_to_file))` before calling `process_covariates()`.

## Workflow demonstration variables

The live example below processes six variables spanning standard MERRA-2 and
the new public FWI product. The point and polygon extraction chunks focus on
the two FWI layers so the extracted values stay dense over the Durham example
locations and hexagons, and a separate hourly `BCEMAN` example shows how to
roll a 1-hour product up to a daily summary for extraction.

```{r merra2-demo-reference, echo = FALSE, results = "asis"}
knitr::kable(
  merra2_demo_variables,
  col.names = c(
    "Collection",
    "Variable shown in the workflow",
    "Why it is included"
  ),
  escape = FALSE
)
```

## Download representative requests

You can use the helper function `get_merra2_info()` to query available variables from each collection.

```{r download, eval = live_run}

merra2_demo_specs <- data.frame(
  collection = c(
    "inst1_2d_int_Nx",
    "inst3_2d_gas_Nx",
    "statD_2d_slv_Nx",
    "tavg1_2d_adg_Nx",
    "fwi",
    "fwi"
  ),
  variable = c("CPT", "AODANA", "HOURNORAIN", "BCEMAN", "FFMC", "FWI"),
  date = rep("2024-08-11", 6),
  use_for_extraction = c(FALSE, FALSE, FALSE, FALSE, TRUE, TRUE),
  stringsAsFactors = FALSE
)

directory_to_save <- file.path(tempdir(), "merra2_workflow")
download_data(
  dataset_name = "merra2",
  collection = unique(merra2_demo_specs$collection),
  date = "2024-08-11",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)

get_merra2_info(path = paste0(directory_to_save,"/inst1_2d_int_Nx"))
```

This single request intentionally mixes authenticated GES DISC collections with
the public `fwi` collection so the vignette demonstrates both download paths in
the same workflow.
## Process six workflow-ready data products

```{r process, eval = live_run}
processed_examples <- setNames(
  vector("list", nrow(merra2_demo_specs)),
  merra2_demo_specs$variable
)

find_merra2_collection_dir <- function(root, collection) {
  pattern <- if (collection == "fwi") {
    "^FWI\\..*\\.nc$"
  } else {
    paste0("^MERRA2_[0-9]{3}\\.", collection, "\\..*\\.nc4$")
  }
  files <- list.files(
    root,
    pattern = pattern,
    recursive = TRUE,
    full.names = TRUE
  )
  if (length(files) == 0) {
    stop("Could not locate files for collection ", collection, ".")
  }
  dirname(files[1])
}

for (i in seq_len(nrow(merra2_demo_specs))) {
  spec <- merra2_demo_specs[i, ]
  processed_examples[[spec$variable]] <- process_covariates(
    covariate = "merra2",
    variable = spec$variable,
    date = spec$date,
    path = find_merra2_collection_dir(directory_to_save, spec$collection)
  )
}
processed_summary <- data.frame(
  variable = merra2_demo_specs$variable,
  collection = merra2_demo_specs$collection,
  n_layers = vapply(processed_examples, function(x) as.integer(terra::nlyr(x)), integer(1)),
  first_output_layer = vapply(processed_examples, function(x) {
    names(x)[1]
  }, character(1)),
  stringsAsFactors = FALSE
)
knitr::kable(
  processed_summary,
  col.names = c(
    "Variable",
    "Collection",
    "Layers returned",
    "First layer name in the processed raster"
  )
)
```

## Plot one processed raster

```{r plot-processed-raster, eval = live_run, fig.width = 12, fig.height = 8, fig.alt = "Global map of the processed daily Fire Weather Index raster for the workflow date."}
terra::plot(
  processed_examples$FWI,
  main = "MERRA-2 corrected FWI on 2024-08-11",
  plg = list(x = "bottom", title = "FWI")
)
```

## Demonstrate hourly MERRA-2 data with `tavg1_2d_adg_Nx`

Many MERRA-2 datasets are at 1, 3, or 6 hour increments. Here, we inspect the hourly timestamps.

```{r bceman-hourly-summary, eval = live_run}
bceman_hourly_layers <- data.frame(
  layer_name = names(processed_examples$BCEMAN),
  time_utc = as.character(terra::time(processed_examples$BCEMAN)),
  stringsAsFactors = FALSE
)
knitr::kable(
  head(bceman_hourly_layers, 8),
  col.names = c("Hourly layer name", "Timestamp (UTC)")
)
```



## View the daily BCEMAN summary

Process and calculate functions currently return results in the native time resolution of the data. Here, again, that is hourly. To calculate a daily summary we simply use an apply style function from `terra`. 

```{r bceman-daily-summary, eval = live_run}
bceman_daily <- terra::app(processed_examples$BCEMAN, mean, na.rm = TRUE)
names(bceman_daily) <- "BCEMAN_20240811_0000"
terra::time(bceman_daily) <- as.POSIXct("2024-08-11 00:00:00", tz = "UTC")
```
```{r plot-bceman-daily, eval = live_run, fig.width = 12, fig.height = 8, fig.alt = "Global map of the daily mean BCEMAN raster summarized from hourly MERRA-2 data."}
terra::plot(
  bceman_daily,
  main = "Daily mean BCEMAN from tavg1_2d_adg_Nx on 2024-08-11",
  plg = list(x = "bottom", title = "BCEMAN")
)
```

## Calculate daily BCEMAN at points 

```{r calculate-bceman-daily, eval = live_run}


df <- data.frame(
  site_id = c("site_1", "site_2", "site_3","site_4"),
  lon = c(-78.6382, -47.8825, 116.4074, 36.8219),
  lat = c(35.7796, -15.7942, 39.9042, -1.2921)
)
example_points_sf <- sf::st_as_sf(
  df,
  coords = c("lon", "lat"),
  crs = 4326
)

bceman_point_values <- calculate_covariates(
  covariate = "merra2",
  from = bceman_daily,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)

print(bceman_point_values)
```

## Calculate covariates at points with dense-coverage FWI layers

```{r calculate-points, eval = live_run}
point_ffmc <- calculate_covariates(
  covariate = "merra2",
  from = processed_examples$FFMC,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
point_fwi <- calculate_covariates(
  covariate = "merra2",
  from = processed_examples$FWI,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
point_values <- dplyr::left_join(
  point_ffmc,
  sf::st_drop_geometry(point_fwi),
  by = c("site_id", "time")
)

print(point_values)

```


## Visualize the point outputs

```{r plot-points, eval = live_run, fig.width = 12, fig.height = 8, fig.alt = "Map of point-based extraction results for the daily Fire Weather Index layer used in this workflow."}

point_basemap <-
  sf::st_as_sf(maps::map("world", plot = FALSE, fill = TRUE))

ggplot() + geom_sf(data = point_basemap) + geom_sf(data = point_values,aes(color = `MERRA2.CORRECTED.FWI_0`)) + ggtitle("FWI values at example points on 2024-08-11")


```

