---
title: "NASA MODIS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA MODIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  fig.width = 9,
  fig.height = 7,
  out.width = "100%",
  fig.align = "center"
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


modis_workflow_path <- function(product, process_step, calculate_step) {
  download_step <- sprintf(
    '`download_data(dataset_name = "modis", product = "%s", ...)`',
    product
  )
  paste(download_step, process_step, calculate_step, sep = "<br>")
}

workflow_merge <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_modis_merge(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"modis\", preprocess = process_modis_merge, ...)`"
  )
}

workflow_swath <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_modis_swath(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"modis\", preprocess = process_modis_swath, ...)`"
  )
}

workflow_blackmarble <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_blackmarble(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"modis\", preprocess = process_blackmarble, ...)`"
  )
}

workflow_mcd14dl <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_mcd14dl(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"mcd14dl\", ...)`"
  )
}

collapse_subdatasets <- function(x) paste(x, collapse = "<br>")

modis_subdataset_reference <- c(
  MOD09GA = collapse_subdatasets(c(
    "num_observations_500m",
    "sur_refl_b01_1",
    "sur_refl_b02_1",
    "sur_refl_b03_1",
    "sur_refl_b04_1",
    "sur_refl_b05_1",
    "sur_refl_b06_1",
    "sur_refl_b07_1",
    "QC_500m_1",
    "obscov_500m_1",
    "iobs_res_1",
    "q_scan_1"
  )),
  MYD09GA = collapse_subdatasets(c(
    "num_observations_500m",
    "sur_refl_b01_1",
    "sur_refl_b02_1",
    "sur_refl_b03_1",
    "sur_refl_b04_1",
    "sur_refl_b05_1",
    "sur_refl_b06_1",
    "sur_refl_b07_1",
    "QC_500m_1",
    "obscov_500m_1",
    "iobs_res_1",
    "q_scan_1"
  )),
  MOD09GQ = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD09GQ = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD09A1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD09A1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD09Q1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD09Q1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD11A1 = collapse_subdatasets(c(
    "LST_Day_1km",
    "QC_Day",
    "Day_view_time",
    "Day_view_angl",
    "LST_Night_1km",
    "QC_Night",
    "Night_view_time",
    "Night_view_angl",
    "Emis_31",
    "Emis_32",
    "Clear_day_cov",
    "Clear_night_cov"
  )),
  MYD11A1 = collapse_subdatasets(c(
    "LST_Day_1km",
    "QC_Day",
    "Day_view_time",
    "Day_view_angl",
    "LST_Night_1km",
    "QC_Night",
    "Night_view_time",
    "Night_view_angl",
    "Emis_31",
    "Emis_32",
    "Clear_day_cov",
    "Clear_night_cov"
  )),
  MOD11A2 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD11A2 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD11B1 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD11B1 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD13A1 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MYD13A1 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MOD13A2 = collapse_subdatasets(c(
    "\"1 km 16 days NDVI\"",
    "\"1 km 16 days EVI\"",
    "\"1 km 16 days VI Quality\"",
    "\"1 km 16 days red reflectance\"",
    "\"1 km 16 days NIR reflectance\"",
    "\"1 km 16 days blue reflectance\"",
    "\"1 km 16 days MIR reflectance\"",
    "\"1 km 16 days view zenith angle\"",
    "\"1 km 16 days sun zenith angle\"",
    "\"1 km 16 days relative azimuth angle\"",
    "\"1 km 16 days composite day of the year\"",
    "\"1 km 16 days pixel reliability\""
  )),
  MYD13A2 = collapse_subdatasets(c(
    "\"1 km 16 days NDVI\"",
    "\"1 km 16 days EVI\"",
    "\"1 km 16 days VI Quality\"",
    "\"1 km 16 days red reflectance\"",
    "\"1 km 16 days NIR reflectance\"",
    "\"1 km 16 days blue reflectance\"",
    "\"1 km 16 days MIR reflectance\"",
    "\"1 km 16 days view zenith angle\"",
    "\"1 km 16 days sun zenith angle\"",
    "\"1 km 16 days relative azimuth angle\"",
    "\"1 km 16 days composite day of the year\"",
    "\"1 km 16 days pixel reliability\""
  )),
  MOD13A3 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MYD13A3 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MOD13Q1 = "250m 16 days (NDVI|EVI) layers",
  MYD13Q1 = "250m 16 days (NDVI|EVI) layers",
  MCD12Q1 = "(LC_Type) layers (e.g., LC_Type1-5)",
  MOD14A1 = "(FireMask) layers",
  MYD14A1 = "(FireMask) layers",
  MOD14A2 = "(FireMask) layers",
  MYD14A2 = "(FireMask) layers",
  MOD14CM1 = "(FireMask) layers",
  MYD14CM1 = "(FireMask) layers",
  MCD64A1 = "(Burn Date|BurnDate) layers",
  MCD64CMQ = "(Burn Date|BurnDate) layers",
  VNP64A1 = "(BurnDate) layers",
  MOD16A2 = "(ET_500m|PET_500m) layers",
  MYD16A2 = "(ET_500m|PET_500m) layers",
  MOD06_L2 = collapse_subdatasets(c(
    "Cloud_Phase_Infrared_1km",
    "cloud_top_pressure_1km",
    "cloud_top_height_1km",
    "cloud_top_temperature_1km",
    "Cloud_Effective_Radius",
    "Cloud_Optical_Thickness",
    "Cloud_Water_Path",
    "Cloud_Phase_Optical_Properties",
    "Cloud_Multi_Layer_Flag",
    "Cirrus_Reflectance"
  )),
  MCD14DL = "N/A (.txt active fire detections)",
  MCD19A2 = collapse_subdatasets(c(
    "Optical_Depth_047_1",
    "Optical_Depth_047_2",
    "Optical_Depth_047_3",
    "Optical_Depth_055_1",
    "Optical_Depth_055_2",
    "Optical_Depth_055_3",
    "AOD_Uncertainty_1",
    "AOD_Uncertainty_2",
    "AOD_Uncertainty_3",
    "Column_WV_1",
    "Column_WV_2",
    "Column_WV_3",
    "AngstromExp_470-780_1",
    "AngstromExp_470-780_2",
    "AngstromExp_470-780_3",
    "AOD_QA_1",
    "AOD_QA_2",
    "AOD_QA_3",
    "FineModeFraction_1",
    "FineModeFraction_2",
    "FineModeFraction_3",
    "Injection_Height_1",
    "Injection_Height_2",
    "Injection_Height_3"
  )),
  VNP46A2 = collapse_subdatasets(c(
    "DNB_BRDF-Corrected_NTL",
    "DNB_Lunar_Irradiance",
    "Gap_Filled_DNB_BRDF-Corrected_NTL",
    "Latest_High_Quality_Retrieval",
    "Mandatory_Quality_Flag",
    "QF_Cloud_Mask",
    "Snow_Flag"
  ))
)

modis_reference <- data.frame(
  family = c(
    "Surface reflectance", "Surface reflectance",
    "Surface reflectance", "Surface reflectance",
    "Surface reflectance", "Surface reflectance",
    "Surface reflectance", "Surface reflectance",
    "Land surface temperature", "Land surface temperature",
    "Land surface temperature", "Land surface temperature",
    "Land surface temperature", "Land surface temperature",
    "Vegetation indices", "Vegetation indices",
    "Vegetation indices", "Vegetation indices",
    "Vegetation indices", "Vegetation indices",
    "Vegetation indices", "Vegetation indices",
    "Land cover",
    "Fire grids", "Fire grids", "Fire grids", "Fire grids",
    "Fire grids", "Fire grids",
    "Burned area", "Burned area", "Burned area",
    "Evapotranspiration", "Evapotranspiration",
    "Cloud swaths",
    "Active fire detections",
    "MAIAC aerosol",
    "Nighttime lights"
  ),
  product = c(
    "MOD09GA", "MYD09GA",
    "MOD09GQ", "MYD09GQ",
    "MOD09A1", "MYD09A1",
    "MOD09Q1", "MYD09Q1",
    "MOD11A1", "MYD11A1",
    "MOD11A2", "MYD11A2",
    "MOD11B1", "MYD11B1",
    "MOD13A1", "MYD13A1",
    "MOD13A2", "MYD13A2",
    "MOD13A3", "MYD13A3",
    "MOD13Q1", "MYD13Q1",
    "MCD12Q1",
    "MOD14A1", "MYD14A1",
    "MOD14A2", "MYD14A2",
    "MOD14CM1", "MYD14CM1",
    "MCD64A1", "MCD64CMQ", "VNP64A1",
    "MOD16A2", "MYD16A2",
    "MOD06_L2",
    "MCD14DL",
    "MCD19A2",
    "VNP46A2"
  ),
  level = c(
    "L2G", "L2G", "L2G", "L2G",
    "L3", "L3", "L3", "L3",
    "L3", "L3", "L3", "L3", "L3", "L3",
    "L3", "L3", "L3", "L3", "L3", "L3", "L3", "L3",
    "L3",
    "L3", "L3", "L3", "L3", "L3", "L3",
    "L3", "L3", "L3 (VIIRS)",
    "L4", "L4",
    "L2",
    "L2 (derived NRT)",
    "L2G",
    "L3 (VIIRS)"
  ),
  platform = c(
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra + Aqua",
    "Terra", "Aqua", "Terra", "Aqua", "Terra", "Aqua",
    "Terra + Aqua", "Terra + Aqua", "Suomi-NPP (VIIRS)",
    "Terra", "Aqua",
    "Terra",
    "Terra + Aqua",
    "Terra + Aqua",
    "Suomi-NPP (VIIRS)"
  ),
  description = c(
    "Terra daily surface reflectance grid",
    "Aqua daily surface reflectance grid",
    "Terra daily 250 m surface reflectance grid",
    "Aqua daily 250 m surface reflectance grid",
    "Terra 8-day surface reflectance composite",
    "Aqua 8-day surface reflectance composite",
    "Terra 8-day 250 m surface reflectance composite",
    "Aqua 8-day 250 m surface reflectance composite",
    "Terra daily land surface temperature and emissivity",
    "Aqua daily land surface temperature and emissivity",
    "Terra 8-day land surface temperature and emissivity",
    "Aqua 8-day land surface temperature and emissivity",
    "Terra daily land surface temperature and emissivity",
    "Aqua daily land surface temperature and emissivity",
    "Terra 16-day vegetation indices",
    "Aqua 16-day vegetation indices",
    "Terra 16-day vegetation indices",
    "Aqua 16-day vegetation indices",
    "Terra monthly vegetation indices",
    "Aqua monthly vegetation indices",
    "Terra 16-day vegetation indices at 250 m",
    "Aqua 16-day vegetation indices at 250 m",
    "Combined MODIS annual global land cover",
    "Terra daily thermal anomalies / fire mask grid",
    "Aqua daily thermal anomalies / fire mask grid",
    "Terra 8-day thermal anomalies / fire mask grid",
    "Aqua 8-day thermal anomalies / fire mask grid",
    "Terra monthly fire climate modeling grid",
    "Aqua monthly fire climate modeling grid",
    "Combined MODIS monthly burned area",
    "Combined MODIS monthly climate-model burned area",
    "VIIRS monthly burned area",
    "Terra 8-day evapotranspiration and potential ET",
    "Aqua 8-day evapotranspiration and potential ET",
    "Terra cloud swath granules",
    "Combined MODIS active fire detections text feed",
    "Combined MODIS MAIAC aerosol optical depth product",
    "VIIRS Black Marble nighttime lights"
  ),
  cadence = c(
    rep("Daily", 4),
    rep("Once every 8 days", 4),
    rep("Daily", 2),
    rep("Once every 8 days", 2),
    rep("Daily", 2),
    rep("Once every 16 days", 6),
    rep("Monthly", 2),
    "Yearly",
    rep("Daily", 2),
    rep("Once every 8 days", 2),
    rep("Monthly", 2),
    rep("Monthly", 3),
    rep("Once every 8 days", 2),
    "Daily swath granules",
    "Daily text feed",
    "Daily",
    "Daily"
  ),
  spatial_form = c(
    rep("Tiled gridded", 34),
    "Swath granules",
    "Text feed detections",
    "Tiled gridded",
    "Tiled gridded"
  ),
  native_format = c(
    rep(".hdf", 32),
    ".h5",
    ".hdf",
    ".hdf",
    ".txt",
    ".hdf",
    ".h5"
  ),
  version_handling = c(
    rep("Uses user/default version (default 061)", 27),
    "Forces version 005",
    "Forces version 005",
    "Uses user/default version (default 061)",
    "Forces version 006",
    "No version string sent to CMR",
    rep("Uses user/default version (default 061)", 2),
    "Forces version 6.1",
    "Forces version 6.1NRT",
    "Uses user/default version (default 061)",
    "No version string sent to CMR"
  ),
  cross_year_range = c(
    rep("No", 27),
    "Yes",
    "Yes",
    "Yes",
    "Yes",
    "Yes",
    "No",
    "No",
    "Yes",
    "Yes",
    "No",
    "No"
  ),
  workflow = c(
    workflow_merge("MOD09GA"),
    workflow_merge("MYD09GA"),
    workflow_merge("MOD09GQ"),
    workflow_merge("MYD09GQ"),
    workflow_merge("MOD09A1"),
    workflow_merge("MYD09A1"),
    workflow_merge("MOD09Q1"),
    workflow_merge("MYD09Q1"),
    workflow_merge("MOD11A1"),
    workflow_merge("MYD11A1"),
    workflow_merge("MOD11A2"),
    workflow_merge("MYD11A2"),
    workflow_merge("MOD11B1"),
    workflow_merge("MYD11B1"),
    workflow_merge("MOD13A1"),
    workflow_merge("MYD13A1"),
    workflow_merge("MOD13A2"),
    workflow_merge("MYD13A2"),
    workflow_merge("MOD13A3"),
    workflow_merge("MYD13A3"),
    workflow_merge("MOD13Q1"),
    workflow_merge("MYD13Q1"),
    workflow_merge("MCD12Q1"),
    workflow_merge("MOD14A1"),
    workflow_merge("MYD14A1"),
    workflow_merge("MOD14A2"),
    workflow_merge("MYD14A2"),
    workflow_merge("MOD14CM1"),
    workflow_merge("MYD14CM1"),
    workflow_merge("MCD64A1"),
    workflow_merge("MCD64CMQ"),
    workflow_merge("VNP64A1"),
    workflow_merge("MOD16A2"),
    workflow_merge("MYD16A2"),
    workflow_swath("MOD06_L2"),
    workflow_mcd14dl("MCD14DL"),
    workflow_merge("MCD19A2"),
    workflow_blackmarble("VNP46A2")
  ),
  stringsAsFactors = FALSE
)

modis_reference$subdatasets <- unname(modis_subdataset_reference[modis_reference$product])
modis_reference$subdatasets[is.na(modis_reference$subdatasets)] <-
  "Inspect with terra::describe(path, sds = TRUE)"
```

This article demonstrates how MODIS, combined MODIS, and VIIRS-style products move through `amadeus` via `download_data()`, `process_modis_merge()`, `process_modis_swath()`, `process_blackmarble()`, `process_mcd14dl()`, and `calculate_covariates()`. MODIS downloads require a NASA EarthData token.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces created within the MODIS example extent: two `sf` points in the upper half of the bounding box for point extraction, and a subset of packaged Durham County Uber H3 resolution-8 hexagons filtered to the same upper-half extent for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "modis", ...)` wraps `download_modis()`.

```{r modis-product-reference, echo = FALSE, results = "asis"}
knitr::kable(
  modis_reference[, c(
    "family",
    "product",
    "level",
    "platform",
    "description",
    "cadence",
    "spatial_form",
    "native_format",
    "subdatasets",
    "version_handling",
    "cross_year_range",
    "workflow"
  )],
  col.names = c(
    "Family",
    "Product",
    "NASA level",
    "Platform",
    "Description",
    "Temporal cadence",
    "Spatial form",
    "Native file",
    "Subdataset names",
    "Version handling",
    "Cross-year date range",
    "Workflow in amadeus"
  ),
  escape = FALSE
)
```

- `version` defaults to `"061"`, but `download_modis()` internally rewrites `MOD06_L2` to `6.1`, `MOD14CM1` / `MYD14CM1` to `005`, `MCD64CMQ` to `006`, `MCD14DL` to `6.1NRT`, and omits the version string for `VNP46A2` and `VNP64A1`.
- `date` can be a single day or a range. Most products must stay within one calendar year; cross-year ranges are supported for `MOD06_L2`, `MOD14CM1`, `MYD14CM1`, `MCD14DL`, `MCD64A1`, `MCD64CMQ`, and `VNP64A1`.
- `extent` limits the NASA CMR query to a study-area bounding box so only intersecting tiles are downloaded.
- Downloads are saved as raw `.hdf` or `.h5` granules, except `MCD14DL`, which is saved as `.txt`. The table now includes a subdataset-name column populated from local granule inspection (`terra::rast(path); names(r)`) where available, plus product-specific selector patterns. Use `terra::describe(path, sds = TRUE)` to inspect the complete subdataset list for your exact files and collection version.



## Live representative workflow: `MOD11A1`

The evaluated chunks below keep one compact, end-to-end `MOD11A1` land-surface temperature example that still exercises the download, process, calculate, and plotting stages. We will demonstate downloading, processing with temporal aggregation, processing with daily resolution, covariate calculating with aggregation and daily resolutions, and plotting. The workflow for other products is similar, but the exact subdataset names and processing parameters may differ.

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "modis_workflow", "MOD11A1")
modis_extent <- c(-79.2, 35.8, -78.6, 36.3)
download_data(
  dataset_name = "modis",
  product = "MOD11A1",
  version = "061",
  extent = modis_extent,
  date = c("2019-08-15","2019-08-18"),
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```
## Process one workflow-ready land-surface temperature product

```{r process, eval = live_run}
modis_files <- list.files(
  directory_to_save,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

# Process with daily resolution, no temporal aggregation
processed_daily <- process_modis_daily(
  path = modis_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km"
)

# Processs with temporal averaging across the 4-day window
processed_avg <- process_modis_merge(
  path = modis_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km",
  fun_agg = "mean"
)
```

```{r plot-processed-rasters, eval = live_run, fig.alt = "Raster plot of the processed MOD11A1 land-surface temperature surface for the example date."}
terra::plot(
  processed_avg,
  main = "MOD11A1 LST_Day_1km averaged over 4 Days (2019-08-15 to 2019-08-18)"
)

terra::plot(
  processed_daily,
  main = "MOD11A1 LST_Day_1km daily resolution (2019-08-15 to 2019-08-18)"
)
```

## Calculate covariates at points - demonstating cacluations at points with buffers

```{r calculate-points, eval = live_run}
# Build two example points in the upper half of the MODIS extent and
# subset H3 hexagons to the same upper-half window.
if (!exists("modis_extent")) {
  modis_extent <- c(-79.2, 35.8, -78.6, 36.3)
}

modis_mid_lat <- mean(modis_extent[c(2, 4)])
example_points_sf <- sf::st_as_sf(
  data.frame(
    site_id = c("northwest_point", "northeast_point"),
    lon = c(-79.05, -78.78),
    lat = c(36.22, 36.18)
  ),
  coords = c("lon", "lat"),
  crs = 4326
)

modis_upper_half <- sf::st_as_sfc(sf::st_bbox(
  c(
    xmin = modis_extent[1],
    ymin = modis_mid_lat,
    xmax = modis_extent[3],
    ymax = modis_extent[4]
  ),
  crs = sf::st_crs(4326)
))

durham_hex <- sf::st_filter(
  sf::st_transform(durham_hex, 4326),
  modis_upper_half,
  .predicate = sf::st_intersects
)

# Demonstate calculation at points with the output from the `process` step with daily resolution. The result in a WIDE format sf object

point_values_process <- calculate_covariates(
  covariate = "modis",
  from = processed_daily,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  name_covariates = names(processed_daily),
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.02 - 273.15"
) |> dplyr::select(-time)

# calculate_modis (or the calculate_covariate version ) can do the process and calculate step all in one. The result is LONG format sf object with a time column. This is the recommended workflow for most users, but the two-step process above allows users to inspect the processed rasters before calculating covariates and to use the same processed rasters for multiple calculate_covariates runs with different parameters.

point_values <- calculate_covariates(
  covariate = "modis",
  from = modis_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "LST_Day_1km",
  name_covariates = "LST_",
  fun_summary = "mean",
  geom = "terra",
  scale = "* 0.02 - 273.15"
)

```

## Calculate covariates across H3 hexagons - demonstating calculations across polygons 

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "modis",
  from = modis_files,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "LST_Day_1km",
  name_covariates = "LST_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.02 - 273.15"
)
```

## Terra+Aqua fusion for better temporal coverage

`process_modis_merge()` `process_modis_daily()` and `calculate_modis()` support optional Terra+Aqua fusion through `path_secondary` / `from_secondary` and `fusion_method`.

For paired products (for example `MOD13Q1` + `MYD13Q1`), this allows:

- pixel-wise mean blending where both products exist (`fusion_method = "mean"`)
- fallback to whichever product is available on a date
- optional priority fallback (`"primary_first"` / `"secondary_first"`)

```{r process complementary TERRA/AQUA , eval = live_run}
directory_aqua <- file.path(tempdir(), "modis_workflow", "MYD11A1")

download_data(
  dataset_name = "modis",
  product = "MYD11A1",
  version = "061",
  extent = modis_extent,
  date = c("2019-08-15","2019-08-18"),
  directory_to_save = directory_aqua,
  acknowledgement = TRUE
)

aqua_files <- list.files(
  directory_aqua,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

processed_comb <- process_modis_daily(
  path = modis_files,
  path_secondary = aqua_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km",
  fun_agg = "mean"
)

# Process and calulate together with fusion in one step with calculate_covariates. The result is LONG format sf object with a time column.
calculated_comb <- calculate_covariates(
  covariate = "modis",
  from = modis_files,
  from_secondary = aqua_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_daily,
  subdataset = "LST_Day_1km",
  name_covariates = "LST_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.02 - 273.15"
)

# Use the process stage output from the complementary fusion as input to calculate_covariates for a more traditional workflow. The result is LONG format sf object with a time column.
calculated_comb_process <- calculate_covariates(
  covariate = "modis",
  from = processed_comb,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  name_covariates = names(processed_comb),
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.02 - 273.15"
) |> dplyr::select(-time)
```


## Visualize the process and calculate results

First, we visualize the process daily rasters with the point caclulations overlayed.

```{r plot-daily, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
r_time <- point_values$time


raster_plot_df <- do.call(rbind, lapply(seq_len(terra::nlyr(processed_daily)), function(i) {
  r_i <- processed_daily[[i]] * 0.02 - 273.15
  r_i <- terra::aggregate(r_i, fact = 6, fun = mean, na.rm = TRUE)
  df_i <- terra::as.data.frame(r_i, xy = TRUE, na.rm = TRUE)
  names(df_i)[3] <- "lst_c"
  df_i$plot_date <- as.Date(r_time[i])
  df_i
}))

point_values_modis <- terra::project(point_values, terra::crs(processed_daily))
point_df <- cbind(
  terra::as.data.frame(point_values_modis),
  terra::geom(point_values_modis)[, c("x", "y")]
)
point_value_col <- grep("^LST_|^lst_", names(point_df), value = TRUE)[1]
point_df$point_value <- point_df[[point_value_col]]
point_df$plot_date <- as.Date(point_df$time)
point_df <- point_df[point_df$plot_date %in% unique(raster_plot_df$plot_date), ]

shared_limits <- range(
  c(raster_plot_df$lst_c, point_df$point_value),
  na.rm = TRUE,
  finite = TRUE
)

ggplot2::ggplot() +
  ggplot2::geom_raster(
    data = raster_plot_df,
    ggplot2::aes(x = x, y = y, fill = lst_c)
  ) +
  ggplot2::geom_point(
    data = point_df,
    ggplot2::aes(x = x, y = y),
    shape = 17,
    size = 4.8,
    color = "black",
    alpha = 0.95
  ) +
  ggplot2::geom_point(
    data = point_df,
    ggplot2::aes(x = x, y = y, color = point_value),
    shape = 17,
    size = 3.6,
    alpha = 0.95
  ) +
  ggplot2::facet_wrap(~plot_date, ncol = 2) +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::scale_fill_viridis_c(option = "C", limits = shared_limits) +
  ggplot2::scale_color_viridis_c(option = "C", limits = shared_limits) +
  ggplot2::labs(
    title = "MOD11A1 daily LST with same-day point covariate overlays",
    fill = "Raster LST (C)",
    color = "Point LST (C)"
  ) +
  ggplot2::theme_minimal(base_size = 13) +
  ggplot2::theme(
    legend.position = "bottom",
    legend.key.width = grid::unit(2, "cm"),
    legend.title = ggplot2::element_text(face = "bold")
  )

```

## Workflow template for using .by_time to create space-time summaries 

Here we demonstrate with EVI a monthly temporal aggregation by polygon. For MODIS datasets, it is faster and recommended to do the calculate step directly and provide the process via the `preprocess` input. A demonstration of the recommended approach and the alternative (which requires using terra::tapp) are below.

#### Vegetation indices: `MOD13*` or `MYD13*`

```{r vegetation-template, eval = FALSE}
vegetation_dir <- file.path(tempdir(), "modis_workflow", "vegetation")
vegetation_window <- c("2019-01-01", "2019-04-30")
download_modis(
  product = "MOD13A2",
  version = "061",
  date = vegetation_window,
  extent = modis_extent,
  directory_to_save = vegetation_dir,
  acknowledgement = TRUE
)

vegetation_files <- list.files(
  vegetation_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

For MODIS data, it is faster to do the process and calculate in one step with calculate_covariates and provide the process as the preprocess function.

evi_points_monthly <- calculate_covariates(
  covariate = "modis",
  from = vegetation_files, # provide the path to the files from download_modis
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "(EVI)",
  name_covariates = "evi_",
  .by_time = "month", # specify the temporal aggregation period for the output points
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.0001"
)

print(evi_points_monthly)

# If we want to view the processed rasters we can use process_modis_merge or process_modis_daily. 
# process_modis_merge without .by_time will give us one raster with the temporal aggregation across the whole window. process_modis_daily will give us daily rasters that we can then aggregate with terra::tapp or similar functions.
vegetation_raster <- process_modis_merge(
  path = vegetation_files,
  date = vegetation_window,
  subdataset = "(EVI)",
  fun_agg = "mean"
)

vegetation_daily_raster <- process_modis_daily(
  path = vegetation_files,
  date = vegetation_window,
  subdataset = "(EVI)"
)

plot(vegetation_raster, main = "MOD13A2 EVI averaged over Jan-Apr 2019")
plot(vegetation_daily_raster)

```

This same averaging pattern applies to `MOD13A1`, `MYD13A1`, and `MYD13A2` when you use a short window that spans two valid 16-day composites. For the monthly products `MOD13A3` and `MYD13A3`, request a date range covering the target month so the monthly granule is included.

### Fire grids: `MOD14A1`, `MYD14A1`, `MOD14CM1`, and `MYD14CM1`

When using fire-grid products with the `FireMask` layer, the raw values are typically interpreted as follows.

```{r fire-mask-reference, echo = FALSE, results = "asis"}
fire_mask_reference <- data.frame(
  raw_value = 0:9,
  meaning = c(
    "not processed, missing input",
    "obsolete, not used since Collection 1",
    "not processed, other reason",
    "non-fire water pixel",
    "cloud, land or water",
    "non-fire land pixel",
    "unknown, land or water",
    "fire, low confidence",
    "fire, nominal confidence",
    "fire, high confidence"
  ),
  binary_fire_mask = c(
    "NA / no observation",
    "NA",
    "NA",
    "0",
    "NA or 0, depending on analysis",
    "0",
    "NA",
    "1, or exclude for stricter mask",
    "1",
    "1"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  fire_mask_reference,
  col.names = c("Raw value", "Meaning", "Binary fire mask?")
)
```

Given the typical interpretation of the `FireMask` layer, users often want to create a binary fire mask where 1 indicates a fire detection and 0 indicates no fire. The exact values to include in the binary fire mask depend on the confidence level you want to include. For example, you might choose to include only high-confidence fires (raw value 9) or to include both nominal and high-confidence fires (raw values 8 and 9). Low-confidence fires (raw value 7) can be included for a more inclusive mask, but they may also include more false positives. To create the binary fire mask from the process or calculate amadeus functions, we simply add a scale argument to recode the raw values to 1s and 0s based on the confidence levels we want to include. For example, if we want to include both nominal and high-confidence fires, we can use `scale = " %in% c(8, 9)"` to recode raw values 8 and 9 to 1 and all other values to 0.

```{r fire-grid-template, eval = FALSE}
fire_grid_dir <- file.path(tempdir(), "modis_workflow", "fire_grid")
fire_grid_window <- c("2019-01-01", "2019-04-30")
download_data(
  dataset_name = "modis",
  product = "MOD14A1",
  version = "061",
  date = fire_grid_window,
  extent = modis_extent,
  directory_to_save = fire_grid_dir,
  acknowledgement = TRUE
)


fire_grid_files <- list.files(
  fire_grid_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)


fire_grid_points <- calculate_covariates(
  covariate = "modis",
  from = fire_grid_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset =  process_modis_sds("MOD14A1"),
  name_covariates = "firemask_",
  fun_summary = "mean",
  geom = "sf",
  scale = " %in% c(8, 9)" # use c(7, 8, 9) to include low-confidence fires
)

# Let's also have a looksy at the processed data 

fire_grid_raster <- process_modis_daily(
  path = fire_grid_files,
  date = fire_grid_window,
  subdataset = process_modis_sds("MOD14A1")
)
fire_grid_raster <- fire_grid_raster %in% c(8, 9) # use c(7, 8, 9) to include low-confidence fires

terra::plot(fire_grid_raster, main = "MOD14A1 processed raster")
```



### MAIAC aerosol: `MCD19A2`

MAIAC MCD19A2 provides information on daily atmospheric properties, of which the most helpful are likely aerosol optical depth and smoke plume height. 

```{r maiac-template, eval = FALSE}
maiac_dir <- file.path(tempdir(), "modis_workflow", "maiac")
MAIAC_grid_window <- c("2019-05-01", "2019-06-30")
maiac_extent <- c(-124.5, 32.5, -114.0, 42.0) # California
download_data(
  dataset_name = "modis",
  product = "MCD19A2",
  version = "061",
  date = MAIAC_grid_window,
  extent = maiac_extent,
  directory_to_save = maiac_dir,
  acknowledgement = TRUE
)

maiac_files <- list.files(
  maiac_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

maiac_raster <- process_modis_daily(
  path = maiac_files,
  date = MAIAC_grid_window,
  subdataset = "Optical_Depth_047",
  scale = "* 0.001"
)

maiac_points <- calculate_covariates(
  covariate = "modis",
  from = maiac_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "Optical_Depth_047",
  name_covariates = "aod_047_",
  fun_summary = "mean",
  geom = "sf", 
  scale = "* 0.001"
)

terra::plot(maiac_raster, main = "MCD19A2 AOD processed raster")

maiac_PH <- process_modis_daily(
  path = maiac_files,
  date = MAIAC_grid_window,
  subdataset = "(Injection_Height)"
)

terra::plot(maiac_PH, main = "MCD19A2 Plume Injection Height processed raster")

```

### Cloud swaths: `MOD06_L2`

```{r cloud-template, eval = FALSE}
cloud_dir <- file.path(tempdir(), "modis_workflow", "cloud")
swath_window <- c("2019-05-01", "2019-05-30")
swath_extent <- c(-124.5, 32.5, -114.0, 42.0) # California
download_data(
  dataset_name = "modis",
  product = "MOD06_L2",
  version = "061",
  date = swath_window,
  extent = swath_extent,
  directory_to_save = cloud_dir,
  acknowledgement = TRUE
)

cloud_files <- list.files(
  cloud_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

cloud_raster <- process_modis_swath(
  path = cloud_files,
  date = "2019-08-15",
  subdataset = "Cloud_Fraction",
  suffix = ":mod06:",
  resolution = 0.05
)

cloud_points <- calculate_covariates(
  covariate = "modis",
  from = cloud_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_swath,
  subdataset = "Cloud_Fraction",
  suffix = ":mod06:",
  resolution = 0.05,
  name_covariates = "cloud_fraction_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(cloud_raster, main = "MOD06_L2 processed raster")
plot_histogram_points(cloud_points, "MOD06_L2 point covariate histogram")
```

### Active fire detections: `MCD14DL`

```{r mcd14dl-template, eval = FALSE}
fire_txt_dir <- file.path(tempdir(), "modis_workflow", "mcd14dl")
download_data(
  dataset_name = "modis",
  product = "MCD14DL",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = fire_txt_dir,
  acknowledgement = TRUE
)

fire_txt_files <- list.files(
  fire_txt_dir,
  pattern = "\\.txt$",
  recursive = TRUE,
  full.names = TRUE
)

fire_detections <- process_mcd14dl(
  path = fire_txt_files,
  date = "2019-08-15",
  extent = modis_extent
)

fire_counts <- calculate_covariates(
  covariate = "mcd14dl",
  from = fire_detections,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = c(0L, 1000L),
  geom = "sf"
)

plot_points(fire_detections, "MCD14DL processed detections")
plot_histogram_points(fire_counts, "MCD14DL point covariate histogram")
```

### Nighttime lights: `VNP46A2`

```{r blackmarble-template, eval = FALSE}
blackmarble_dir <- file.path(tempdir(), "modis_workflow", "blackmarble")
download_data(
  dataset_name = "modis",
  product = "VNP46A2",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = blackmarble_dir,
  acknowledgement = TRUE
)

blackmarble_files <- list.files(
  blackmarble_dir,
  pattern = "\\.h5$",
  recursive = TRUE,
  full.names = TRUE
)

blackmarble_raster <- process_blackmarble(
  path = blackmarble_files,
  date = "2019-08-15",
  tile_df = process_blackmarble_corners(hrange = c(8, 10), vrange = c(4, 5)),
  subdataset = 3L,
  crs = "EPSG:4326"
)

blackmarble_points <- calculate_covariates(
  covariate = "modis",
  from = blackmarble_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_blackmarble,
  tile_df = process_blackmarble_corners(hrange = c(8, 10), vrange = c(4, 5)),
  subdataset = 3L,
  name_covariates = "blackmarble_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(blackmarble_raster, main = "VNP46A2 processed raster")
plot_histogram_points(blackmarble_points, "VNP46A2 point covariate histogram")
```

## Notes

- `calculate_covariates(covariate = "modis", ...)` always receives the raw file list in `from =`; the preprocessing helper named in `preprocess =` is what turns those files into a day-specific raster during extraction.
- `MCD14DL` is the main exception in the MODIS family: it downloads as text, preprocesses with `process_mcd14dl()`, and then extracts with `calculate_covariates(covariate = "mcd14dl", ...)`.
- Because `process_mcd14dl()` returns detections as points rather than a raster, its template plots the processed detections and a histogram of extracted counts instead of a processed raster.
- The example dates are intentionally chosen to align with each representative product cadence: daily products use a single day, while non-daily products are handled with short 15-30 day windows so the plotted rasters and extracted point / H3 summaries are averaged over available composites instead of relying on a single sparse date.
- Apply product-specific scale factors from the NASA product documentation when you move from these templates to analysis-ready covariates. The `MOD11A1` example includes a temperature conversion because that scale is already well established in the package examples.
