---
title: "NASA MODIS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA MODIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

aqs_sample_path <- c(
  file.path("tests", "testdata", "aqs", "aqs-location-sample.rds"),
  file.path("..", "tests", "testdata", "aqs", "aqs-location-sample.rds")
)
aqs_sample_path <- aqs_sample_path[file.exists(aqs_sample_path)][1]
if (is.na(aqs_sample_path) || aqs_sample_path == "") {
  stop("Could not locate tests/testdata/aqs/aqs-location-sample.rds")
}
example_points_sf <- sf::st_as_sf(readRDS(aqs_sample_path))
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)


pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  numeric_candidates <- candidates[vapply(x[candidates], is.numeric, logical(1))]
  ranked <- if (length(numeric_candidates) > 0) numeric_candidates else candidates
  non_missing_share <- vapply(ranked, function(col) {
    mean(!is.na(x[[col]]))
  }, numeric(1))
  ranked[which.max(non_missing_share)]
}

pick_numeric_covariate_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c(
    "site_id", "h3_id", "resolution", "area_km2",
    "name", "NAME", "time", geom_col
  )
  candidates <- setdiff(names(x), excluded)
  numeric_candidates <- candidates[vapply(x[candidates], is.numeric, logical(1))]
  if (length(numeric_candidates) == 0) {
    stop("No numeric covariate columns found for histogram plotting.")
  }
  numeric_candidates[1]
}

coord_local_sf <- function(x) {
  x_wgs84 <- sf::st_transform(sf::st_as_sf(x), 4326)
  bbox <- sf::st_bbox(x_wgs84)
  pad_x <- max((bbox[["xmax"]] - bbox[["xmin"]]) * 0.08, 0.05)
  pad_y <- max((bbox[["ymax"]] - bbox[["ymin"]]) * 0.08, 0.05)
  ggplot2::coord_sf(
    xlim = c(bbox[["xmin"]] - pad_x, bbox[["xmax"]] + pad_x),
    ylim = c(bbox[["ymin"]] - pad_y, bbox[["ymax"]] + pad_y),
    expand = FALSE,
    datum = NA
  )
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(color = !!rlang::sym(value_col)),
      size = 2
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::guides(
      color = ggplot2::guide_colorbar(title.position = "top")
    ) +
    ggplot2::theme_minimal(base_size = 13) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.key.width = grid::unit(2, "cm"),
      legend.title = ggplot2::element_text(face = "bold")
    )
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot() +
    ggplot2::geom_sf(
      data = x,
      ggplot2::aes(fill = !!rlang::sym(value_col)),
      color = NA
    ) +
    coord_local_sf(x) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::guides(
      fill = ggplot2::guide_colorbar(title.position = "top")
    ) +
    ggplot2::theme_minimal(base_size = 13) +
    ggplot2::theme(
      legend.position = "bottom",
      legend.key.width = grid::unit(2, "cm"),
      legend.title = ggplot2::element_text(face = "bold")
    )
}

plot_histogram_points <- function(x, title) {
  value_col <- pick_numeric_covariate_column(x)
  x_hist <- sf::st_drop_geometry(sf::st_as_sf(x))
  ggplot2::ggplot(
    x_hist,
    ggplot2::aes(x = !!rlang::sym(value_col))
  ) +
    ggplot2::geom_histogram(bins = 20, fill = "steelblue", color = "white") +
    ggplot2::labs(title = title, x = value_col, y = "Count") +
    ggplot2::theme_minimal(base_size = 13)
}

summarize_na_counts <- function(x, value_pattern) {
  data <- sf::st_drop_geometry(sf::st_as_sf(x))
  value_cols <- grep(value_pattern, names(data), value = TRUE)
  if (length(value_cols) == 0) {
    stop("No covariate columns matched ", value_pattern, ".")
  }
  data.frame(
    variable = value_cols,
    missing_values = vapply(data[value_cols], function(col) sum(is.na(col)), integer(1)),
    stringsAsFactors = FALSE
  )
}

average_modis_window <- function(paths, dates, preprocess, ...) {
  rasters <- lapply(dates, function(date_i) {
    rlang::exec(preprocess, path = paths, date = date_i, ...)
  })
  if (length(rasters) == 0) {
    stop("No rasters were created for the requested averaging window.")
  }
  if (length(rasters) == 1) {
    return(rasters[[1]])
  }
  raster_stack <- do.call(c, rasters)
  terra::app(raster_stack, mean, na.rm = TRUE)
}

average_covariates_over_time <- function(x, id_col) {
  x_sf <- sf::st_as_sf(x)
  geom_col <- attr(x_sf, "sf_column")
  x_tbl <- sf::st_drop_geometry(x_sf)
  numeric_cols <- setdiff(
    names(x_tbl)[vapply(x_tbl, is.numeric, logical(1))],
    "time"
  )
  if (length(numeric_cols) == 0) {
    stop("No numeric covariate columns found for time averaging.")
  }
  x_avg <- stats::aggregate(
    x_tbl[numeric_cols],
    by = list(x_tbl[[id_col]]),
    FUN = function(value) mean(value, na.rm = TRUE)
  )
  names(x_avg)[1] <- id_col
  x_geom <- x_sf[!duplicated(x_sf[[id_col]]), c(id_col, geom_col)]
  dplyr::left_join(x_geom, x_avg, by = id_col)
}

modis_workflow_path <- function(product, process_step, calculate_step) {
  download_step <- sprintf(
    '`download_data(dataset_name = "modis", product = "%s", ...)`',
    product
  )
  paste(download_step, process_step, calculate_step, sep = "<br>")
}

workflow_merge <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_modis_merge(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"modis\", preprocess = process_modis_merge, ...)`"
  )
}

workflow_swath <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_modis_swath(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"modis\", preprocess = process_modis_swath, ...)`"
  )
}

workflow_blackmarble <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_blackmarble(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"modis\", preprocess = process_blackmarble, ...)`"
  )
}

workflow_mcd14dl <- function(product) {
  modis_workflow_path(
    product = product,
    process_step = "`process_mcd14dl(...)`",
    calculate_step =
      "`calculate_covariates(covariate = \"mcd14dl\", ...)`"
  )
}

collapse_subdatasets <- function(x) paste(x, collapse = "<br>")

modis_subdataset_reference <- c(
  MOD09GA = collapse_subdatasets(c(
    "num_observations_500m",
    "sur_refl_b01_1",
    "sur_refl_b02_1",
    "sur_refl_b03_1",
    "sur_refl_b04_1",
    "sur_refl_b05_1",
    "sur_refl_b06_1",
    "sur_refl_b07_1",
    "QC_500m_1",
    "obscov_500m_1",
    "iobs_res_1",
    "q_scan_1"
  )),
  MYD09GA = collapse_subdatasets(c(
    "num_observations_500m",
    "sur_refl_b01_1",
    "sur_refl_b02_1",
    "sur_refl_b03_1",
    "sur_refl_b04_1",
    "sur_refl_b05_1",
    "sur_refl_b06_1",
    "sur_refl_b07_1",
    "QC_500m_1",
    "obscov_500m_1",
    "iobs_res_1",
    "q_scan_1"
  )),
  MOD09GQ = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD09GQ = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD09A1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD09A1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD09Q1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD09Q1 = "(sur_refl_b0) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD11A1 = collapse_subdatasets(c(
    "LST_Day_1km",
    "QC_Day",
    "Day_view_time",
    "Day_view_angl",
    "LST_Night_1km",
    "QC_Night",
    "Night_view_time",
    "Night_view_angl",
    "Emis_31",
    "Emis_32",
    "Clear_day_cov",
    "Clear_night_cov"
  )),
  MYD11A1 = collapse_subdatasets(c(
    "LST_Day_1km",
    "QC_Day",
    "Day_view_time",
    "Day_view_angl",
    "LST_Night_1km",
    "QC_Night",
    "Night_view_time",
    "Night_view_angl",
    "Emis_31",
    "Emis_32",
    "Clear_day_cov",
    "Clear_night_cov"
  )),
  MOD11A2 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD11A2 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD11B1 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MYD11B1 = "(LST_) pattern; inspect full names with terra::describe(path, sds = TRUE)",
  MOD13A1 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MYD13A1 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MOD13A2 = collapse_subdatasets(c(
    "\"1 km 16 days NDVI\"",
    "\"1 km 16 days EVI\"",
    "\"1 km 16 days VI Quality\"",
    "\"1 km 16 days red reflectance\"",
    "\"1 km 16 days NIR reflectance\"",
    "\"1 km 16 days blue reflectance\"",
    "\"1 km 16 days MIR reflectance\"",
    "\"1 km 16 days view zenith angle\"",
    "\"1 km 16 days sun zenith angle\"",
    "\"1 km 16 days relative azimuth angle\"",
    "\"1 km 16 days composite day of the year\"",
    "\"1 km 16 days pixel reliability\""
  )),
  MYD13A2 = collapse_subdatasets(c(
    "\"1 km 16 days NDVI\"",
    "\"1 km 16 days EVI\"",
    "\"1 km 16 days VI Quality\"",
    "\"1 km 16 days red reflectance\"",
    "\"1 km 16 days NIR reflectance\"",
    "\"1 km 16 days blue reflectance\"",
    "\"1 km 16 days MIR reflectance\"",
    "\"1 km 16 days view zenith angle\"",
    "\"1 km 16 days sun zenith angle\"",
    "\"1 km 16 days relative azimuth angle\"",
    "\"1 km 16 days composite day of the year\"",
    "\"1 km 16 days pixel reliability\""
  )),
  MOD13A3 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MYD13A3 = "(NDVI/EVI) layers; inspect full names with terra::describe(path, sds = TRUE)",
  MOD13Q1 = "250m 16 days (NDVI|EVI) layers",
  MYD13Q1 = "250m 16 days (NDVI|EVI) layers",
  MCD12Q1 = "(LC_Type) layers (e.g., LC_Type1-5)",
  MOD14A1 = "(FireMask) layers",
  MYD14A1 = "(FireMask) layers",
  MOD14A2 = "(FireMask) layers",
  MYD14A2 = "(FireMask) layers",
  MOD14CM1 = "(FireMask) layers",
  MYD14CM1 = "(FireMask) layers",
  MCD64A1 = "(Burn Date|BurnDate) layers",
  MCD64CMQ = "(Burn Date|BurnDate) layers",
  VNP64A1 = "(BurnDate) layers",
  MOD16A2 = "(ET_500m|PET_500m) layers",
  MYD16A2 = "(ET_500m|PET_500m) layers",
  MOD06_L2 = collapse_subdatasets(c(
    "Cloud_Phase_Infrared_1km",
    "cloud_top_pressure_1km",
    "cloud_top_height_1km",
    "cloud_top_temperature_1km",
    "Cloud_Effective_Radius",
    "Cloud_Optical_Thickness",
    "Cloud_Water_Path",
    "Cloud_Phase_Optical_Properties",
    "Cloud_Multi_Layer_Flag",
    "Cirrus_Reflectance"
  )),
  MCD14DL = "N/A (.txt active fire detections)",
  MCD19A2 = collapse_subdatasets(c(
    "Optical_Depth_047_1",
    "Optical_Depth_047_2",
    "Optical_Depth_047_3",
    "Optical_Depth_055_1",
    "Optical_Depth_055_2",
    "Optical_Depth_055_3",
    "AOD_Uncertainty_1",
    "AOD_Uncertainty_2",
    "AOD_Uncertainty_3",
    "Column_WV_1",
    "Column_WV_2",
    "Column_WV_3",
    "AngstromExp_470-780_1",
    "AngstromExp_470-780_2",
    "AngstromExp_470-780_3",
    "AOD_QA_1",
    "AOD_QA_2",
    "AOD_QA_3",
    "FineModeFraction_1",
    "FineModeFraction_2",
    "FineModeFraction_3",
    "Injection_Height_1",
    "Injection_Height_2",
    "Injection_Height_3"
  )),
  VNP46A2 = collapse_subdatasets(c(
    "DNB_BRDF-Corrected_NTL",
    "DNB_Lunar_Irradiance",
    "Gap_Filled_DNB_BRDF-Corrected_NTL",
    "Latest_High_Quality_Retrieval",
    "Mandatory_Quality_Flag",
    "QF_Cloud_Mask",
    "Snow_Flag"
  ))
)

modis_reference <- data.frame(
  family = c(
    "Surface reflectance", "Surface reflectance",
    "Surface reflectance", "Surface reflectance",
    "Surface reflectance", "Surface reflectance",
    "Surface reflectance", "Surface reflectance",
    "Land surface temperature", "Land surface temperature",
    "Land surface temperature", "Land surface temperature",
    "Land surface temperature", "Land surface temperature",
    "Vegetation indices", "Vegetation indices",
    "Vegetation indices", "Vegetation indices",
    "Vegetation indices", "Vegetation indices",
    "Vegetation indices", "Vegetation indices",
    "Land cover",
    "Fire grids", "Fire grids", "Fire grids", "Fire grids",
    "Fire grids", "Fire grids",
    "Burned area", "Burned area", "Burned area",
    "Evapotranspiration", "Evapotranspiration",
    "Cloud swaths",
    "Active fire detections",
    "MAIAC aerosol",
    "Nighttime lights"
  ),
  product = c(
    "MOD09GA", "MYD09GA",
    "MOD09GQ", "MYD09GQ",
    "MOD09A1", "MYD09A1",
    "MOD09Q1", "MYD09Q1",
    "MOD11A1", "MYD11A1",
    "MOD11A2", "MYD11A2",
    "MOD11B1", "MYD11B1",
    "MOD13A1", "MYD13A1",
    "MOD13A2", "MYD13A2",
    "MOD13A3", "MYD13A3",
    "MOD13Q1", "MYD13Q1",
    "MCD12Q1",
    "MOD14A1", "MYD14A1",
    "MOD14A2", "MYD14A2",
    "MOD14CM1", "MYD14CM1",
    "MCD64A1", "MCD64CMQ", "VNP64A1",
    "MOD16A2", "MYD16A2",
    "MOD06_L2",
    "MCD14DL",
    "MCD19A2",
    "VNP46A2"
  ),
  level = c(
    "L2G", "L2G", "L2G", "L2G",
    "L3", "L3", "L3", "L3",
    "L3", "L3", "L3", "L3", "L3", "L3",
    "L3", "L3", "L3", "L3", "L3", "L3", "L3", "L3",
    "L3",
    "L3", "L3", "L3", "L3", "L3", "L3",
    "L3", "L3", "L3 (VIIRS)",
    "L4", "L4",
    "L2",
    "L2 (derived NRT)",
    "L2G",
    "L3 (VIIRS)"
  ),
  platform = c(
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra", "Aqua", "Terra", "Aqua",
    "Terra + Aqua",
    "Terra", "Aqua", "Terra", "Aqua", "Terra", "Aqua",
    "Terra + Aqua", "Terra + Aqua", "Suomi-NPP (VIIRS)",
    "Terra", "Aqua",
    "Terra",
    "Terra + Aqua",
    "Terra + Aqua",
    "Suomi-NPP (VIIRS)"
  ),
  description = c(
    "Terra daily surface reflectance grid",
    "Aqua daily surface reflectance grid",
    "Terra daily 250 m surface reflectance grid",
    "Aqua daily 250 m surface reflectance grid",
    "Terra 8-day surface reflectance composite",
    "Aqua 8-day surface reflectance composite",
    "Terra 8-day 250 m surface reflectance composite",
    "Aqua 8-day 250 m surface reflectance composite",
    "Terra daily land surface temperature and emissivity",
    "Aqua daily land surface temperature and emissivity",
    "Terra 8-day land surface temperature and emissivity",
    "Aqua 8-day land surface temperature and emissivity",
    "Terra daily land surface temperature and emissivity",
    "Aqua daily land surface temperature and emissivity",
    "Terra 16-day vegetation indices",
    "Aqua 16-day vegetation indices",
    "Terra 16-day vegetation indices",
    "Aqua 16-day vegetation indices",
    "Terra monthly vegetation indices",
    "Aqua monthly vegetation indices",
    "Terra 16-day vegetation indices at 250 m",
    "Aqua 16-day vegetation indices at 250 m",
    "Combined MODIS annual global land cover",
    "Terra daily thermal anomalies / fire mask grid",
    "Aqua daily thermal anomalies / fire mask grid",
    "Terra 8-day thermal anomalies / fire mask grid",
    "Aqua 8-day thermal anomalies / fire mask grid",
    "Terra monthly fire climate modeling grid",
    "Aqua monthly fire climate modeling grid",
    "Combined MODIS monthly burned area",
    "Combined MODIS monthly climate-model burned area",
    "VIIRS monthly burned area",
    "Terra 8-day evapotranspiration and potential ET",
    "Aqua 8-day evapotranspiration and potential ET",
    "Terra cloud swath granules",
    "Combined MODIS active fire detections text feed",
    "Combined MODIS MAIAC aerosol optical depth product",
    "VIIRS Black Marble nighttime lights"
  ),
  cadence = c(
    rep("Daily", 4),
    rep("Once every 8 days", 4),
    rep("Daily", 2),
    rep("Once every 8 days", 2),
    rep("Daily", 2),
    rep("Once every 16 days", 6),
    rep("Monthly", 2),
    "Yearly",
    rep("Daily", 2),
    rep("Once every 8 days", 2),
    rep("Monthly", 2),
    rep("Monthly", 3),
    rep("Once every 8 days", 2),
    "Daily swath granules",
    "Daily text feed",
    "Daily",
    "Daily"
  ),
  spatial_form = c(
    rep("Tiled gridded", 34),
    "Swath granules",
    "Text feed detections",
    "Tiled gridded",
    "Tiled gridded"
  ),
  native_format = c(
    rep(".hdf", 32),
    ".h5",
    ".hdf",
    ".hdf",
    ".txt",
    ".hdf",
    ".h5"
  ),
  version_handling = c(
    rep("Uses user/default version (default 061)", 27),
    "Forces version 005",
    "Forces version 005",
    "Uses user/default version (default 061)",
    "Forces version 006",
    "No version string sent to CMR",
    rep("Uses user/default version (default 061)", 2),
    "Forces version 6.1",
    "Forces version 6.1NRT",
    "Uses user/default version (default 061)",
    "No version string sent to CMR"
  ),
  cross_year_range = c(
    rep("No", 27),
    "Yes",
    "Yes",
    "Yes",
    "Yes",
    "Yes",
    "No",
    "No",
    "Yes",
    "Yes",
    "No",
    "No"
  ),
  workflow = c(
    workflow_merge("MOD09GA"),
    workflow_merge("MYD09GA"),
    workflow_merge("MOD09GQ"),
    workflow_merge("MYD09GQ"),
    workflow_merge("MOD09A1"),
    workflow_merge("MYD09A1"),
    workflow_merge("MOD09Q1"),
    workflow_merge("MYD09Q1"),
    workflow_merge("MOD11A1"),
    workflow_merge("MYD11A1"),
    workflow_merge("MOD11A2"),
    workflow_merge("MYD11A2"),
    workflow_merge("MOD11B1"),
    workflow_merge("MYD11B1"),
    workflow_merge("MOD13A1"),
    workflow_merge("MYD13A1"),
    workflow_merge("MOD13A2"),
    workflow_merge("MYD13A2"),
    workflow_merge("MOD13A3"),
    workflow_merge("MYD13A3"),
    workflow_merge("MOD13Q1"),
    workflow_merge("MYD13Q1"),
    workflow_merge("MCD12Q1"),
    workflow_merge("MOD14A1"),
    workflow_merge("MYD14A1"),
    workflow_merge("MOD14A2"),
    workflow_merge("MYD14A2"),
    workflow_merge("MOD14CM1"),
    workflow_merge("MYD14CM1"),
    workflow_merge("MCD64A1"),
    workflow_merge("MCD64CMQ"),
    workflow_merge("VNP64A1"),
    workflow_merge("MOD16A2"),
    workflow_merge("MYD16A2"),
    workflow_swath("MOD06_L2"),
    workflow_mcd14dl("MCD14DL"),
    workflow_merge("MCD19A2"),
    workflow_blackmarble("VNP46A2")
  ),
  stringsAsFactors = FALSE
)

modis_reference$subdatasets <- unname(modis_subdataset_reference[modis_reference$product])
modis_reference$subdatasets[is.na(modis_reference$subdatasets)] <-
  "Inspect with terra::describe(path, sds = TRUE)"
```

This article demonstrates how MODIS, combined MODIS, and VIIRS-style products move through `amadeus` via `download_data()`, `process_modis_merge()`, `process_modis_swath()`, `process_blackmarble()`, `process_mcd14dl()`, and `calculate_covariates()`. MODIS downloads require a NASA EarthData token.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces created within the MODIS example extent: two `sf` points in the upper half of the bounding box for point extraction, and a subset of packaged Durham County Uber H3 resolution-8 hexagons filtered to the same upper-half extent for polygon extraction.

## Available inputs and data availability

`download_data(dataset_name = "modis", ...)` wraps `download_modis()`.

```{r modis-product-reference, echo = FALSE, results = "asis"}
knitr::kable(
  modis_reference,
  col.names = c(
    "Family",
    "Product",
    "NASA level",
    "Platform",
    "Description",
    "Temporal cadence",
    "Spatial form",
    "Native file",
    "Subdataset names",
    "Version handling",
    "Cross-year date range",
    "Workflow in amadeus"
  ),
  escape = FALSE
)
```

- `version` defaults to `"061"`, but `download_modis()` internally rewrites `MOD06_L2` to `6.1`, `MOD14CM1` / `MYD14CM1` to `005`, `MCD64CMQ` to `006`, `MCD14DL` to `6.1NRT`, and omits the version string for `VNP46A2` and `VNP64A1`.
- `date` can be a single day or a range. Most products must stay within one calendar year; cross-year ranges are supported for `MOD06_L2`, `MOD14CM1`, `MYD14CM1`, `MCD14DL`, `MCD64A1`, `MCD64CMQ`, and `VNP64A1`.
- `extent` limits the NASA CMR query to a study-area bounding box so only intersecting tiles are downloaded.
- Downloads are saved as raw `.hdf` or `.h5` granules, except `MCD14DL`, which is saved as `.txt`. The table now includes a subdataset-name column populated from local granule inspection (`terra::rast(path); names(r)`) where available, plus product-specific selector patterns. Use `terra::describe(path, sds = TRUE)` to inspect the complete subdataset list for your exact files and collection version.



## Live representative workflow: `MOD11A1`

The evaluated chunks below keep one compact, end-to-end `MOD11A1` land-surface temperature example that still exercises the download, process, calculate, and plotting stages. We will demonstate downloading, processing with temporal aggregation, processing with daily resolution, covariate calculating with aggregation and daily resolutions, and plotting. The workflow for other products is similar, but the exact subdataset names and processing parameters may differ.

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "modis_workflow", "MOD11A1")
modis_extent <- c(-79.2, 35.8, -78.6, 36.3)
download_data(
  dataset_name = "modis",
  product = "MOD11A1",
  version = "061",
  extent = modis_extent,
  date = c("2019-08-15","2019-08-18"),
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```
## Process one workflow-ready land-surface temperature product

```{r process, eval = live_run}
modis_files <- list.files(
  directory_to_save,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

# Process with daily resolution, no temporal aggregation
processed_daily <- process_modis_daily(
  path = modis_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km"
)

# Processs with temporal averaging across the 4-day window
processed_avg <- process_modis_merge(
  path = modis_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km",
  fun_agg = "mean"
)
```

```{r plot-processed-rasters, eval = live_run, fig.alt = "Raster plot of the processed MOD11A1 land-surface temperature surface for the example date."}
terra::plot(
  processed_avg,
  main = "MOD11A1 LST_Day_1km averaged over 4 Days (2019-08-15 to 2019-08-18)"
)

terra::plot(
  processed_daily,
  main = "MOD11A1 LST_Day_1km daily resolution (2019-08-15 to 2019-08-18)"
)
```

## Calculate covariates at points - demonstating cacluations at points with buffers

```{r calculate-points, eval = live_run}
# Build two example points in the upper half of the MODIS extent and
# subset H3 hexagons to the same upper-half window.
if (!exists("modis_extent")) {
  modis_extent <- c(-79.2, 35.8, -78.6, 36.3)
}

modis_mid_lat <- mean(modis_extent[c(2, 4)])
example_points_sf <- sf::st_as_sf(
  data.frame(
    site_id = c("northwest_point", "northeast_point"),
    lon = c(-79.05, -78.78),
    lat = c(36.22, 36.18)
  ),
  coords = c("lon", "lat"),
  crs = 4326
)

modis_upper_half <- sf::st_as_sfc(sf::st_bbox(
  c(
    xmin = modis_extent[1],
    ymin = modis_mid_lat,
    xmax = modis_extent[3],
    ymax = modis_extent[4]
  ),
  crs = sf::st_crs(4326)
))

durham_hex <- sf::st_filter(
  sf::st_transform(durham_hex, 4326),
  modis_upper_half,
  .predicate = sf::st_intersects
)

point_values <- calculate_covariates(
  covariate = "modis",
  from = modis_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "LST_Day_1km",
  name_covariates = "LST_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.02 - 273.15"
)
```

## Calculate covariates across H3 hexagons - demonstating calculations across polygons 

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "modis",
  from = modis_files,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "LST_Day_1km",
  name_covariates = "LST_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.02 - 273.15"
)
```

```{r process complementary TERRA/AQUA , eval = live_run}
directory_aqua <- file.path(tempdir(), "modis_workflow", "MYD11A1")

download_data(
  dataset_name = "modis",
  product = "MYD11A1",
  version = "061",
  extent = modis_extent,
  date = c("2019-08-15","2019-08-18"),
  directory_to_save = directory_aqua,
  acknowledgement = TRUE
)

aqua_files <- list.files(
  directory_aqua,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

processed_aqua <- process_modis_daily(
  path = aqua_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km"
)

processed_comb <- process_modis_daily(
  path = modis_files,
  path_secondary = aqua_files,
  date = c("2019-08-15","2019-08-18"),
  subdataset = "LST_Day_1km",
  fun_agg = "mean"
)
```




## Terra+Aqua fusion for better temporal coverage

`process_modis_merge()` `process_modis_daily()` and `calculate_modis()` support optional Terra+Aqua fusion through `path_secondary` / `from_secondary` and `fusion_method`.

For paired products (for example `MOD13Q1` + `MYD13Q1`), this allows:

- pixel-wise mean blending where both products exist (`fusion_method = "mean"`)
- fallback to whichever product is available on a date
- optional priority fallback (`"primary_first"` / `"secondary_first"`)



```{r modis-na-summary, eval = live_run}
point_na_summary <- summarize_na_counts(point_values, "^LST_")
polygon_na_summary <- summarize_na_counts(polygon_values, "^LST_")
knitr::kable(point_na_summary, col.names = c("Variable", "Point missing values"))
knitr::kable(polygon_na_summary, col.names = c("Variable", "Polygon missing values"))
```

## Histogram of the point calculations

```{r plot-point-histogram, eval = live_run, fig.alt = "Histogram of point-based MOD11A1 covariate values extracted for the example date."}
plot_histogram_points(
  point_values,
  "MOD11A1 point covariate histogram"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("NASA MODIS", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("NASA MODIS", ": Durham H3 polygons"))
```

## Live non-daily composite example: `MOD13A2` averaged over 17 days

For non-daily products, a short averaging window is more robust than a single composite date. The `MOD13A2` example below downloads two 16-day NDVI composites, averages their processed rasters, then averages the extracted point and H3 values over the same 17-day window.

```{r vegetation-download-window, eval = live_run}
vegetation_dir_live <- file.path(tempdir(), "modis_workflow", "mod13a2")
vegetation_window <- c("2019-08-13", "2019-08-29")
vegetation_dates <- c("2019-08-13", "2019-08-29")
download_data(
  dataset_name = "modis",
  product = "MOD13A2",
  version = "061",
  date = vegetation_window,
  extent = modis_extent,
  directory_to_save = vegetation_dir_live,
  acknowledgement = TRUE
)

vegetation_files_live <- list.files(
  vegetation_dir_live,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)
```

```{r vegetation-process-window, eval = live_run}
vegetation_raster_mean <- average_modis_window(
  paths = vegetation_files_live,
  dates = vegetation_dates,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD13A2")
)
```

```{r vegetation-calculate-window-points, eval = live_run}
vegetation_points_raw <- calculate_covariates(
  covariate = "modis",
  from = vegetation_files_live,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD13A2"),
  name_covariates = "ndvi_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.0001"
)
vegetation_points_mean <- average_covariates_over_time(
  vegetation_points_raw,
  id_col = "site_id"
)
```

```{r vegetation-calculate-window-polygons, eval = live_run}
vegetation_polygons_raw <- calculate_covariates(
  covariate = "modis",
  from = vegetation_files_live,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD13A2"),
  name_covariates = "ndvi_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.0001"
)
vegetation_polygons_mean <- average_covariates_over_time(
  vegetation_polygons_raw,
  id_col = "h3_id"
)
```

```{r vegetation-plot-raster, eval = live_run, fig.alt = "Raster plot of a 17-day average MOD13A2 NDVI surface."}
terra::plot(
  vegetation_raster_mean,
  main = "MOD13A2 17-day average NDVI"
)
```

```{r vegetation-plot-histogram, eval = live_run, fig.alt = "Histogram of 17-day average MOD13A2 NDVI values extracted at the example points."}
plot_histogram_points(
  vegetation_points_mean,
  "MOD13A2 17-day average point histogram"
)
```

```{r vegetation-plot-points, eval = live_run, fig.alt = "Map of 17-day average MOD13A2 NDVI values extracted at the example points."}
plot_points(
  vegetation_points_mean,
  "MOD13A2 17-day average: point extraction"
)
```

```{r vegetation-plot-polygons, eval = live_run, fig.alt = "Map of 17-day average MOD13A2 NDVI values extracted across Durham County H3 polygons."}
plot_polygons(
  vegetation_polygons_mean,
  "MOD13A2 17-day average: Durham H3 polygons"
)
```

## Workflow templates for the remaining supported product families

The table above covers every product accepted by `download_modis()`. The chunks below show the download -> process -> calculate pattern for each remaining helper family. Swap Terra / Aqua product codes within the same family to move between `MOD*` and `MYD*` variants while keeping the same `amadeus` helper path.

### Surface reflectance grids: `MOD09*` and `MYD09*`

```{r reflectance-template, eval = FALSE}
reflectance_dir <- file.path(tempdir(), "modis_workflow", "reflectance")
download_data(
  dataset_name = "modis",
  product = "MOD09GA",
  version = "061",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = reflectance_dir,
  acknowledgement = TRUE
)

reflectance_files <- list.files(
  reflectance_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

reflectance_raster <- process_modis_merge(
  path = reflectance_files,
  date = "2019-08-15",
  subdataset = process_modis_sds("MOD09GA")
)

reflectance_points <- calculate_covariates(
  covariate = "modis",
  from = reflectance_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD09GA"),
  name_covariates = "reflectance_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(reflectance_raster, main = "MOD09GA processed raster")
plot_histogram_points(reflectance_points, "MOD09GA point covariate histogram")
```

For `MOD09GQ` and `MYD09GQ`, keep the same daily request pattern. For the 8-day composites `MOD09A1`, `MYD09A1`, `MOD09Q1`, and `MYD09Q1`, switch to a short multi-composite window such as `date = c("2019-08-13", "2019-08-29")`, then average the processed rasters and extracted values with `average_modis_window()` and `average_covariates_over_time()`.

### Vegetation indices: `MOD13*` and `MYD13*`

```{r vegetation-template, eval = FALSE}
vegetation_dir <- file.path(tempdir(), "modis_workflow", "vegetation")
vegetation_window <- c("2019-08-13", "2019-08-29")
vegetation_dates <- c("2019-08-13", "2019-08-29")
download_data(
  dataset_name = "modis",
  product = "MOD13A2",
  version = "061",
  date = vegetation_window,
  extent = modis_extent,
  directory_to_save = vegetation_dir,
  acknowledgement = TRUE
)

vegetation_files <- list.files(
  vegetation_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

vegetation_raster <- average_modis_window(
  paths = vegetation_files,
  dates = vegetation_dates,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD13A2")
)

vegetation_points_raw <- calculate_covariates(
  covariate = "modis",
  from = vegetation_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD13A2"),
  name_covariates = "ndvi_",
  fun_summary = "mean",
  geom = "sf",
  scale = "* 0.0001"
)
vegetation_points <- average_covariates_over_time(vegetation_points_raw, "site_id")

terra::plot(vegetation_raster, main = "MOD13A2 processed raster")
plot_histogram_points(vegetation_points, "MOD13A2 point covariate histogram")
```

This same averaging pattern applies to `MOD13A1`, `MYD13A1`, and `MYD13A2` when you use a short window that spans two valid 16-day composites. For the monthly products `MOD13A3` and `MYD13A3`, request a date range covering the target month so the monthly granule is included.

### Fire grids: `MOD14A1`, `MYD14A1`, `MOD14CM1`, and `MYD14CM1`

```{r fire-grid-template, eval = FALSE}
fire_grid_dir <- file.path(tempdir(), "modis_workflow", "fire_grid")
download_data(
  dataset_name = "modis",
  product = "MOD14A1",
  version = "061",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = fire_grid_dir,
  acknowledgement = TRUE
)

fire_grid_files <- list.files(
  fire_grid_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

fire_grid_raster <- process_modis_merge(
  path = fire_grid_files,
  date = "2019-08-15",
  subdataset = process_modis_sds("MOD14A1")
)

fire_grid_points <- calculate_covariates(
  covariate = "modis",
  from = fire_grid_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = process_modis_sds("MOD14A1"),
  name_covariates = "firemask_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(fire_grid_raster, main = "MOD14A1 processed raster")
plot_histogram_points(fire_grid_points, "MOD14A1 point covariate histogram")
```

The monthly climate-modeling grids `MOD14CM1` and `MYD14CM1` use the same merge-and-calculate pattern, but their filenames are monthly. Request a date inside the target month, such as `"2019-08-15"`, and `download_modis()` will automatically request the correct internal version (`005`).

### MAIAC aerosol: `MCD19A2`

```{r maiac-template, eval = FALSE}
maiac_dir <- file.path(tempdir(), "modis_workflow", "maiac")
download_data(
  dataset_name = "modis",
  product = "MCD19A2",
  version = "061",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = maiac_dir,
  acknowledgement = TRUE
)

maiac_files <- list.files(
  maiac_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

maiac_raster <- process_modis_merge(
  path = maiac_files,
  date = "2019-08-15",
  subdataset = "Optical_Depth_047"
)

maiac_points <- calculate_covariates(
  covariate = "modis",
  from = maiac_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_merge,
  subdataset = "Optical_Depth_047",
  name_covariates = "aod_047_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(maiac_raster, main = "MCD19A2 processed raster")
plot_histogram_points(maiac_points, "MCD19A2 point covariate histogram")
```

### Cloud swaths: `MOD06_L2`

```{r cloud-template, eval = FALSE}
cloud_dir <- file.path(tempdir(), "modis_workflow", "cloud")
download_data(
  dataset_name = "modis",
  product = "MOD06_L2",
  version = "061",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = cloud_dir,
  acknowledgement = TRUE
)

cloud_files <- list.files(
  cloud_dir,
  pattern = "\\.hdf$",
  recursive = TRUE,
  full.names = TRUE
)

cloud_raster <- process_modis_swath(
  path = cloud_files,
  date = "2019-08-15",
  subdataset = "Cloud_Fraction",
  suffix = ":mod06:",
  resolution = 0.05
)

cloud_points <- calculate_covariates(
  covariate = "modis",
  from = cloud_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_modis_swath,
  subdataset = "Cloud_Fraction",
  suffix = ":mod06:",
  resolution = 0.05,
  name_covariates = "cloud_fraction_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(cloud_raster, main = "MOD06_L2 processed raster")
plot_histogram_points(cloud_points, "MOD06_L2 point covariate histogram")
```

### Active fire detections: `MCD14DL`

```{r mcd14dl-template, eval = FALSE}
fire_txt_dir <- file.path(tempdir(), "modis_workflow", "mcd14dl")
download_data(
  dataset_name = "modis",
  product = "MCD14DL",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = fire_txt_dir,
  acknowledgement = TRUE
)

fire_txt_files <- list.files(
  fire_txt_dir,
  pattern = "\\.txt$",
  recursive = TRUE,
  full.names = TRUE
)

fire_detections <- process_mcd14dl(
  path = fire_txt_files,
  date = "2019-08-15",
  extent = modis_extent
)

fire_counts <- calculate_covariates(
  covariate = "mcd14dl",
  from = fire_detections,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = c(0L, 1000L),
  geom = "sf"
)

plot_points(fire_detections, "MCD14DL processed detections")
plot_histogram_points(fire_counts, "MCD14DL point covariate histogram")
```

### Nighttime lights: `VNP46A2`

```{r blackmarble-template, eval = FALSE}
blackmarble_dir <- file.path(tempdir(), "modis_workflow", "blackmarble")
download_data(
  dataset_name = "modis",
  product = "VNP46A2",
  date = "2019-08-15",
  extent = modis_extent,
  directory_to_save = blackmarble_dir,
  acknowledgement = TRUE
)

blackmarble_files <- list.files(
  blackmarble_dir,
  pattern = "\\.h5$",
  recursive = TRUE,
  full.names = TRUE
)

blackmarble_raster <- process_blackmarble(
  path = blackmarble_files,
  date = "2019-08-15",
  tile_df = process_blackmarble_corners(hrange = c(8, 10), vrange = c(4, 5)),
  subdataset = 3L,
  crs = "EPSG:4326"
)

blackmarble_points <- calculate_covariates(
  covariate = "modis",
  from = blackmarble_files,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  preprocess = amadeus::process_blackmarble,
  tile_df = process_blackmarble_corners(hrange = c(8, 10), vrange = c(4, 5)),
  subdataset = 3L,
  name_covariates = "blackmarble_",
  fun_summary = "mean",
  geom = "sf"
)

terra::plot(blackmarble_raster, main = "VNP46A2 processed raster")
plot_histogram_points(blackmarble_points, "VNP46A2 point covariate histogram")
```

## Notes

- `calculate_covariates(covariate = "modis", ...)` always receives the raw file list in `from =`; the preprocessing helper named in `preprocess =` is what turns those files into a day-specific raster during extraction.
- `MCD14DL` is the main exception in the MODIS family: it downloads as text, preprocesses with `process_mcd14dl()`, and then extracts with `calculate_covariates(covariate = "mcd14dl", ...)`.
- Because `process_mcd14dl()` returns detections as points rather than a raster, its template plots the processed detections and a histogram of extracted counts instead of a processed raster.
- The example dates are intentionally chosen to align with each representative product cadence: daily products use a single day, while non-daily products are handled with short 15-30 day windows so the plotted rasters and extracted point / H3 summaries are averaged over available composites instead of relying on a single sparse date.
- Apply product-specific scale factors from the NASA product documentation when you move from these templates to analysis-ready covariates. The `MOD11A1` example includes a temperature conversion because that scale is already well established in the package examples.
