---
title: "US EPA National Emissions Inventory (NEI)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{US EPA National Emissions Inventory (NEI)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})


```

This article demonstrates a compact workflow for county-level NEI summaries.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf`, a saved subset of AQS monitor locations from `tests/testdata/aqs/aqs-location-sample.rds`, for point extraction; and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Available inputs and data availability

`download_nei()` is intentionally narrow and currently targets EPA on-road NEI summary releases:

- `year` is the main selector and currently supports the two NEI releases wired into `amadeus`: `2017` and `2020`.
- Requests can include one year or both available years, yielding one annual on-road summary file per year requested.
- NEI downloads arrive as zip archives and can be unzipped automatically with `unzip = TRUE`; the workflow then processes the extracted annual summary tables.
- No jurisdiction argument is needed because the source files are EPA by-region summaries rather than state-by-state download variants.
- NEI does not require authentication. Temporal resolution is annual.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "nei_workflow")
download_data(
  dataset_name = "nei",
  year = 2020L,
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
county <- tigris::counties("NC", year = 2021, cb = TRUE, class = "sf")
data_path = paste0(directory_to_save, "/data_files")
county$FIPS <- county$GEOID
processed_data <- process_covariates(
  covariate = "nei",
  path = data_path,
  county = county,
  year = 2021
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "nei",
  from = data_path,
  locs = example_points_sf,
  locs_id = "site_id",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "nei",
  from = data_path,
  locs = durham_hex,
  locs_id = "h3_id",
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("US EPA National Emissions Inventory (NEI)", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("US EPA National Emissions Inventory (NEI)", ": Durham H3 polygons"))
```

## Notes

- The download request spans both supported NEI years so the live workflow exercises more than one on-road summary release.
- `process_nei()` requires county polygons with a `GEOID` field, which is why the vignette builds counties explicitly before processing.
