---
title: "NOAA GOES Aerosol Detection Product (ADP)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NOAA GOES Aerosol Detection Product (ADP)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)
library(ggplot2)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})


```

This article demonstrates a compact workflow for NOAA GOES Aerosol Detection Product (ADP) data.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Data description

GOES ADP is a geostationary satellite aerosol detection product that reports smoke and dust presence from ABI observations.

| Component | Description |
| --- | --- |
| Products | `ADP-C` (CONUS), `ADP-F` (Full Disk), `ADP-M` (Mesoscale) |
| Variables available in `process_covariates(covariate = "goes", ...)` | `Smoke`, `Dust` |
| Spatial domain | GOES-East (`satellite = "16"`) and GOES-West (`satellite = "18"`) viewing domains; sector depends on product (`C`, `F`, `M`) |
| Spatial resolution | Native GOES fixed-grid product (approximately 2 km at nadir; effective footprint increases away from nadir) |
| Temporal domain | Daily archive by UTC day (`YYYY-MM-DD`), with intra-day granules available throughout the day |
| Temporal resolution | High-frequency geostationary updates (sector-dependent cadence; often every few minutes for CONUS/Full Disk and faster for Mesoscale) |
| Output format | NetCDF (`.nc`) files from NOAA Open Data S3 |

## Available inputs and data availability

`download_data(dataset_name = "goes", ...)` wraps `download_goes()`.

- `satellite` accepts `"16"` (GOES-East) or `"18"` (GOES-West).
- `product` accepts `"ADP-C"` (CONUS), `"ADP-F"` (Full Disk), or `"ADP-M"` (Mesoscale).
- `date` accepts either a single day or a start/end range in `YYYY-MM-DD` format.
- GOES ADP does not require authentication.
- `process_covariates(covariate = "goes", ...)` supports `variable = "Smoke"` or `"Dust"`.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "goes_workflow")
download_goes(
  date = c("2024-01-01","2024-01-01"),
  satellite = "16",
  product = "ADP-C",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
goes_path <- file.path(directory_to_save, "ABI-L2-ADPC", "2024", "001")

processed_mean <- process_covariates(
  covariate = "goes",
  variable = "Smoke",
  date = "2024-01-01",
  path = goes_path,
  daily_agg = TRUE,
  fun = "mean"
)

processed_sum <- process_covariates(
  covariate = "goes",
  variable = "Smoke",
  date = "2024-01-01",
  path = goes_path,
  daily_agg = TRUE,
  fun = "sum"
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
domain_x <- c(terra::xmin(processed_data), terra::xmax(processed_data))
domain_y <- c(terra::ymin(processed_data), terra::ymax(processed_data))
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
example_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))

point_values <- calculate_covariates(
  covariate = "goes",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 0,
  fun = "mean",
  geom = "sf"
)
```

## Visualize the point outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
smoke_col <- grep("^Smoke_", names(point_values), value = TRUE)[1]
if (is.na(smoke_col) || smoke_col == "") {
  excluded <- c("site_id", "h3_id", attr(point_values, "sf_column"))
  fallback_cols <- setdiff(names(point_values), excluded)
  fallback_numeric <- fallback_cols[
    vapply(point_values[fallback_cols], is.numeric, logical(1))
  ]
  smoke_col <- fallback_numeric[1]
}

ggplot(data = point_values) +
  geom_sf(aes(color = .data[[smoke_col]])) +
  ggtitle("Daily GOES Smoke values")
```
