---
title: "NASA SEDAC Population Density"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA SEDAC Population Density}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

```

This article demonstrates a compact workflow for the NASA SEDAC population-density rasters. These downloads require a NASA EarthData token.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

`download_data(dataset_name = "sedac_population", ...)` wraps `download_population()`.

- Inputs are global GPW population-density rasters keyed by `year`; request a single year or `year = "all"` for the all-years product.
- `data_resolution` supports `"30 second"`, `"2.5 minute"`, `"15 minute"`, `"30 minute"`, and `"60 minute"`. The all-years product is not available at 30-second resolution, so `amadeus` automatically falls back to 2.5-minute resolution in that case.
- `data_format` accepts `"GeoTIFF"`, `"ASCII"`, or `"netCDF"`. The all-years product is only available as netCDF and will be downloaded that way even if another format is requested.
- Downloads are global rather than extent-based; crop to your study area during processing.
- Files arrive as zip archives, with `unzip` and `remove_zip` controlling archive handling. Access may require NASA EarthData authentication, and `amadeus` will use `NASA_EARTHDATA_TOKEN` when needed.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "population_workflow")
download_data(
  dataset_name = "sedac_population",
  year = "2020",
  data_format = "GeoTIFF",
  data_resolution = "15 minute",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)

```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "population",
  path = list.files(
    directory_to_save,
    pattern = "\\.tif$",
    recursive = TRUE,
    full.names = TRUE
  )[1]
)
terra::plot(log10(processed_data))
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
df <- data.frame(
  site_id = c("site_1", "site_2", "site_3","site_4"),
  lon = c(-157.8583, -155.5319, -155.5828,-159.5),
  lat = c(21.3069, 19.5, 19.8968, 22.0)
)
example_points_sf <- sf::st_as_sf(
  df,
  coords = c("lon", "lat"),
  crs = 4326
)


point_values <- calculate_covariates(
  covariate = "population",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 1000,
  fun = "mean",
  geom = "sf"
)

print(point_values)
```


## Calculate covariates at polygons and demonstrate the `weighted_mean` function

Here we calculate and compare a simple mean and a population-weighted mean for the same set of polygons. The `weighted_mean` function uses the population density values as weights, so it will give more influence to areas with higher population density within the specified radius.

```{r calculate-pop-weighted, eval = live_run}

nc <- st_read(system.file("shape/nc.shp", package="sf"))
directory_prism <- file.path(tempdir(), "population_workflow","prism")

# Get prism data 
download_data(
  dataset_name = "prism",
  time = "201005",
  element = "tmax",
  data_type = "ts",
  directory_to_save = directory_prism,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)

processed_prism <- process_covariates(
  covariate = "prism",
  path = list.files(
    paste0(directory_prism,"/data_files/"),
    pattern = ".nc",
    recursive = TRUE,
    full.names = TRUE
  )[1],
  element = "tmax",
  time = "201005"
)

pop_weighted_values <- calculate_covariates(
  covariate = "prism",
  from = processed_prism,
  locs = nc,
  locs_id = "FIPS",
  weights = processed_data,
  geom = "sf"
)

print(pop_weighted_values)

area_weighted_values <- calculate_covariates(
  covariate = "prism",
  from = processed_prism,
  locs = nc,
  locs_id = "FIPS",
  geom = "sf"
)

print(area_weighted_values)

```