---
title: "NASA SEDAC Population Density"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NASA SEDAC Population Density}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

example_points <- data.frame(
  site_id = c("durham_armory", "bahama"),
  lon = c(-78.9040, -78.8803),
  lat = c(36.0330, 36.1702)
)
example_points_sf <- sf::st_as_sf(
  example_points,
  coords = c("lon", "lat"),
  crs = 4326
)
durham_hex_path <- system.file(
  "extdata", "data_files", "durham_h3_res8.rds", package = "amadeus"
)
if (durham_hex_path == "") {
  source_candidates <- c(
    file.path("inst", "extdata", "data_files", "durham_h3_res8.rds"),
    file.path("..", "inst", "extdata", "data_files", "durham_h3_res8.rds")
  )
  durham_hex_path <- source_candidates[file.exists(source_candidates)][1]
}
if (is.na(durham_hex_path) || durham_hex_path == "") {
  stop("Could not locate inst/extdata/data_files/durham_h3_res8.rds")
}
durham_hex <- readRDS(durham_hex_path)

pick_value_column <- function(x) {
  geom_col <- attr(x, "sf_column")
  excluded <- c("site_id", "h3_id", "resolution", "area_km2", "name", "NAME", geom_col)
  candidates <- setdiff(names(x), excluded)
  if (length(candidates) == 0) {
    stop("No plottable covariate columns found.")
  }
  preferred <- candidates[vapply(x[candidates], function(col) {
    is.numeric(col) || is.character(col) || is.factor(col)
  }, logical(1))]
  if (length(preferred) > 0) preferred[1] else candidates[1]
}

plot_points <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot(x) +
    ggplot2::geom_sf(ggplot2::aes(color = !!rlang::sym(value_col)), size = 2) +
    ggplot2::labs(title = title, color = value_col) +
    ggplot2::theme_minimal()
}

plot_polygons <- function(x, title) {
  value_col <- pick_value_column(x)
  ggplot2::ggplot(x) +
    ggplot2::geom_sf(ggplot2::aes(fill = !!rlang::sym(value_col)), color = NA) +
    ggplot2::labs(title = title, fill = value_col) +
    ggplot2::theme_minimal()
}
```

This article demonstrates a compact workflow for the NASA SEDAC population-density rasters. These downloads require a NASA EarthData token.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

Each workflow uses two small example surfaces: `example_points_sf` for point extraction and the packaged Durham County Uber H3 resolution-8 hexagons at `system.file("extdata", "data_files", "durham_h3_res8.rds", package = "amadeus")` for polygon extraction.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "population_workflow")
download_data(
  dataset_name = "sedac_population",
  year = "2020",
  data_format = "GeoTIFF",
  data_resolution = "30 second",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
download_data(
  dataset_name = "sedac_population",
  year = "2015",
  data_format = "GeoTIFF",
  data_resolution = "2.5 minute",
  directory_to_save = file.path(directory_to_save, "pop_2015"),
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)
```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "population",
  path = list.files(
    directory_to_save,
    pattern = "\\.tif$",
    recursive = TRUE,
    full.names = TRUE
  )[1],
  extent = terra::ext(-79.2, -78.6, 35.8, 36.3)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "population",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 1000,
  fun = "mean",
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "population",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 1000,
  fun = "mean",
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("NASA SEDAC Population Density", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("NASA SEDAC Population Density", ": Durham H3 polygons"))
```

## Notes

- The live workflow now demonstrates two population density requests by varying both year and resolution.
- `download_data()` uses the historical wrapper name `"sedac_population"`, while processing and calculation use the public-facing `"population"` covariate name.
