---
title: "PRISM Climate Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PRISM Climate Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})


```

This article demonstrates a compact workflow for PRISM climate rasters.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

`download_data(dataset_name = "prism", ...)` combines `time`, `element`, and `data_type`.

| `data_type` | Resolution | Supported `element` values | `time` expectations |
| --- | --- | --- | --- |
| `ts` | 4 km time series | `ppt`, `tmin`, `tmax`, `tmean`, `tdmean`, `vpdmin`, `vpdmax` | `YYYYMMDD` for daily data (1981-01-01 through yesterday), `YYYYMM` for monthly data (1981-01 through last month), or `YYYY` for annual data (1981 through last year) |
| `normals_800` | 800 m normals | `ppt`, `tmin`, `tmax`, `tmean`, `tdmean`, `vpdmin`, `vpdmax`, `solslope`, `soltotal`, `solclear`, `soltrans` | Monthly normals use `MM` or `14` for annual normal; daily normals use `MMDD` |
| `normals` | 4 km normals | `ppt`, `tmin`, `tmax`, `tmean`, `tdmean`, `vpdmin`, `vpdmax`, `solslope`, `soltotal`, `solclear`, `soltrans` | Monthly normals use `MM` or `14` for annual normal; daily normals use `MMDD` |

- Historical PRISM bundles also accept `time = "YYYY"` for years 1895-1980 and return a zip file containing 12 monthly grids plus the annual grid.
- `format` is only used for `data_type = "ts"` and can be `nc`, `asc`, or `grib2`.
- Major constraints:
  - `sol*` elements are available for normals only, not for `ts`.
  - For normals, `format` is ignored.
  - The PRISM API always returns a zip file, even when a time-series format is requested.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "prism_workflow")
download_data(
  dataset_name = "prism",
  time = "201005",
  element = "tmin",
  data_type = "ts",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE,
  unzip = TRUE,
  remove_zip = FALSE
)

```

## Process one workflow-ready data product

```{r process, eval = live_run}
processed_data <- process_covariates(
  covariate = "prism",
  path = list.files(
    paste0(directory_to_save,"/data_files/"),
    pattern = "PRISM_ppt.*201005.*\.(nc|tif|grib2|asc)$",
    recursive = TRUE,
    full.names = TRUE
  )[1],
  element = "ppt",
  time = "201005",
  extent = terra::ext(-162.5, -153.5, 18, 23)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
point_values <- calculate_covariates(
  covariate = "prism",
  from = processed_data,
  locs = example_points_sf,
  locs_id = "site_id",
  radius = 1000,
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "prism",
  from = processed_data,
  locs = durham_hex,
  locs_id = "h3_id",
  radius = 1000,
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("PRISM Climate Data", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("PRISM Climate Data", ": Durham H3 polygons"))
```

## Notes

- The download section now covers two PRISM elements (`ppt` and `tmax`) for the same monthly time stamp.
- Pattern-based file discovery is safer than hard-coding extracted PRISM filenames because the zip contents vary by data type and service response.
