---
title: "Reproducible Unit and Mocked Testing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducible Unit and Mocked Testing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Scope and objective

This article describes how to make the `amadeus` unit and mocked tests more
reproducible while retaining the detailed dataset-specific coverage already in
the package. It focuses on the three public workflow stages:

```text
download_data()
    -> process_covariates()
        -> calculate_covariates()
```

Live tests are intentionally outside the main scope. A live test contacts the
real upstream provider and is useful for detecting URL, authentication, and
file-format changes. A unit or mocked test must instead run without network
access and must produce the same result from the same repository revision and
fixture.

The recommended test structure has two CRAN-safe levels:

1. **Unit tests** exercise one function, validation rule, or branch with small
   controlled inputs.
2. **Mocked integration tests** exercise several real package functions
   together while replacing only external network or expensive I/O boundaries.

The ecoregion examples below are derived from `test-download.R`,
`test-ecoregion.R`, `helper-mocks-download.R`, and the standalone ecoregion
simulation. The EDGAR tests provide an additional model, particularly the
fixture factories and parameterized discovery-to-calculation workflow in
`test-edgar.R`.

# Reproducibility requirements

A reproducible unit or mocked test should satisfy all of the following:

| Requirement | Implementation |
|---|---|
| No external network dependency | Mock URL checks and the download runner |
| Controlled input | Use a small fixture committed under `tests/testdata/` |
| Isolated filesystem | Use `withr::local_tempdir()` or `withr::with_tempdir()` |
| Deterministic locations | Use fixed coordinates or deterministic feature/cell selection |
| Explicit contracts | Check class, dimensions, schema, CRS, IDs, and values |
| Actionable failure messages | Use typed expectations and `info` where needed |
| No persistent side effects | Keep routine test artifacts inside temporary directories |
| Stable test naming | Use `"<fn>(<arg=value>, ...): <expected behavior>"` |
| Separate performance evidence | Record timings diagnostically; do not use fragile wall-time limits |

The same fixture must always yield the same semantic result. Timestamps and
temporary paths may vary, but they must not affect calculated covariates.

# Download tests

## Existing patterns

`test-download.R` provides broad coverage of acknowledgement checks, directory
validation, URL handling, hashes, retries, progress modes, zero-byte files,
existing files, path handling, and argument forwarding. Dataset files such as
`test-edgar.R` also test URL construction and error branches.

A typical live-oriented download test has this form:

```{r download-original-live}
testthat::test_that(
  "download_ecoregion(show_progress=FALSE, hash=FALSE): downloads archive",
  {
    skip_on_cran()
    skip_if_offline()

    withr::with_tempdir({
      result <- download_ecoregion(
        directory_to_save = ".",
        acknowledgement = TRUE,
        download = TRUE,
        unzip = FALSE,
        show_progress = FALSE,
        hash = FALSE
      )

      testthat::expect_null(result)
      testthat::expect_true(dir.exists("zip_files"))
      zip_files <- list.files(
        "zip_files",
        pattern = "\\.zip$",
        full.names = TRUE
      )
      testthat::expect_gt(length(zip_files), 0L)
      testthat::expect_true(all(file.size(zip_files) > 0L))
    })
  }
)
```

This checks real service availability, but its outcome depends on DNS, the
provider, network speed, and the current remote file. It belongs in a live test,
not in the routine mocked suite.

A simple mock commonly replaces the external functions with no-ops:

```{r download-original-mock}
testthat::local_mocked_bindings(
  check_url_status = function(...) TRUE,
  download_run_method = function(...) invisible(NULL),
  download_unzip = function(...) invisible(NULL),
  download_remove_zips = function(...) invisible(NULL),
  download_hash = function(hash, directory) {
    if (isTRUE(hash)) "fakehash" else NULL
  },
  .package = "amadeus"
)
```

This is deterministic, but a no-op downloader and unzipper do not verify the
filesystem contract that connects downloading to processing.

## Improvement: preserve download side effects without using the network

The improved mock should replace the network boundary while preserving the
expected archive and extracted-data layout:

```{r download-improved-mock}
fixture_path <- testthat::test_path(
  "..",
  "testdata",
  "ecoregions",
  "eco_l3_clip.gpkg"
)
download_root <- withr::local_tempdir()

archive_url <- NULL
archive_destination <- NULL
unzip_destination <- NULL
download_call_count <- 0L

local_download_mocks(
  download_run_method = function(urls, destfiles, ...) {
    archive_url <<- urls
    archive_destination <<- destfiles
    download_call_count <<- download_call_count + length(urls)

    dir.create(
      dirname(destfiles),
      recursive = TRUE,
      showWarnings = FALSE
    )
    writeBin(charToRaw("mock ecoregion archive"), destfiles)

    list(
      success = length(destfiles),
      failed = 0L,
      skipped = 0L
    )
  },
  download_unzip = function(
    file_name,
    directory_to_unzip,
    ...
  ) {
    unzip_destination <<- directory_to_unzip
    dir.create(
      directory_to_unzip,
      recursive = TRUE,
      showWarnings = FALSE
    )

    copied <- file.copy(
      fixture_path,
      file.path(directory_to_unzip, basename(fixture_path)),
      overwrite = TRUE
    )
    if (!all(copied)) {
      stop("Could not install the ecoregion fixture.", call. = FALSE)
    }

    invisible(NULL)
  },
  download_remove_zips = function(...) invisible(NULL)
)
```

Specific changes from the simple mock are:

| Original | Improved | Reason |
|---|---|---|
| `function(...) invisible(NULL)` | Capture `urls` and `destfiles` | Make the requested operation observable |
| Return no download status | Return typed success/failure/skip counts | Preserve the internal function contract |
| Create no archive | Write a small fake archive at `destfiles` | Validate archive path construction |
| Unzip is a no-op | Copy a real fixture to `data_files/` | Validate the download-to-process boundary |
| Assume copying succeeded | Stop when `file.copy()` fails | Prevent false-positive tests |
| Report a hard-coded count | Increment `download_call_count` | Keep diagnostic counts synchronized with behavior |

The fake archive is deliberately not a valid ZIP. The unzip operation is mocked
and real archive extraction is tested separately. The extracted GPKG is real,
because it becomes input to the real process function.

## Improvement: exercise the public download dispatcher

Some dataset tests call `download_ecoregion()` directly. Keep those focused unit
tests, but add one mocked integration test through the public dispatcher:

```{r download-improved-dispatch}
testthat::test_that(
  "download_data(dataset_name=ecoregion): creates archive and extracted fixture",
  {
    # Install the side-effect-preserving mocks defined above.

    result <- amadeus::download_data(
      dataset_name = "ecoregion",
      directory_to_save = download_root,
      acknowledgement = TRUE,
      hash = FALSE,
      show_progress = FALSE,
      rate_limit = 0
    )

    extracted_path <- file.path(
      unzip_destination,
      basename(fixture_path)
    )

    testthat::expect_null(result)
    testthat::expect_identical(download_call_count, 1L)
    testthat::expect_match(
      archive_url,
      "us_eco_l3_state_boundaries\\.zip$"
    )
    testthat::expect_identical(
      basename(archive_destination),
      "us_eco_l3_state_boundaries.zip"
    )
    testthat::expect_true(file.exists(archive_destination))
    testthat::expect_gt(file.size(archive_destination), 0L)
    testthat::expect_true(file.exists(extracted_path))
    testthat::expect_identical(
      unname(tools::md5sum(extracted_path)),
      unname(tools::md5sum(fixture_path))
    )
  }
)
```

This test adds evidence for all of the following:

- `download_data()` dispatches `"ecoregion"` correctly.
- Arguments reach the selected dataset function.
- The correct URL and archive name are constructed.
- The archive is written to the expected directory.
- Extraction produces the expected process input.
- The copied fixture is byte-for-byte identical to the source fixture.

It does not replace unit tests for acknowledgement, invalid paths, hashes,
existing files, retry behavior, or other datasets.

## Download validation checklist

A dataset-specific mocked download test should check, where applicable:

```{r download-contract}
testthat::expect_length(urls, expected_file_count)
testthat::expect_length(destfiles, expected_file_count)
testthat::expect_true(all(grepl("^https://", urls)))
testthat::expect_false(anyDuplicated(destfiles) > 0L)
testthat::expect_true(all(file.exists(destfiles)))
testthat::expect_true(all(file.size(destfiles) > 0L))
testthat::expect_true(all(file.exists(extracted_files)))
```

For parameterized datasets such as EDGAR, also verify that the discovery matrix
and file matrix agree:

```{r download-edgar-contract}
testthat::expect_equal(length(discovery$urls), discovery$n_files)
testthat::expect_equal(length(discovery$destfiles), discovery$n_files)
testthat::expect_false(anyDuplicated(discovery$destfiles) > 0L)
```

# Process tests

## Existing pattern

The existing ecoregion process test verifies that the fixture can be read and
that a nonempty, valid `SpatVector` is returned:

```{r process-original}
path_eco <- testthat::test_path(
  "..",
  "testdata",
  "ecoregions",
  "eco_l3_clip.gpkg"
)

testthat::expect_no_error(
  eco <- process_ecoregion(path_eco)
)
testthat::expect_s4_class(eco, "SpatVector")
testthat::expect_true(all(terra::is.valid(eco)))
testthat::expect_gt(terra::nrow(eco), 0L)
```

These checks are useful, but `expect_no_error()` plus a class check does not
prove that the output retains the schema, CRS, time information, and data needed
by calculation.

## Improvement: validate the process contract

```{r process-improved-contract}
testthat::test_that(
  "process_covariates(covariate=ecoregion): returns calculation-ready data",
  {
    fixture_path <- testthat::test_path(
      "..",
      "testdata",
      "ecoregions",
      "eco_l3_clip.gpkg"
    )
    fixture_hash_before <- unname(tools::md5sum(fixture_path))

    processed <- amadeus::process_covariates(
      covariate = "ecoregion",
      path = fixture_path
    )

    required_fields <- c(
      "L2_KEY",
      "L3_KEY",
      "NA_L2NAME",
      "US_L3NAME",
      "NA_L3NAME",
      "time"
    )
    missing_fields <- setdiff(required_fields, names(processed))

    testthat::expect_s4_class(processed, "SpatVector")
    testthat::expect_gt(terra::nrow(processed), 0L)
    testthat::expect_true(all(terra::is.valid(processed)))
    testthat::expect_length(
      missing_fields,
      0L,
      info = paste(
        "Missing calculation fields:",
        paste(missing_fields, collapse = ", ")
      )
    )
    testthat::expect_true(nzchar(terra::crs(processed)))
    testthat::expect_false(anyNA(processed$L3_KEY))
    testthat::expect_identical(
      unname(tools::md5sum(fixture_path)),
      fixture_hash_before,
      info = "Processing must not modify its source fixture."
    )
  }
)
```

The specific improvements are:

| Original assertion | Improved assertion | Failure detected |
|---|---|---|
| No unexpected error | Typed spatial class | Incorrect return type |
| Row count is positive | Row count plus key checks | Empty or malformed features |
| Geometry is valid | Geometry validity plus CRS | Invalid or unreferenced spatial output |
| No schema check | Required-field contract | Fields dropped or renamed during processing |
| No input-side-effect check | Fixture hash before and after | Accidental modification of test data |

Do not assert a specific CRS unless the function contract requires one. Instead,
assert that it is nonempty and, where required, compare it with the documented
target CRS.

## Improvement: compare wrapper and direct dispatch

At least one test per dataset should verify that the public wrapper selects the
same implementation as the direct function:

```{r process-dispatch-parity}
via_wrapper <- amadeus::process_covariates(
  covariate = "ecoregion",
  path = fixture_path
)
via_direct <- amadeus::process_ecoregion(
  path = fixture_path
)

testthat::expect_identical(
  names(via_wrapper),
  names(via_direct)
)
testthat::expect_equal(
  terra::nrow(via_wrapper),
  terra::nrow(via_direct)
)
testthat::expect_true(
  terra::same.crs(via_wrapper, via_direct)
)
```

This detects dispatch errors without duplicating every dataset-specific process
test at the wrapper level.

## Raster process contract

For EDGAR and other raster datasets, use a raster-specific contract:

```{r process-raster-contract}
processed <- amadeus::process_edgar(path = raster_path)

testthat::expect_s4_class(processed, "SpatRaster")
testthat::expect_gt(terra::nlyr(processed), 0L)
testthat::expect_gt(terra::ncell(processed), 0L)
testthat::expect_true(nzchar(terra::crs(processed)))
testthat::expect_false(anyDuplicated(names(processed)) > 0L)
testthat::expect_true(all(grepl("^edgar_", names(processed))))
testthat::expect_true(any(!is.na(terra::values(processed))))
```

The existing EDGAR tests already check the `SpatRaster` class, layer count,
informative names, and parsed time. Adding CRS, duplicate-name, and non-NA-value
checks would complete the calculation-ready contract.

## Process determinism

For a small fixture, run processing twice and compare semantic properties:

```{r process-determinism}
processed_first <- amadeus::process_ecoregion(fixture_path)
processed_second <- amadeus::process_ecoregion(fixture_path)

testthat::expect_identical(
  names(processed_first),
  names(processed_second)
)
testthat::expect_equal(
  terra::nrow(processed_first),
  terra::nrow(processed_second)
)
testthat::expect_true(
  terra::same.crs(processed_first, processed_second)
)
testthat::expect_equal(
  as.data.frame(processed_first),
  as.data.frame(processed_second)
)
```

This catches hidden dependence on row ordering, session state, or nondeterministic
parallel execution. It should only be used with sufficiently small fixtures.

# Calculate tests

## Existing pattern

The existing ecoregion test uses one hard-coded site and checks specific output
columns:

```{r calculate-original}
site_faux <- data.frame(
  site_id = "37999109988101",
  lon = -77.576,
  lat = 39.40,
  date = as.Date("2022-01-01")
)

site_faux <- terra::vect(
  site_faux,
  geom = c("lon", "lat"),
  keepgeom = TRUE,
  crs = "EPSG:4326"
)
site_faux <- terra::project(site_faux, "EPSG:5070")

ecor_res <- calculate_ecoregion(
  from = processed,
  locs = site_faux,
  locs_id = "site_id"
)

testthat::expect_s3_class(ecor_res, "data.frame")
testthat::expect_equal(
  colnames(ecor_res)[-(1:2)],
  c("DUM_E2083_00000", "DUM_E3064_00000")
)
```

The exact expected columns are a valuable semantic oracle, but one location does
not validate row preservation, identifier uniqueness, or behavior across more
than one ecoregion.

## Improvement: deterministic multi-location probes

```{r calculate-deterministic-locations}
sample_size <- min(3L, terra::nrow(processed))
centroids <- terra::centroids(
  processed[seq_len(sample_size), ]
)
coordinates <- terra::crds(centroids)

locations <- terra::vect(
  data.frame(
    site_id = sprintf("site_%02d", seq_len(sample_size)),
    x = coordinates[, 1L],
    y = coordinates[, 2L]
  ),
  geom = c("x", "y"),
  crs = terra::crs(processed)
)
```

This changes the test as follows:

| Original | Improved | Benefit |
|---|---|---|
| One manually selected point | Three feature-derived points | Exercises multiple result rows |
| Manual CRS assignment and projection | Inherit processed CRS | Avoids CRS mismatch in test setup |
| Coordinate may leave fixture after fixture edits | Centroids remain within selected features | More robust fixture evolution |
| One site identifier | Stable `site_01`, `site_02`, ... IDs | Enables order and uniqueness checks |

The selection is deterministic; no random-number seed is required.

For raster datasets, select fixed cell centers instead:

```{r calculate-raster-locations}
probe_cells <- unique(c(
  1L,
  ceiling(terra::ncell(processed) / 2),
  terra::ncell(processed)
))
probe_xy <- terra::xyFromCell(processed, probe_cells)

locations <- terra::vect(
  data.frame(
    site_id = sprintf("site_%02d", seq_along(probe_cells)),
    x = probe_xy[, 1L],
    y = probe_xy[, 2L]
  ),
  geom = c("x", "y"),
  crs = terra::crs(processed)
)
```

## Improvement: validate calculation invariants

```{r calculate-improved-contract}
calculated <- amadeus::calculate_covariates(
  covariate = "ecoregion",
  from = processed,
  locs = locations,
  locs_id = "site_id",
  frac = FALSE,
  drop = TRUE
)

indicator_names <- grep(
  "^DUM_E[23]",
  names(calculated),
  value = TRUE
)

testthat::expect_s3_class(calculated, "data.frame")
testthat::expect_equal(nrow(calculated), sample_size)
testthat::expect_identical(
  calculated$site_id,
  locations$site_id
)
testthat::expect_length(
  unique(calculated$site_id),
  sample_size
)
testthat::expect_gt(length(indicator_names), 0L)
testthat::expect_true(
  all(vapply(
    calculated[indicator_names],
    function(values) {
      all(values %in% c(0L, 1L), na.rm = TRUE)
    },
    logical(1)
  ))
)
```

These assertions detect:

- Incorrect return class.
- Missing or duplicated rows.
- Lost or reordered location identifiers.
- Missing indicator columns.
- Invalid indicator values outside the binary domain.

For EDGAR, replace the binary-domain assertion with numeric and nonmissing-value
contracts:

```{r calculate-edgar-contract}
value_columns <- setdiff(names(calculated), "site_id")

testthat::expect_s3_class(calculated, "data.frame")
testthat::expect_identical(calculated$site_id, locations$site_id)
testthat::expect_gt(length(value_columns), 0L)
testthat::expect_true(
  all(vapply(calculated[value_columns], is.numeric, logical(1)))
)
testthat::expect_true(
  any(!is.na(as.matrix(calculated[value_columns])))
)
```

Only assert that values are nonnegative if the documented dataset contract
guarantees nonnegative values.

## Improvement: combine invariants with exact semantic expectations

Dynamic invariant checks should not replace known-answer tests. Keep at least
one small oracle whose expected classifications or values are explicit:

```{r calculate-known-answer}
expected_positive_indicators <- list(
  site_01 = c("DUM_E2083_00000", "DUM_E3045_00000"),
  site_02 = c("DUM_E2083_00000", "DUM_E3064_00000"),
  site_03 = c("DUM_E2083_00000", "DUM_E3064_00000")
)

positive_indicators <- lapply(
  seq_len(nrow(calculated)),
  function(row_number) {
    sort(indicator_names[vapply(
      indicator_names,
      function(column_name) {
        value <- calculated[[column_name]][row_number]
        !is.na(value) && value > 0
      },
      logical(1)
    )])
  }
)
names(positive_indicators) <- calculated$site_id

testthat::expect_identical(
  positive_indicators,
  expected_positive_indicators
)
```

The invariant test answers “is the output structurally valid?” The known-answer
test answers “is the scientific classification still correct?” Both are needed.

## Improvement: wrapper and direct-function parity

```{r calculate-dispatch-parity}
via_wrapper <- amadeus::calculate_covariates(
  covariate = "ecoregion",
  from = processed,
  locs = locations,
  locs_id = "site_id",
  frac = FALSE,
  drop = TRUE
)

via_direct <- amadeus::calculate_ecoregion(
  from = processed,
  locs = locations,
  locs_id = "site_id",
  frac = FALSE,
  drop = TRUE
)

testthat::expect_equal(via_wrapper, via_direct)
```

One parity test is sufficient. Dataset-specific tests should continue to cover
the full argument matrix directly, including fraction, radius, geometry-return,
column-naming, missing-location, and invalid-input branches.

# Mocked end-to-end workflow test

After the three stage contracts are tested separately, add one compact mocked
integration test. The test should reuse the side-effect-preserving download mock
and keep processing and calculation real:

```{r mocked-workflow}
testthat::test_that(
  "ecoregion workflow(n=3): mocked download feeds process and calculate",
  {
    download_root <- withr::local_tempdir()

    # Install local_download_mocks() so the fake archive and real fixture are
    # created under download_root without a network request.

    amadeus::download_data(
      dataset_name = "ecoregion",
      directory_to_save = download_root,
      acknowledgement = TRUE,
      hash = FALSE,
      show_progress = FALSE,
      rate_limit = 0
    )

    raw_path <- file.path(
      unzip_destination,
      basename(fixture_path)
    )
    testthat::expect_true(file.exists(raw_path))

    processed <- amadeus::process_covariates(
      covariate = "ecoregion",
      path = raw_path
    )
    testthat::expect_s4_class(processed, "SpatVector")

    # Construct deterministic locations from processed features.
    calculated <- amadeus::calculate_covariates(
      covariate = "ecoregion",
      from = processed,
      locs = locations,
      locs_id = "site_id",
      frac = FALSE,
      drop = TRUE
    )

    testthat::expect_s3_class(calculated, "data.frame")
    testthat::expect_identical(
      calculated$site_id,
      locations$site_id
    )
  }
)
```

This does not replace the stage-specific tests. Its purpose is to prove that the
output contract of each stage satisfies the input contract of the next stage.

# Efficiency of the routine test suite

Reproducibility and efficiency reinforce one another. Apply these practices to
the default test suite:

1. Use small committed fixtures instead of full datasets.
2. Mock only external boundaries; keep package transformations real.
3. Process a fixture once per test block and reuse the result for option checks.
4. Use parameterized case lists rather than duplicating nearly identical setup.
5. Do not sleep for rate-limit tests; mock the sleep boundary and inspect its
   requested duration.
6. Keep actual network and full-data comparisons in `test-<dataset>-live.R`.
7. Avoid loading `sf`, `terra`, or other large packages in tests that do not use
   them.
8. Avoid `expect_no_error()` around calls followed by stronger typed
   expectations; unexpected errors already fail the test.

The EDGAR discovery matrix is a good parameterized example:

```{r efficient-case-matrix}
discovery_cases <- list(
  yearly_sector = list(
    species = "CO",
    temp_res = "yearly",
    sector_yearly = "ENE",
    year_range = 2021
  ),
  monthly_sector = list(
    species = "SO2",
    temp_res = "monthly",
    sector_monthly = "BUILDINGS"
  ),
  timeseries = list(
    species = "NOx",
    temp_res = "timeseries"
  )
)
```

Each case should be labeled in assertion diagnostics so a failure identifies the
input combination.

# Performance diagnostics

Timing code measures performance; it does not itself make the package faster.
Do not place strict elapsed-time assertions in routine unit tests because CI
hardware, filesystem load, and spatial-library builds vary.

For a standalone reproducibility run, time each stage separately:

```{r performance-diagnostics}
download_timing <- system.time(
  amadeus::download_data(...)
)
process_timing <- system.time(
  processed <- amadeus::process_covariates(...)
)
calculate_timing <- system.time(
  calculated <- amadeus::calculate_covariates(...)
)

performance <- data.frame(
  stage = c("download", "process", "calculate"),
  elapsed_seconds = c(
    unname(download_timing[["elapsed"]]),
    unname(process_timing[["elapsed"]]),
    unname(calculate_timing[["elapsed"]])
  )
)
```

Use repeated measurements and medians before drawing performance conclusions:

```{r repeated-performance}
measure_elapsed <- function(operation, iterations = 5L) {
  timings <- replicate(
    iterations,
    system.time(operation())[["elapsed"]]
  )

  data.frame(
    iterations = iterations,
    minimum = min(timings),
    median = stats::median(timings),
    maximum = max(timings)
  )
}
```

Run performance diagnostics outside the routine unit suite, save the raw
measurements, and compare like-for-like fixtures and environments.

# Reproducibility metadata for standalone runs

Routine tests should leave no persistent files. A standalone simulation or
benchmark should save enough provenance to reproduce the result:

```{r provenance}
provenance <- data.frame(
  field = c(
    "timestamp_utc",
    "r_version",
    "platform",
    "amadeus_version",
    "git_commit",
    "fixture_path",
    "fixture_md5"
  ),
  value = c(
    format(Sys.time(), tz = "UTC"),
    R.version.string,
    R.version$platform,
    as.character(utils::packageVersion("amadeus")),
    system2(
      "git",
      c("-C", repository, "rev-parse", "HEAD"),
      stdout = TRUE
    ),
    fixture_path,
    unname(tools::md5sum(fixture_path))
  )
)
```

At minimum, save:

- The Git commit.
- Whether the worktree contained uncommitted changes.
- R and package versions.
- Platform and relevant spatial-library versions.
- Fixture path and checksum.
- Exact function arguments.
- Per-stage timing.
- Contract results and calculated output.

# Recommended file organization

Use the following division for each dataset:

```text
tests/testthat/
|-- helper-fixtures.R
|-- helper-mocks-download.R
|-- test-<dataset>.R
|-- test-<dataset>-download-mock.R
`-- test-<dataset>-live.R
```

Responsibilities are:

| File | Responsibility |
|---|---|
| `helper-fixtures.R` | Small deterministic input factories and committed-fixture paths |
| `helper-mocks-download.R` | Reusable network and download-boundary mocks |
| `test-<dataset>.R` | Process, calculate, validation, and dataset-specific unit tests |
| `test-<dataset>-download-mock.R` | Network-free download and mocked workflow integration |
| `test-<dataset>-live.R` | Explicitly gated real-provider integration |

# Adoption checklist

For each dataset, adopt the changes in this order:

1. Identify or create the smallest scientifically representative fixture.
2. Add typed unit assertions for download discovery and validation.
3. Add a side-effect-preserving download mock.
4. Validate process class, dimensions, schema, CRS, validity, names, and time.
5. Create deterministic point or polygon probes.
6. Validate calculation class, IDs, row count, names, value domain, and one
   known-answer result.
7. Add one public-wrapper parity test per stage.
8. Add one compact mocked download-to-calculate integration test.
9. Keep network calls and full datasets out of the routine suite.
10. Record performance and provenance in a separate diagnostic workflow.

# Ordered implementation and comparison plan

The following plan translates the checklist into concrete edits. Apply one
change at a time, run the focused tests after each change, and record both test
time and the additional contracts exercised. A stronger test can perform a few
more local operations than a no-op test; efficiency must therefore be evaluated
using reliability and coverage as well as elapsed time.

## Step 1: extend the reusable download helper

**Current code:** `helper-mocks-download.R` returns statuses but leaves download
and unzip as no-ops:

```{r adoption-helper-before}
download_run_method = function(...) {
  list(success = success, failed = failed, skipped = skipped)
},
download_unzip = function(...) NULL
```

**Recommended addition:** retain these fast defaults and add a separate fixture
installer for tests that need a real download-to-process boundary:

```{r adoption-helper-after}
local_fixture_download_mocks <- function(
  fixture_path,
  downloaded_name = basename(fixture_path),
  envir = parent.frame()
) {
  local_download_mocks(
    download_run_method = function(urls, destfiles, ...) {
      vapply(
        destfiles,
        function(destination) {
          dir.create(
            dirname(destination),
            recursive = TRUE,
            showWarnings = FALSE
          )
          writeBin(charToRaw("mock archive"), destination)
          TRUE
        },
        logical(1)
      )
      list(
        success = length(destfiles),
        failed = 0L,
        skipped = 0L
      )
    },
    download_unzip = function(
      file_name,
      directory_to_unzip,
      ...
    ) {
      dir.create(
        directory_to_unzip,
        recursive = TRUE,
        showWarnings = FALSE
      )
      copied <- file.copy(
        fixture_path,
        file.path(directory_to_unzip, downloaded_name),
        overwrite = TRUE
      )
      if (!all(copied)) {
        stop("Could not install fixture.", call. = FALSE)
      }
      invisible(NULL)
    },
    download_remove_zips = function(...) invisible(NULL),
    envir = envir
  )
}
```

**Expected comparison:** the original helper should remain fastest for narrow
return-value tests. The fixture helper should additionally create an archive and
process-ready file. Do not judge this step only by milliseconds; compare the
number of verified filesystem and stage-boundary contracts.

## Step 2: add one network-free dataset download test

**Current code:** a live ecoregion test calls `download_ecoregion()` and depends
on the upstream service:

```{r adoption-download-before}
skip_on_cran()
skip_if_offline()
result <- download_ecoregion(
  directory_to_save = ".",
  acknowledgement = TRUE,
  unzip = FALSE,
  show_progress = FALSE,
  hash = FALSE
)
```

**Recommended addition:** put a focused test in
`test-ecoregion-download-mock.R`, install the fixture mock, and call the public
dispatcher:

```{r adoption-download-after}
testthat::test_that(
  "download_data(dataset_name=ecoregion): creates process-ready fixture",
  {
    output <- withr::local_tempdir()
    local_fixture_download_mocks(fixture_path)

    result <- amadeus::download_data(
      dataset_name = "ecoregion",
      directory_to_save = output,
      acknowledgement = TRUE,
      hash = FALSE,
      show_progress = FALSE,
      rate_limit = 0
    )

    archive <- file.path(
      output,
      "zip_files",
      "us_eco_l3_state_boundaries.zip"
    )
    extracted <- file.path(
      output,
      "data_files",
      basename(fixture_path)
    )

    testthat::expect_null(result)
    testthat::expect_true(file.exists(archive))
    testthat::expect_gt(file.size(archive), 0L)
    testthat::expect_true(file.exists(extracted))
    testthat::expect_identical(
      unname(tools::md5sum(extracted)),
      unname(tools::md5sum(fixture_path))
    )
  }
)
```

**Expected comparison:** the live test validates the provider; the new mocked
test validates package orchestration. The mocked test should pass offline on
every routine run. Retain one separately gated live test rather than replacing
provider validation entirely.

## Step 3: use the smallest representative fixture

**Current approach:** live tests or local scripts may process the complete
national file.

**Recommended change:** use the committed GPKG for routine tests and document its
checksum:

```{r adoption-fixture}
fixture_path <- testthat::test_path(
  "..",
  "testdata",
  "ecoregions",
  "eco_l3_clip.gpkg"
)
fixture_md5 <- unname(tools::md5sum(fixture_path))
```

**Expected comparison:** record file size, feature count, and processing time for
the fixture and full file. The fixture should be substantially faster while
retaining every field and classification used by the unit test. A full-data run
is diagnostic evidence, not a routine test dependency.

## Step 4: replace minimal process checks with a stage contract

**Current code:**

```{r adoption-process-before}
testthat::expect_no_error(
  processed <- process_ecoregion(fixture_path)
)
testthat::expect_s4_class(processed, "SpatVector")
testthat::expect_gt(terra::nrow(processed), 0L)
```

**Recommended code:**

```{r adoption-process-after}
processed <- amadeus::process_covariates(
  covariate = "ecoregion",
  path = fixture_path
)
missing_fields <- setdiff(required_fields, names(processed))

testthat::expect_s4_class(processed, "SpatVector")
testthat::expect_gt(terra::nrow(processed), 0L)
testthat::expect_true(all(terra::is.valid(processed)))
testthat::expect_true(nzchar(terra::crs(processed)))
testthat::expect_length(missing_fields, 0L)
testthat::expect_false(anyNA(processed$L3_KEY))
```

**Expected comparison:** processing time should remain approximately the same,
because both versions call the process function once. The recommended version
checks more failure modes and should identify a schema or CRS regression before
calculation begins.

## Step 5: reuse each processed object

**Inefficient pattern to avoid:**

```{r adoption-reuse-before}
testthat::expect_s4_class(
  process_ecoregion(fixture_path),
  "SpatVector"
)
testthat::expect_gt(
  terra::nrow(process_ecoregion(fixture_path)),
  0L
)
testthat::expect_true(
  all(terra::is.valid(process_ecoregion(fixture_path)))
)
```

**Recommended pattern:**

```{r adoption-reuse-after}
processed <- process_ecoregion(fixture_path)

testthat::expect_s4_class(processed, "SpatVector")
testthat::expect_gt(terra::nrow(processed), 0L)
testthat::expect_true(all(terra::is.valid(processed)))

calc_zero <- calculate_ecoregion(
  from = processed,
  locs = locations,
  radius = 0
)
calc_buffer <- calculate_ecoregion(
  from = processed,
  locs = locations,
  radius = 1000
)
```

**Expected comparison:** count process calls in addition to timing. The first
pattern performs three reads and transformations; the recommended pattern
performs one and reuses the result. Existing tests that already process once and
reuse the object require no change.

## Step 6: add one linked mocked workflow

**Current separated pattern:** the download mock returns success, while the
process test independently reads `fixture_path`. Both tests can pass even if the
download output path is incompatible with processing.

```{r adoption-workflow-before}
local_download_mocks()
download_ecoregion(...)

# In a separate test:
processed <- process_ecoregion(fixture_path)
```

**Recommended linked pattern:**

```{r adoption-workflow-after}
local_fixture_download_mocks(fixture_path)
amadeus::download_data(
  dataset_name = "ecoregion",
  directory_to_save = output,
  acknowledgement = TRUE,
  rate_limit = 0
)

raw_path <- file.path(
  output,
  "data_files",
  basename(fixture_path)
)
processed <- amadeus::process_covariates(
  covariate = "ecoregion",
  path = raw_path
)
calculated <- amadeus::calculate_covariates(
  covariate = "ecoregion",
  from = processed,
  locs = locations,
  locs_id = "site_id",
  drop = TRUE
)
```

**Expected comparison:** the key new metric is
`download_output_feeds_process = TRUE`. Elapsed time may be slightly higher than
separated no-op mocks because the recommended test writes and reads a real local
fixture. It remains independent of the network and provides much stronger
workflow evidence.

## Step 7: parameterize repeated cases

**Current duplicated pattern:** separate blocks repeat fixture construction,
processing, and nearly identical assertions for each option.

**Recommended pattern:** use a named case matrix, as in the EDGAR workflow test:

```{r adoption-cases}
cases <- list(
  zero_radius = list(radius = 0, suffix = "_0"),
  buffered = list(radius = 1000, suffix = "_1000")
)

for (case_name in names(cases)) {
  current <- cases[[case_name]]
  result <- calculate_edgar(
    from = processed,
    locs = locations,
    locs_id = "site_id",
    radius = current$radius
  )
  value_names <- setdiff(names(result), "site_id")
  testthat::expect_true(
    all(endsWith(value_names, current$suffix)),
    info = paste("Failed case:", case_name)
  )
}
```

**Expected comparison:** runtime may be similar, but setup duplication and the
number of repeated process calls should decline. Maintenance becomes safer
because a new case requires one list entry rather than another full test block.

## Step 8: compare performance outside routine assertions

For a stage-by-stage comparison of the original `test-*.R` blocks with the
recommended download, process, and calculate tests for ecoregion, drought, and
CropScape, see
[Original versus recommended Amadeus tests](original-vs-recommended-test-comparison.md).
For the ordered implementation checklist, verification gates, commit sequence,
and pull-request structure, see
[Incremental testing implementation plan](testing-implementation-plan.md).

**Current behavior:** routine tests usually report only pass, fail, and skip.

**Recommended diagnostic:** measure each strategy repeatedly, save every run,
and compare medians. Do not repeatedly benchmark a live download.

The repository includes a runnable, network-free comparison script:

```{sh adoption-run-comparison, eval=FALSE}
Rscript vignettes/scripts/compare-test-strategies.r \
  . \
  5 \
  /tmp/amadeus-test-strategy-comparison
```

The equivalent named form is less sensitive to argument order:

```{sh adoption-run-comparison-named, eval=FALSE}
Rscript vignettes/scripts/compare-test-strategies.r \
  --repository=. \
  --iterations=5 \
  --output=/tmp/amadeus-test-strategy-comparison \
  --dataset=ecoregion,drought,cropscape
```

`drought` expands to the three distinct specifications `spei`, `eddi`, and
`usdm`. Use `--dataset=all` for every implemented specification, or select one
directly, for example `--dataset=spei`. The default remains `ecoregion` for
backward compatibility.

When the default of five iterations is acceptable, the output directory can be
the second positional argument:

```{sh adoption-run-comparison-default-iterations, eval=FALSE}
Rscript vignettes/scripts/compare-test-strategies.r \
  . \
  /tmp/amadeus-test-strategy-comparison
```

It runs two strategies:

| Strategy | Download | Process input | Calculation probes |
|---|---|---|---|
| `original_separated` | Successful no-op mock | Fixture prepared independently | Dataset-specific deterministic locations |
| `recommended_linked` | Process-ready file or extracted fixture | Exact simulated download output | The same dataset-specific calculation contract |

The runner defines a `strategy_specs` registry. Each entry supplies only the
parts that vary by source: download arguments, fixture installation, process
input discovery, processing contracts, calculation locations, and calculation
contracts. Timing, repetition, mock setup, summaries, and report writing occur
once in the shared runner. Adding another source therefore requires one new
specification rather than a copy of the benchmarking script.

| Specification | Simulated download output | Real processed object | Main calculation contract |
|---|---|---|---|
| `ecoregion` | ZIP plus extracted GPKG | `SpatVector` | Binary ecoregion indicators and preserved IDs |
| `spei` | `spei01.nc` | Three-layer `SpatRaster` | Three dated rows per location and `spei_01_0` |
| `eddi` | Dated ASCII grid | One-layer `SpatRaster` | Correct filename-derived date and `eddi_01_0` |
| `usdm` | Two ZIPs plus complete shapefile bundles | Two-date `SpatVector` | Drought class plus bounded 1,000 m proportions summing to one |
| `cropscape` | `tar.gz` plus extracted TIFF | Classified `SpatRaster` | Bounded 300 m class fractions summing to one |

The active `test-cropscape.R` implements the CropScape specification for both
GMU and USDA archive layouts. Its focused tests additionally cover exact year
metadata, raster values and CRS, zero-radius crop classes, `sf` geometry,
location-ID preservation, and nonmissing 300 m fractions.

The USDM specification deliberately constructs locations from USDM polygon
centroids. Such locations retain source attributes named `DM`, `date`, and
`source`, which previously collided with identically named fields during
`terra::intersect()`. The package now reduces the buffered location object to a
stable internal row identifier before intersection. This keeps site-to-result
mapping while ensuring that `DM` always refers to the authoritative drought
polygon. The regression test verifies both dates, the dominant class, five
proportion columns, bounded values, and row sums of one.

The `recommended_linked` strategy is the **linked simulation**. Simulation does
not mean that the whole workflow is fake. Only external boundaries are replaced;
the package workflow remains real:

| Component | Mode | What the code does | Produced evidence |
|---|---|---|---|
| `check_url_status()` | Simulated | Returns `TRUE` without contacting the provider | No network dependency |
| `check_destfile()` | Simulated | Returns `TRUE` so the download branch runs | Predictable branch selection |
| `download_run_method()` | Simulated | Writes a small archive or process-ready raster | Dataset-specific local download artifact |
| `download_unzip()` / `archive_extract()` | Simulated | Installs a committed GPKG, TIFF, or shapefile bundle | Process-ready extracted files |
| `download_remove_zips()` | Simulated | Does nothing | Archive remains inspectable |
| `download_data()` | Real | Dispatches to the selected source downloader | Wrapper and alias coverage |
| Source `download_*()` | Real | Builds URLs, destinations, and directories | Real download orchestration |
| `process_covariates()` | Real | Dispatches to the selected source processor | Wrapper and alias coverage |
| Source `process_*()` | Real | Reads and transforms the installed fixture | Real `SpatVector` or `SpatRaster` output |
| `calculate_covariates()` | Real | Dispatches to calculation | Wrapper coverage |
| Source `calculate_*()` | Real | Extracts deterministic locations | Real covariate rows and semantic values |
| Contract checks | Real | Checks files, schema, CRS, IDs, and values | Pass rates and contract counts |

The efficiency gain comes from replacing remote and large-data operations, not
from replacing the scientific transformations being tested.

It writes:

- `strategy-comparison-report.txt`: a formatted plain-text report with run
  metadata, a compact summary, key findings, failures, per-iteration results,
  the simulation boundary, and interpretation guidance.
- `strategy-comparison-runs.csv`: every measured iteration.
- `strategy-comparison-summary.csv`: median/minimum/maximum time, pass rate,
  artifact creation, workflow linkage, and contract counts.
- `simulation-component-map.csv`: a machine-readable list of what was simulated,
  what remained real, and the evidence produced by each block.

## Producing retained simulation outputs

The strategy-comparison runner uses temporary directories for each iteration so
repeated measurements do not reuse files from earlier runs. It retains the
reports and metrics, but removes the per-iteration fake archives and extracted
fixtures when each iteration finishes.

Use `SimulateDownloadProcessCalculate.r` when the intermediate workflow files
and calculated covariates must also be retained. From the `amadeus` repository,
run a simulation without a full-data comparison as follows:

```{sh adoption-run-retained-simulation, eval=FALSE}
Rscript ../SimulateDownloadProcessCalculate.r \
  . \
  "" \
  /tmp/ecoregion-simulation-results
```

To compare the simulated fixture with a full downloaded ecoregion file:

```{sh adoption-run-retained-comparison, eval=FALSE}
Rscript ../SimulateDownloadProcessCalculate.r \
  . \
  /path/to/us_eco_l3_state_boundaries.shp \
  /tmp/ecoregion-simulation-results
```

Each run creates a timestamped directory containing:

| Output | Code block that produces it | Purpose |
|---|---|---|
| `simulated-download/zip_files/*.zip` | Simulated `download_run_method()` | Confirms archive destination behavior |
| `simulated-download/data_files/eco_l3_clip.gpkg` | Simulated `download_unzip()` | Provides real input to processing |
| `simulation-test-report.txt` | `SummaryReporter` plus `writeLines()` | Retains assertion status and process messages |
| `simulation-summary.csv` | `simulation_summary` plus `write.csv()` | Records requests, files, features, locations, rows, and columns |
| `simulation-covariates.csv` | Calculated result plus `write.csv()` | Retains the actual scientific output |
| `performance.csv` | Three `system.time()` blocks | Separates download, process, and calculate time |
| `comparison-summary.csv` | Simulated and actual summary block | Compares scale and output dimensions |
| `probe-comparison.csv` | Positive-indicator comparison block | Compares classifications at identical locations |
| `contract-checks.csv` | Schema, geometry, CRS, row, and classification checks | Identifies the exact failed contract |
| `comparison-report.txt` | Comparison status plus `writeLines()` | Provides a readable simulated-versus-actual conclusion |

The two scripts answer different questions:

| Script | Comparison | Files retained |
|---|---|---|
| `compare-test-strategies.r` | Original separated tests versus linked simulation | Text report, summary CSV, run CSV, component map |
| `SimulateDownloadProcessCalculate.r` | Simulated fixture workflow versus optional full-data workflow | Archive, extracted fixture, covariates, timings, contracts, reports |

The simulation blocks improve test efficiency in the following specific ways:

1. `check_url_status = function(...) TRUE` removes URL requests and retry time.
2. `check_destfile = function(...) TRUE` makes branch selection deterministic.
3. The simulated `download_run_method()` writes only a few bytes instead of
   transferring a large archive.
4. The simulated `download_unzip()` copies a small committed fixture instead of
   extracting the complete national dataset.
5. The real `process_ecoregion()` processes 23 fixture features while still
   exercising geometry, CRS, schema, and time handling.
6. Deterministic centroids limit calculation to three known locations.
7. Stage-specific timers show whether remaining cost is in local download
   setup, spatial processing, or calculation.
8. The optional full-data comparison verifies that the faster fixture workflow
   preserves the classifications needed for scientific confidence.

Thus, the simulation is not merely a fake successful return. It replaces costly
external operations with controlled artifacts and then passes those artifacts
through the real processing and calculation implementations.

The runner performs one unmeasured warm-up for each dataset and strategy and
alternates strategy order across measured iterations. This reduces systematic bias from
package initialization and filesystem caching, although it is still a
diagnostic comparison rather than a formal microbenchmark.

Interpret the output using this matrix:

| Field | Expected interpretation |
|---|---|
| `pass_rate` | `1` for both strategies across repeated runs |
| `downloaded_artifacts_created` | `TRUE` for the recommended strategy |
| `process_input_created` | `TRUE` for both; only the recommended input comes from download |
| `download_output_feeds_process` | `TRUE` for the recommended strategy |
| `process_contracts` | More explicit contracts in the recommended strategy |
| `calculate_contracts` | More explicit contracts in the recommended strategy |
| `median_elapsed_seconds` | Low and stable; interpret with added coverage |

A significant improvement does not require the recommended no-network test to
beat the no-op mock by elapsed time. The practical success criteria are:

1. It remains fast enough for routine CI.
2. It passes consistently without a network connection.
3. It produces the archive and extracted-file layout expected by users.
4. Its downloaded output is the actual input passed to processing.
5. It checks more process and calculate contracts.
6. It produces identical semantic results over repeated runs.

Do not compare a five-iteration mocked median with one live download and call the
difference a package runtime improvement. Report it as a test-suite reliability
and efficiency improvement. Runtime optimization of `download_*()` or
`process_*()` requires a separate, like-for-like benchmark of the implementation.

The result is a test suite with three complementary forms of evidence:

```text
unit contracts
    + mocked cross-stage workflow
        + separately gated live-provider checks
```

Unit contracts identify the function that failed. The mocked workflow identifies
incompatible stage boundaries. Live tests, maintained separately, identify
changes in external services.
