---
title: "Testing amadeus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Testing amadeus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

This vignette documents the testing layout and conventions used by
`amadeus`. It is intended for contributors adding tests for an existing
dataset or wiring up tests for a new one.

# Layout

```
tests/
  testthat.R
  testthat/
    helper-fixtures.R          # canonical AOIs, sample dates, tiny rasters
    helper-mocks-download.R    # check_url_status, download_run_method, ...
    helper-mocks-process.R     # terra/sf canned objects, file-listing mocks
    helper-skips.R             # skip_if_no_live_tests, skip_if_no_credentials
    test-<dataset>.R           # mocked / fixture-driven (CRAN/CI safe)
    test-<dataset>-live.R      # live API tests (scheduled CI only)
  testdata/                    # static fixtures
  test_report/                 # quality-scorecard rmd
```

Helper files are auto-loaded by `testthat` (any file starting with `helper-`
in `tests/testthat/`).

# Two tiers of tests

## Mocked / fixture tests (`test-<dataset>.R`)

- Run on every PR via `check-standard.yaml` and in `test-coverage-local.yaml`.
- Must be CRAN-safe: no network calls, no external credentials.
- Use the helpers in `helper-mocks-*.R` to intercept download/IO internals.
- Should cover dispatch logic, branching, error paths, and input validation.

## Live tests (`test-<dataset>-live.R`)

- Run only on the scheduled `test-live` workflow (weekly + `workflow_dispatch`)
  or locally with `AMADEUS_LIVE_TESTS=true`.
- Exercise the real upstream API: one minimal real download per function,
  asserting the file is reachable, non-empty, and (where cheap) parseable.
- Each test must start with `skip_if_no_live_tests()` and any required
  `skip_if_no_credentials("EARTHDATA_TOKEN")` calls.

# Mocking conventions

Use `testthat::local_mocked_bindings(..., .package = "amadeus")` as the only
mocking idiom. Reusable mocks live in `helper-mocks-*.R` as **factory
functions** returning named lists of bindings.

```{r}
testthat::test_that(
  "download_aqs(hash=TRUE, unzip=TRUE): returns hash string",
  {
    local_download_mocks(hash_value = "hash-ok")
    res <- amadeus::download_aqs(
      resolution_temporal = "daily",
      parameter_code = "88101",
      year = 2020,
      directory_to_save = withr::local_tempdir(),
      acknowledgement = TRUE,
      unzip = TRUE,
      hash = TRUE
    )
    testthat::expect_identical(res, "hash-ok")
  }
)
```

To extend or override a factory:

```{r}
local_download_mocks(success = 0L, failed = 1L, url_ok = FALSE)
```

Dataset-specific bindings used in only one file (e.g.
`download_normalize_aqs_unzip`) can be defined inline at the call site.

# Skip conventions

All conditional skipping goes through `helper-skips.R`:

| Helper | When to use |
|---|---|
| `skip_if_no_live_tests()` | Top of every `test-*-live.R` test_that block |
| `skip_if_no_credentials("EARTHDATA_TOKEN")` | Live tests requiring credentials |
| `skip_if_pkg_missing("rstac")` | Tests depending on optional Suggests |

Do not use bare `testthat::skip_on_cran()` in test bodies — that is the job
of the live/mocked split.

# Assertion conventions

- Always namespace: `testthat::expect_*`.
- Prefer typed expectations:

| Avoid | Prefer |
|---|---|
| `expect_true(inherits(x, "SpatRaster"))` | `expect_s4_class(x, "SpatRaster")` |
| `expect_true(inherits(x, "data.frame"))` | `expect_s3_class(x, "data.frame")` |
| `expect_true(file.exists(p))` | keep + `expect_gt(file.info(p)$size, 0)` |
| `expect_true(length(x) > 0)` | `expect_gt(length(x), 0)` or `expect_length` |
| `expect_no_error(f(...))` | assign result and assert on its class/value |

A static check is shipped in `tests/lint_tests.R`. It reports occurrences of
the patterns in the "Avoid" column. Run it locally with:

```bash
Rscript tests/lint_tests.R           # advisory (always exits 0)
Rscript tests/lint_tests.R --strict  # fail build on any hit
```

The CI `lint` workflow runs this script in advisory mode. As legacy weak
assertions are migrated, contributors should aim to keep the count strictly
decreasing.

# Test naming

Test descriptions must encode the input combination under test:

```
test_that("<fn>(<arg=value>, ...): <expected behavior>", { ... })
```

Examples:

- `"download_aqs(resolution_temporal='daily', hash=TRUE): returns hash string"`
- `"download_geos(collection=<bad>, acknowledgement=TRUE): errors on bad collection"`
- `"process_modis_swath(path=<missing>): errors on non-existent path"`

For matrix-style cases (e.g. "all datasets reject `acknowledgement = FALSE`")
use `patrick::with_parameters_test_that()` so each parameter row appears as a
separate test in failure output.

# Adding tests for a new dataset

Follow this checklist (mirroring the dataset-addition workflow in
`AGENTS.md`):

1. Create `tests/testthat/test-<dataset>.R` with:
   - One block per input-combination branch in `download_<dataset>()`.
   - Acknowledgement and directory validation tests.
   - Mock-based success path using `local_download_mocks(...)`.
2. Create `tests/testthat/test-<dataset>-live.R` with one minimal real
   download. Gate with `skip_if_no_live_tests()` and any credential helpers.
3. Add mocked tests for `process_<dataset>()` and (if applicable)
   `calc_<dataset>()` in the same file.
4. Make sure `download_data(dataset_name = "<dataset>")` is exercised by
   `test-download.R` (or `test-download-dispatch.R` once split).

# Running tests

```bash
# Mocked tests only (default)
Rscript -e "devtools::test()"

# A filter
Rscript -e "devtools::test(filter = 'aqs')"

# Live tests only
AMADEUS_LIVE_TESTS=true Rscript -e "devtools::test(filter = '-live$')"

# Regenerate test_report.html
Rscript tests/test_report/render_report.R
```
