---
title: "US EPA Toxic Release Inventory (TRI)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{US EPA Toxic Release Inventory (TRI)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

```

This article demonstrates a compact workflow for EPA TRI facility emissions data.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

`download_tri()` works with EPA's annual TRI basic data files and exposes a small set of high-value selectors:

- `jurisdiction` supports the nationwide file (`"US"`), any two-letter state or territory code such as `"NC"`, or the tribal file (`"tbl"`).
- `year` accepts a single year or a start/end pair, so multi-year requests download one annual TRI file per year.
- TRI downloads are delivered directly as CSV files rather than zip archives.
- Output names reflect the jurisdiction requested: U.S.-wide files keep the historical `tri_raw_<year>.csv` pattern, while state and tribal requests append a jurisdiction suffix such as `_NC` or `_tbl`.
- TRI does not require authentication. Because these are annual facility-reported releases and waste-management totals, temporal resolution is yearly.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "tri_workflow")
download_data(
  dataset_name = "tri",
  year = 2023L,
  jurisdiction = "US",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)

```

## Demonstate processing and covariate calculation for a single chemical 

The helper function `get_tri_info()` lists the available chemicals in the downloaded files, which can be used to filter the processing and covariate calculation steps. Here we demonstrate with Polychlorinated biphenyls (PCBs), a group of persistent organic pollutants that were widely used in industrial applications.

```{r process, eval = live_run}

chems <- get_tri_info(path = directory_to_save, type = "chemicals")

processed_pcb <- process_covariates(
  covariate = "tri",
  path = directory_to_save,
  chemical = c("Polychlorinated biphenyls"),
  year = 2023
)
# Note that extent is an option in process_covariates() to limit the domain 
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
domain_x <- c(terra::xmin(processed_pcb), terra::xmax(processed_pcb))
domain_y <- c(terra::ymin(processed_pcb), terra::ymax(processed_pcb))
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
example_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))


point_values_pcb <- calculate_covariates(
  covariate = "tri",
  from = processed_pcb,
  locs = example_points_sf,
  locs_id = "site_id",
  decay_range = 5000, # 5 km decay range for illustrative purposes
  use_threshold = FALSE,
  geom = "sf"
)
```

## Plot the covariates at points along with the facility locations

```{r plot-points, fig.alt = "Calculated TRI PCB values at sample locations, with facility locations overlaid.", eval = live_run}
pcb_sf <- sf::st_as_sf(processed_pcb)
point_basemap <-
  sf::st_as_sf(maps::map("usa", plot = FALSE, fill = FALSE))

ggplot() + 
  geom_sf(data = point_basemap, fill = "gray80", color = "white") +
  geom_sf(data = pcb_sf, color = "red", size = 2) + 
  geom_sf(data = point_values_pcb, aes(color = STACK_AIR_0001336363_05000), size = 3) 

```

## Demonstate with multiple chemicals and a total emissions

```{r process-chems, eval = live_run}


processed_chems <- process_covariates(
  covariate = "tri",
  path = directory_to_save,
  year = 2023,
  variables = "ON-SITE RELEASE TOTAL"
)
```
