---
title: "US EPA Toxic Release Inventory (TRI)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{US EPA Toxic Release Inventory (TRI)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "`r Sys.Date()`"
author: "Kyle Messier, with assistance from GitHub Copilot"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
library(amadeus)
library(sf)

live_run <- local({
  force <- tolower(Sys.getenv("AMADEUS_RUN_VIGNETTES", "")) %in% c("1", "true", "yes")
  on_ci <- nzchar(Sys.getenv("CI")) ||
    nzchar(Sys.getenv("GITHUB_ACTIONS")) ||
    identical(tolower(Sys.getenv("IN_PKGDOWN", "")), "true")
  on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_")) ||
    identical(tolower(Sys.getenv("NOT_CRAN", "")), "false")
  force || !(on_ci || on_cran)
})

```

This article demonstrates a compact workflow for EPA TRI facility emissions data.

This vignette runs its live workflow when rendered locally. The heavy download, processing, extraction, and plotting chunks are skipped automatically on CI, CRAN checks, and pkgdown builds; set `AMADEUS_RUN_VIGNETTES=true` to force live execution in those environments.

## Available inputs and data availability

`download_tri()` works with EPA's annual TRI basic data files and exposes a small set of high-value selectors:

- `jurisdiction` supports the nationwide file (`"US"`), any two-letter state or territory code such as `"NC"`, or the tribal file (`"tbl"`).
- `year` accepts a single year or a start/end pair, so multi-year requests download one annual TRI file per year.
- TRI downloads are delivered directly as CSV files rather than zip archives.
- Output names reflect the jurisdiction requested: U.S.-wide files keep the historical `tri_raw_<year>.csv` pattern, while state and tribal requests append a jurisdiction suffix such as `_NC` or `_tbl`.
- TRI does not require authentication. Because these are annual facility-reported releases and waste-management totals, temporal resolution is yearly.

## Download representative requests

```{r download, eval = live_run}
directory_to_save <- file.path(tempdir(), "tri_workflow")
download_data(
  dataset_name = "tri",
  year = 2023L,
  jurisdiction = "US",
  directory_to_save = directory_to_save,
  acknowledgement = TRUE
)

```

## Process one workflow-ready data product

```{r process, eval = live_run}

chems <- get_tri_info(path = directory_to_save, type = "chemicals")
processed_pcb <- process_covariates(
  covariate = "tri",
  path = directory_to_save,
  chemical = c("Polychlorinated biphenyls"),
  year = 2023,
  extent = terra::ext(-114.9, -102.0, 31.3, 41.1)
)
```

## Calculate covariates at points

```{r calculate-points, eval = live_run}
domain_x <- c(terra::xmin(processed_pcb), terra::xmax(processed_pcb))
domain_y <- c(terra::ymin(processed_pcb), terra::ymax(processed_pcb))
domain_dx <- diff(domain_x)
domain_dy <- diff(domain_y)

candidate_xy <- expand.grid(
  lon = seq(domain_x[1] + 0.12 * domain_dx, domain_x[2] - 0.12 * domain_dx, length.out = 5),
  lat = seq(domain_y[1] + 0.12 * domain_dy, domain_y[2] - 0.12 * domain_dy, length.out = 5)
)
example_points_sf <- sf::st_as_sf(
  candidate_xy,
  coords = c("lon", "lat"),
  crs = 4326
)
example_points_sf$site_id <- paste0("site_", seq_len(nrow(example_points_sf)))


point_values_pcb <- calculate_covariates(
  covariate = "tri",
  from = processed_pcb,
  locs = example_points_sf,
  locs_id = "site_id",
  decay_range = 5000,
  use_threshold = FALSE,
  geom = "sf"
)
```

## Calculate covariates across Durham County H3 hexagons

```{r calculate-polygons, eval = live_run}
polygon_values <- calculate_covariates(
  covariate = "tri",
  from = processed_pcb,
  locs = durham_hex,
  locs_id = "h3_id",
  decay_range = 5000,
  geom = "sf"
)
```

## Visualize the point and polygon outputs

```{r plot-points, eval = live_run, fig.alt = "Map of point-based covariate extraction results for this workflow."}
plot_points(point_values, paste0("US EPA Toxic Release Inventory (TRI)", ": point extraction"))
```

```{r plot-polygons, eval = live_run, fig.alt = "Map of polygon-based covariate extraction results across Durham County H3 hexagons for this workflow."}
plot_polygons(polygon_values, paste0("US EPA Toxic Release Inventory (TRI)", ": Durham H3 polygons"))
```

## Notes

- The vignette now shows both a multi-year national TRI request and a state-specific download for North Carolina.
- TRI calculations usually benefit from positive buffer radii so nearby facilities contribute to the local summary.
