---
title: "amadeus workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{amadeus workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: "2024-07-29"
author: "Mitchell Manware"
---

```{r setup, include = FALSE}
# packages
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(amadeus)
```

The following vignette will demonstrate how to download, process, and calculate covariates using `amadeus` functions.
The examples will utilize air temperature at 2m height (air.2m) data from the NOAA North American Regional Reanalysis (NARR) dataset.\insertRef{web_NARRabout}

## Download

Download data for the years 2021 and 2022 with the `download_data` function.

```{r, eval = FALSE}
download_data(
  dataset_name = "narr",
  variable = "air.2m",
  year = c(2021, 2022),
  directory_to_save = "./inst/extdata",
  acknowledgement = TRUE,
  download = TRUE,
  remove_command = TRUE
)
```

Check for downlaoded files

```{r, eval = FALSE}
list.files("./inst/extdata/air.2m")
```
```{r, echo = FALSE}
cat('[1] "air.2m.2021.nc" "air.2m.2022.nc"\n')
```

## Process

Process data for all dates from December 28, 2021 to January 3, 2022.

```{r}
temp_process <- process_covariates(
  covariate = "narr",
  variable = "air.2m",
  date = c("2021-12-28", "2022-01-03"),
  path = "./inst/extdata/air.2m"
)
```

Check the `SpatVector` object.

```{r}
temp_process
```

## Calculate covariates

Calculate covariates for North Carolina county boundaries.
County boundaries are accessed with the `tigris::counties` function.\insertRef{pacakge_tigris}
`geom = TRUE` will return the covariates as a `SpatVector` object.

```{r}
library(tigris)
temp_covar <- calc_covariates(
  covariate = "narr",
  from = temp_process,
  locs = tigris::counties("NC", year = 2021),
  locs_id = "NAME",
  radius = 0,
  geom = TRUE
)
```

```{r}
temp_covar
```